### 2020-05-18 paper.tex
started the pdf

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigplan,10pt,anonymous,review]{acmart}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Table-Driven Customization of Web Applications}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.

\author{Geoffrey Litt}
\affiliation{%
  \institution{Massachusetts Institute of Technology}
  \city{Cambridge, MA}
  \country{USA}
}
\email{glitt@mit.edu}

\author{Daniel Jackson}
\affiliation{%
  \institution{Massachusetts Institute of Technology}
  \city{Cambridge, MA}
  \country{USA}
}
\email{dnj@csail.mit.edu}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
% \renewcommand{\shortauthors}{Trovato and Tobin, et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  In this paper we propose \emph{table-driven customization,} a new
  paradigm for enabling end users to customize software applications
  without doing traditional programming\emph{.} Users directly
  manipulate a tabular view of the structured data inside the
  application, rather than writing imperative scripts as in most
  customization tools. We extend this simple model with a spreadsheet
  formula language and custom data editing widgets, which provide
  sufficient expressivity to implement many useful customizations.

  We describe Wildcard, a browser extension which demonstrates
  table-driven customization in the context of web applications. Through
  concrete examples, we demonstrate that this paradigm can be used to
  create useful customizations for a variety of real applications. We
  share reflections from our usage of the Wildcard system, including its
  strengths and limitations relative to other customization approaches.
  We further explore how new software architectures might help
  application developers promote this style of end-user customization.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
%% From HERE
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011066.10011069</concept_id>
<concept_desc>Software and its engineering~Integrated and visual development environments</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Integrated and visual development environments}
% To HERE

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{end-user programming, software customization, web browser extensions}

%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
%\begin{teaserfigure}
%  \includegraphics[width=\textwidth]{sampleteaser}
%  \caption{Seattle Mariners at Spring Training, 2010.}
%  \Description{Enjoying the baseball game from the third-base
%  seats. Ichiro Suzuki preparing to bat.}
%  \label{fig:teaser}
%\end{teaserfigure}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

TBD, Summarize the paper

\hypertarget{examples}{%
\section{Examples}\label{examples}}

\begin{itemize}
\tightlist
\item
  use airbnb + expedia
\item
  instacart?
\item
  new screenshots
\item
  mention in text all the different sites we've used it with
\item
  1 externally contributed adapter
\item
  flux story
\end{itemize}

\hypertarget{system-architecture}{%
\section{System architecture}\label{system-architecture}}

The basic data unit in table-driven customization is a Table: an ordered
list of tuples, each of which has an ID and some attributes. The user
always sees a single table in the table editor, but this table is the
result of combining data across one or more underlying tables, much like
viewing the result of a SQL query across multiple database tables. Each
individual table's data is managed by a ``table adapter'' component, and
a query engine coordinates across the various table adapters.

(\emph{Todo: drive this explanation via concrete Instacart example. Add
a figure.}) For example, one table adapter manages scraping data from
the DOM of a site, while another manages extracting data from AJAX
requests. Each table adapter produces a table of data, and the two are
joined together by a shared ID column to produce what looks like a
single table of data extracted from the site. Similarly, when the user
makes annotations on a site, these annotations are handled by a third
table adapter, which manages a data table stored in the browser's local
storage.

\hypertarget{table-adapter-api}{%
\subsection{Table adapter API}\label{table-adapter-api}}

In order to promote uniformity within the system, we've defined a small
abstract API, which all table adapters conform to. (Todo: formalize this
more, make it more specific?)

\emph{Output: data table}:The main purpose of a table adapter is to
return a table representing some data source. The table is
live-updating: a table adapter can produce a new version of its data
table at any point (e.g., updating the data table in response to a
change in a website's DOM). When this happens the new data is pushed to
the editor and the query results are refreshed.

Table editors also accept various incoming commands.

\emph{Input: record edits}:The editor can request to a table adapter to
make an edit to a record. The meaning of making an edit can vary
depending on the adapter: in the local storage adapter, an annotation
can be persisted into local storage; in the DOM scraping adapter, an
edit can represent filling in a form field. An adapter can also mark
values as read-only if it wouldn't be meaningful to edit them; for
example, the DOM scraping adapter typically marks page content as
read-only, except for editable form fields.

\emph{Input: query information}:The query engine also sends each table
adapter information about the entire cross-table query view being
created by the user. This functionality is currently mainly used by the
DOM scraping adapter to make changes to the original web page:

\begin{itemize}
\tightlist
\item
  observes the sort order of the query view to re-order rows in the page
\item
  observes the content of other joined tables to render annotations in
  the page
\item
  observes the currently selected row in the UI, to highlight the
  corresponding row in the page
\end{itemize}

\hypertarget{specific-table-adapters}{%
\subsection{Specific table adapters}\label{specific-table-adapters}}

\emph{(Todo: maybe move this upwards into the worked example?)}

Currently, our system supports three types of table adapters.

\textbf{DOM scraping adapter}: extracts data from the DOM of a web page,
and manipulates the DOM to re-order rows, edit form entries, and inject
annotations. A programmer only needs to write a single function which
returns scraped data and pointers to relevant DOM elements; the Wildcard
framework uses this function to implement the rest of the needed
functionality.

\textbf{AJAX scraping adapter}: intercepts AJAX requests made by a web
page, and extracts information from those requests. A programmer writes
a function which specifies how to extract data from an AJAX request, and
the framework handles the details of actually intercepting requests and
calling the programmer-defined function.

\textbf{Local storage adapter}: simply stores a table of data in the
browser.

\hypertarget{query-engine}{%
\subsection{Query engine}\label{query-engine}}

The query engine is responsible for coordinating across multiple table
adapters. It joins data across multiple tables and creates a single
result table which is shown to the user through the editor. It also
handles all user interactions and routes appropriate messages to each
table adapter.

First, every query involves a primary DOM scraping table adapter which
associates records in the result with elements in the application's user
interface. At minimum, the primary table adapter needs to return record
IDs and have the ability to manipulate the application's UI. It can also
optionally return data about each record.

Next, additional tables (AJAX data, local storage data) are left joined
by ID. (\emph{todo: discuss IDs here?}) Finally, the result table can be
sorted and filtered by any column.

One way to think of this model is a tiny constrained subset of the SQL
query model. We've found that this simple model has proven sufficient
for meeting the needs of customization in practice, and minimizes the
complexity of supporting more general and arbitrary queries. But because
it fits into the general SQL paradigm, it could theoretically be
extended to support more types of queries.

\hypertarget{table-editor-instruments}{%
\subsection{Table editor instruments}\label{table-editor-instruments}}

Our editor is closer to visual SQL query engines like Sieuferd or
Airtable than a freeform spreadsheet.

\hypertarget{future-table-adapters}{%
\subsection{Future table adapters}\label{future-table-adapters}}

However, we have intentionally designed the table adapter API to be
general enough to encapsulate other types of data and additional
functionality in the future. Here are three concrete examples of such
possibilities:

\textbf{Integrated website adapters}: We have taken pains to design a
customization system which is not limited to web scraping as the only
means for integrating with existing sites, and can also accommodate
first party developers adding support for table-driven customization
directly into their own websites. A DOM Scraping table adapter could be
swapped out for such an ``integrated website adapter,'' which implements
the exact same interface as a scraping adapter but by directly accessing
the internal state of the user interface, without needing any scraping
logic.

While we have not yet created a fully operational integrated website
adapter, we think it is possible to create such adapters for existing
frontend web frameworks which result in minimal effort on the part of
the application developers. For example, we have created an early
prototype of a plugin for the Redux state management library, which uses
the Model-View-Update pattern that represents the entire state of a user
interface as a single centralized object. To configure an integrated
website adapter for such an existing application, the user can specify a
function projecting the centralized application state as a table, and
handlers for how data edits to the table should affect the state.

\textbf{Shared storage adapter}: It would be useful to share user
annotations between people and across devices---for example,
collaboratively taking notes with friends on a list of options for
places to stay on Airbnb. The existing Local Storage Adapter could be
extended to share live synchronized data with other users. This could be
achieved through a centralized web server (perhaps an existing service
like Google Sheets capable of storing tabular data), or through P2P
connections which might improve the privacy guarantees available for the
shared annotations.

\textbf{Third party API data adapter}: Currently, the main mechanism for
including data from web APIs in Wildcard is using spreadsheet formulas.
However, a web API could also be wrapped to expose a table API that
would dynamically create tables in response to queries. For example,
when fetching walkability scores for many GPS locations, the query
engine could request a table of walkability scores for various pairs of
latitude and longitude, and the table adapter could dynamically perform
API queries to populate a result table. (\emph{todo: why is this better
than spreadsheet formulas? I don't think it offers any more room for
batching performance optimization because we've already framed formulas
as whole-column, not row specific})

\hypertarget{design-aspects}{%
\section{Design aspects}\label{design-aspects}}

\hypertarget{data-oriented-programming}{%
\subsection{Data-oriented programming}\label{data-oriented-programming}}

There have been many attempts at simplifying aspects of traditional
programming to enable non-programmer end users to perform customization.
End user scripting languages (AppleScript, VBA, Chickenfoot, CoScripter)
have innovated at the syntax level, aiming to improve ease-of-use.
Visual programming environments (Mac Automator, Zapier) go further by
eliminating text syntax entirely, but maintain the same computational
model. Programming-by-demonstration and macro recording environments
(Applescript, Helena, WebVCR) remove some of the initial programming
burden by starting with concrete demonstrations, but typically require
programming later in the process.

However, all of these approaches share something in common: \textbf{an
imperative programming model}. They all involve a list of commands
executed in sequence, with control flow constructs like conditionals and
loops, and mutable variables. This makes some intuitive sense: many
customizations seem most naturally expressed as a series of commands,
each of which has a side effect on the UI: first enter this value, then
click this button, etc. The result is that end users eventually need to
become familiar with all of these traditional imperative programming
constructs, even if only in diluted form. (todo: bolster the case that
this is a barrier to learning)

Imperative programming is not the only possible model. Spreadsheets have
demonstrated another computational model: data editing via direct
manipulation, combined with pure functional expressions for computation,
with automatic reactive updates provided by the runtime. We call this a
data-oriented model, because the user is focused on editing data values
and producing the correct derived data values, rather than encoding a
sequence of commands. The success of spreadsheets with end user
programmers suggests that finding a way to apply data-oriented
programming model to software customization might enable more people to
modify their software. However, it's not immediately obvious how the
spreadsheet paradigm applies to customization.

A key insight from our work is that, rather than representing
customization as a series of commands issued to an existing application,
we can present a projected view of the application's internal state, and
let the user directly edit that projection as a means of customizing the
application. In some sense, this is similar to the DOM inspector, which
provides a user with an alternate representation of the UI that they can
directly manipulate (e.g., by selecting an element and deleting it from
the DOM tree). But rather than provide a low-level view of the user
interface elements, we instead provide a higher-level view of the
semantic state of the application.

\hypertarget{black-box-semantic}{%
\subsection{Black-box + semantic}\label{black-box-semantic}}

Software customization tools typically fall into one of two categories:
black-box, or semantic.

\textbf{Black-box customization tools} enable customization without
using official extension APIs, enabling a broader range of
customizations on top of more applications. For example, web browser
extensions have demonstrated the utility of customizing websites through
manipulating the DOM, without needing explicit extension APIs to be
built in. However, black-box customization comes with a corresponding
cost: these tools can typically only operate at a low level of
abstraction, e.g.~manipulating user interface elements. This makes it
harder for end users to write scripts, and makes the resulting scripts
more brittle. (todo: support this claim more, provide examples)

\textbf{Semantic customization tools} use explicit extension APIs
provided by the application developer. Examples of this include
accessing a backend web API, or writing a customization in Applescript
that uses an application-specific API. The main benefit is that this
allows the extension author to work with meaningful concepts in the
application domain---``create a new calendar event'' rather than ``click
the button that says new event.''---which makes customizations easier to
build and more robust. However, this style limits the range of
extensions that can be built, to only those which the official plugin
API supports.

(Footnote this paragraph?) This categorization is slightly
oversimplified, and there are existing customization ecosystems that
span both categories. For example, part of the success of browser
extensions stems from the fact that the DOM encourages the use of
standardized semantic UI elements; even if an app developer doesn't
anticipate customization, merely using of semantic and accessible HTML
creates a collateral benefit of easier extension. Another example of a
middle ground is AppleScript, which provides system-wide low-level APIs
for manipulating GUIs as a means of black-box customization, while also
enabling developers to optionally add application-specific APIs to add
support for semantic customization.

With data-oriented customization and Wildcard, we use a hybrid approach
that takes the best of both worlds. Programmers implement an API wrapper
that is internally implemented as a black-box customization, but
externally provides a semantic interface to the application. End users
get an ergonomic and simplified customization experience, but without
the need to depend on a first-party application developer exposing
extension APIs.

One way to view this approach is as introducing a new abstraction
barrier into black-box extension. Typically, a black box customization
script combines two responsibilities: 1) mapping the low-level details
of a user interface to semantic constructs (e.g., using CSS selectors to
find certain page elements), and 2) the actual logic of the specific
customization. (todo: could easily show examples of this from browser
extensions, Chickenfoot, etc) Even though the mapping logic is often
more generic than the specific customization (e.g., finding a given
input element is independent of what text to insert into that element),
the intertwining of these two responsibilities in a single script makes
it very difficult to share the mapping logic across scripts. With
Wildcard we propose a decoupling of these two layers: a
community-maintained mapping layer, shared across many specific
customizations by individual users. This architecture has been
successfully used by projects like Gmail.js, an open source project that
creates a convenient API for browser extensions to interface with the
Gmail web email client.

\hypertarget{proximity-to-use}{%
\subsection{Proximity to use}\label{proximity-to-use}}

borrow from the in-place

\begin{itemize}
\tightlist
\item
  here's where ``automation vs customization'' comes in: automation
  frames it as entering a whole new scripting environment; we frame it
  more as just an alternate UI that you can use.
\item
  this is the heart of the gentle slope
\item
  backend APIs: terrible
\item
  the best would be right inline with the GUI (cite Scotty, instrumental
  work) but this has its own problems
\item
  we settle for a compromise: an alternate structured view. (explicitly
  contrast this clearly)
\item
  the CLI GUI thing
\end{itemize}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
%% \bibliographystyle{ACM-Reference-Format}
%%\bibliography{references2.bib}

\end{document}
\endinput
%%
%% End of file `sample-sigconf.tex'.

### 2020-05-18 compile-latex.sh

@@ -1,0 +1,19 @@
+set -e
+
+pandoc \
+  --filter conditional-render \
+  --filter pandoc-crossref \
+  --filter pandoc-citeproc \
+  -s \
+  -o \
+  paper.tex \
+  --template=sigplan.pandoc \
+  --natbib \
+  --metadata=format:pdf \
+  paper.md
+
+pdflatex paper.tex
+# bibtex paper
+# pdflatex -interaction=batchmode paper.tex
+# pdflatex -interaction=batchmode paper.tex
+open paper.pdf

### 2020-05-19 paper.tex
setup pipeline to google docs and pdf

@@
 
-TBD, Summarize the paper
+Many software applications don't meet the precise needs of their users.
+When this happens, the only recourse is to complain to the developers,
+or, more likely, to simply give up. Back in 1977, in Personal Dynamic
+Media \citep{kay1977}, Alan Kay envisioned personal computing as a
+medium that let a user ``mold and channel its power to his own needs,''
+but today, software behaves more like concrete than clay.
+
+\begin{itemize}
+\tightlist
+\item
+  many approaches to customizing software. Most are imperative command
+  based. Can't get around it.
+\item
+  there are successful systems for visual SQL queries: Airtable,
+  Sieuferd. Even spreadsheets.
+\item
+  what if you could see the data queries powering a website and directly
+  modify them yourself?
+\item
+  we call this table-driven customization.
+\end{itemize}
+
+To make it work with the real world, we develop Wildcard, shim this on
+top of existing web apps. Show in Section Examples that it really works!
+can do tons of useful stuff with real apps.
+
+Architecture: Describe implementation of Wildcard. We've developed
+generic abstractions that make this system even more extensible in the
+future. we also propose a future architecture where apps are built for
+this; they expose their data queries and their structured data results.
+
+Design Principles: explain ideas behind the system, generalizable to
+software customization more generally
+
+Related Work: table editors,
 
@@
 
+\emph{This section still todo. Borrow heavily from Convivial paper, but
+extend with newer results}
+
 \begin{itemize}
@@
 \item
-  use airbnb + expedia
+  Just use airbnb + expedia from Convivial paper?
 \item
-  instacart?
+  instacart + hacker news?
 \item
-  new screenshots
+  summarize/mention all 12 sites we've used it with
 \item
-  mention in text all the different sites we've used it with
+  real usage:
+
+  \begin{itemize}
+  \tightlist
   \item
@@
   \item
-  flux story
+    flux adapter in actual use by Sergio
+  \end{itemize}
 \end{itemize}
@@
 
-\hypertarget{specific-table-adapters}{%
-\subsection{Specific table adapters}\label{specific-table-adapters}}
+\hypertarget{table-adapters}{%
+\subsection{Table adapters}\label{table-adapters}}
 
-\emph{(Todo: maybe move this upwards into the worked example?)}
+Here we describe the three types of table adapters which we have
+implemented so far to power the customizations shown above. Then, to
+demonstrate the generality of the table adapter paradigm, we describe
+three more hypothetical types of table adapters which could extend the
+power of the Wildcard system.
 
-Currently, our system supports three types of table adapters.
+\hypertarget{existing-adapters}{%
+\subsubsection{Existing Adapters}\label{existing-adapters}}
 
-\textbf{DOM scraping adapter}: extracts data from the DOM of a web page,
-and manipulates the DOM to re-order rows, edit form entries, and inject
-annotations. A programmer only needs to write a single function which
-returns scraped data and pointers to relevant DOM elements; the Wildcard
-framework uses this function to implement the rest of the needed
-functionality.
+A \textbf{DOM scraping adapter} extracts data from the DOM of a web
+page, and manipulates the DOM to re-order rows, edit form entries, and
+inject annotations. A programmer only needs to write a single function
+which returns scraped data and pointers to relevant DOM elements; the
+Wildcard framework uses this function to implement the rest of the
+needed functionality.
 
-\textbf{AJAX scraping adapter}: intercepts AJAX requests made by a web
+(\emph{do we want to call DOM scraping adapter a ``live'' adapter
+because its data is also shown elsewhere\ldots?})
+
+An \textbf{AJAX scraping adapter} intercepts AJAX requests made by a web
 page, and extracts information from those requests. A programmer writes
@@
 
-\textbf{Local storage adapter}: simply stores a table of data in the
+The \textbf{local storage adapter} simply stores a table of data in the
 browser.
 
-\hypertarget{query-engine}{%
-\subsection{Query engine}\label{query-engine}}
+\hypertarget{future-adapters}{%
+\subsubsection{Future Adapters}\label{future-adapters}}
 
-The query engine is responsible for coordinating across multiple table
-adapters. It joins data across multiple tables and creates a single
-result table which is shown to the user through the editor. It also
-handles all user interactions and routes appropriate messages to each
-table adapter.
-
-First, every query involves a primary DOM scraping table adapter which
-associates records in the result with elements in the application's user
-interface. At minimum, the primary table adapter needs to return record
-IDs and have the ability to manipulate the application's UI. It can also
-optionally return data about each record.
-
-Next, additional tables (AJAX data, local storage data) are left joined
-by ID. (\emph{todo: discuss IDs here?}) Finally, the result table can be
-sorted and filtered by any column.
-
-One way to think of this model is a tiny constrained subset of the SQL
-query model. We've found that this simple model has proven sufficient
-for meeting the needs of customization in practice, and minimizes the
-complexity of supporting more general and arbitrary queries. But because
-it fits into the general SQL paradigm, it could theoretically be
-extended to support more types of queries.
-
-\hypertarget{table-editor-instruments}{%
-\subsection{Table editor instruments}\label{table-editor-instruments}}
-
-Our editor is closer to visual SQL query engines like Sieuferd or
-Airtable than a freeform spreadsheet.
-
-\hypertarget{future-table-adapters}{%
-\subsection{Future table adapters}\label{future-table-adapters}}
-
-However, we have intentionally designed the table adapter API to be
-general enough to encapsulate other types of data and additional
-functionality in the future. Here are three concrete examples of such
-possibilities:
+we have intentionally designed the table adapter API to be general
+enough to encapsulate other types of data and additional functionality
+in the future. Here are three concrete examples of such possibilities:
 
@@
 as whole-column, not row specific})
+
+\hypertarget{query-engine}{%
+\subsection{Query engine}\label{query-engine}}
+
+The query engine is responsible for coordinating across multiple table
+adapters. It joins data across multiple tables and creates a single
+result table which is shown to the user through the editor. It also
+handles all user interactions and routes appropriate messages to each
+table adapter.
+
+First, every query involves a primary DOM scraping table adapter which
+associates records in the result with elements in the application's user
+interface. At minimum, the primary table adapter needs to return record
+IDs and have the ability to manipulate the application's UI. It can also
+optionally return data about each record.
+
+Next, additional tables (AJAX data, local storage data) are left joined
+by ID. (\emph{todo: discuss IDs here?}) Finally, the result table can be
+sorted and filtered by any column.
+
+One way to think of this model is a tiny constrained subset of the SQL
+query model. We've found that this simple model has proven sufficient
+for meeting the needs of customization in practice, and minimizes the
+complexity of supporting more general and arbitrary queries. But because
+it fits into the general SQL paradigm, it could theoretically be
+extended to support more types of queries.
+
+\begin{itemize}
+\tightlist
+\item
+  formulas: do they go here?
+\end{itemize}
+
+\hypertarget{table-editor}{%
+\subsection{Table editor}\label{table-editor}}
+
+\begin{itemize}
+\tightlist
+\item
+  Our editor is closer to visual SQL query engines like Sieuferd or
+  Airtable than a freeform spreadsheet.
+\item
+  cell editors
+\item
+  Note that there could be other entire table editors
+\end{itemize}
 
@@
 
-With data-oriented customization and Wildcard, we use a hybrid approach
-that takes the best of both worlds. Programmers implement an API wrapper
-that is internally implemented as a black-box customization, but
-externally provides a semantic interface to the application. End users
-get an ergonomic and simplified customization experience, but without
-the need to depend on a first-party application developer exposing
-extension APIs.
+With Wildcard, we use a hybrid approach that takes the best of both
+worlds. Programmers implement an API wrapper that is internally
+implemented as a black-box customization, but externally provides a
+semantic interface to the application. End users get an ergonomic and
+simplified customization experience, but without the need to depend on a
+first-party application developer exposing extension APIs.
 
@@
 
-borrow from the in-place
+\emph{borrow from convivial paper's discussion of in-place}
 
@@
 
+\hypertarget{related-work}{%
+\section{Related Work}\label{related-work}}
+
+Find a way to organize all this:
+
+\begin{itemize}
+\tightlist
+\item
+  our own workshop paper. since then\ldots{}
+
+  \begin{itemize}
+  \tightlist
+  \item
+    fundamentally rearchitected around table adapters
+  \item
+    evaluated on many more websites
+  \item
+    more fully describing how the system works
+  \end{itemize}
+\item
+  browser extensions
+\item
+  instrumental interaction, Scotty, Webstrates
+\item
+  customization research tools: Chickenfoot, Coscripter
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Wildcard, like Chickenfoot, wants to hide HTML from users. But we
+    show a structured data view, whereas Chickenfoot shows nothing
+  \end{itemize}
+\item
+  desktop customization: Applescript, VBA, COM
+\item
+  browser dev tools
+\item
+  database GUIs
+\item
+  spreadsheets
+\end{itemize}
+
 %%

### 2020-05-19 compile-html.sh

@@ -1,0 +1,16 @@
+set -e
+
+ pandoc \
+  --filter conditional-render \
+  --filter pandoc-crossref \
+  --filter pandoc-citeproc \
+  --metadata=format:html \
+  -s \
+  --number-sections \
+  -o paper.html \
+  --css draft.css \
+  --toc \
+  --toc-depth=1 \
+  --variable=toc-title:"Contents" \
+  --template=webpage.html \
+  paper.md

### 2020-05-19 paper.tex
skeleton coming together

@@
 
-Many software applications don't meet the precise needs of their users.
-When this happens, the only recourse is to complain to the developers,
-or, more likely, to simply give up. Back in 1977, in Personal Dynamic
-Media \citep{kay1977}, Alan Kay envisioned personal computing as a
-medium that let a user ``mold and channel its power to his own needs,''
-but today, software behaves more like concrete than clay.
+There have been many attempts at empowering end users to customize their
+software by offering them a simplified programming tool. Some scripting
+languages (AppleScript, Chickenfoot) have a friendly syntax that
+resembles natural language. Visual programming tools (Mac Automator,
+Zapier) eliminate text syntax entirely. Macro recorders (Applescript,
+Helena, WebVCR) remove some of the initial programming burden by letting
+a user start by concrete demonstrations.
+
+These approaches have many differences, but they all share something in
+common: an imperative programming model, with mutable variables,
+conditionals and loops. End users ultimately must use these traditional
+programming constructs to express their ideas, and the object of
+interest is fundamentally a \emph{script}, a sequence of commands.
+
+For decades, we have known about an easier model: \emph{direct
+manipulation} \citep{shneiderman1983}, where ``visibility of the object
+of interest'' replaces ``complex command language syntax''. Direct
+manipulation is widely used in GUIs today, but rarely applied for
+software customization. In this work, we ask: what would it look like to
+build a software customization interface that deeply relies on direct
+manipulation?
+
+We take particular inspiration from spreadsheets and visual database
+query interfaces \citep{2020a, bakke2016}. These tools enable users to
+construct queries, edit data, and compute derived values through direct
+manipulation of the data itself, with intermediate feedback available at
+every step. These tools have successfully enabled millions of end users
+to view, edit, and compute with data.
+
+We have developed a technique called \emph{table-driven customization}
+which applies ideas from visual query interfaces in the context of
+software customization. We provide the end user with a table view where
+they can directly manipulate the internal state of an application.
+Changes in the structured table view result in changes to the original
+user interface of the application, enabling the user to customize an
+application with immediate feedback.
+
+To demonstrate the viability of this approach, we have developed a
+browser extension called Wildcard which implements table-driven
+customization in the context of the Web. Wildcard works with existing
+web applications by using web scraping techniques to extract relevant
+application state into a table and to reify changes back into the
+original application UI. In Section~\ref{sec:examples} we present
+examples of using Wildcard to implement useful customizations across
+many real websites, ranging from sorting lists of data to adding whole
+new features to applications.
+
+In Section~\ref{sec:architecture} we explain the architecture of the
+Wildcard system, focusing on the key abstraction of the \emph{table
+adapter} which manages a single table of data. We explain why some of
+these abstractions are general enough to support many customizations
+beyond the ones we've built so far in Wildcard---for example, we explore
+how applications themselves could be re-architected to expose tabular
+data to end users, rather than relying on web scraping techniques.
+
+Table-driven customization embodies several design principles for
+building customizable software systems, described in
+Section~\ref{sec:design-principles}:
 
@@
 \item
-  many approaches to customizing software. Most are imperative command
-  based. Can't get around it.
+  \emph{Data-oriented programming}: End users should be able to
+  customize an application by examining and modifying its data, rather
+  than by issuing imperative commands to an API
 \item
-  there are successful systems for visual SQL queries: Airtable,
-  Sieuferd. Even spreadsheets.
+  \emph{Black-box + semantic}: Typically, ``black-box'' customization
+  tools, which don't rely on official extension APIs, resort to offering
+  low-level APIs for customization. Instead, we suggest a third-party
+  community of semantic wrappers, offering end users a more convenient
+  semantic API for customization, even if it wasn't built in by the
+  original developer.
 \item
-  what if you could see the data queries powering a website and directly
-  modify them yourself?
-\item
-  we call this table-driven customization.
+  \emph{Gentle slope}: Customization tools should offer a ``gentle
+  slope'' from normal use to customization. We explore how our approach
+  offers end users opportunities to start customizing software with
+  almost no effort, which we hope can make customization a viable part
+  of everyday software usage rather than a rare event.
 \end{itemize}
 
-To make it work with the real world, we develop Wildcard, shim this on
-top of existing web apps. Show in Section Examples that it really works!
-can do tons of useful stuff with real apps.
+Table-driven customization relates to existing work in many areas. Our
+goals overlap with many software customization tools, in the browser and
+on the desktop. Our methods overlap with direct manipulation interfaces
+for working with data, including visual database query systems and
+spreadsheets. We explore these connections and more in
+Section~\ref{sec:related-work}.
 
-Architecture: Describe implementation of Wildcard. We've developed
-generic abstractions that make this system even more extensible in the
-future. we also propose a future architecture where apps are built for
-this; they expose their data queries and their structured data results.
-
-Design Principles: explain ideas behind the system, generalizable to
-software customization more generally
-
-Related Work: table editors,
-
-\hypertarget{examples}{%
-\section{Examples}\label{examples}}
+\hypertarget{sec:examples}{%
+\section{Examples}\label{sec:examples}}
 
@@
 
-\hypertarget{system-architecture}{%
-\section{System architecture}\label{system-architecture}}
+\hypertarget{sec:architecture}{%
+\section{System architecture}\label{sec:architecture}}
 
@@
 
-\hypertarget{design-aspects}{%
-\section{Design aspects}\label{design-aspects}}
+\hypertarget{sec:design-principles}{%
+\section{Design principles}\label{sec:design-principles}}
 
@@
 
-\hypertarget{proximity-to-use}{%
-\subsection{Proximity to use}\label{proximity-to-use}}
+\hypertarget{gentle-slope}{%
+\subsection{Gentle slope}\label{gentle-slope}}
 
@@
 
-\hypertarget{related-work}{%
-\section{Related Work}\label{related-work}}
+\hypertarget{sec:related-work}{%
+\section{Related Work}\label{sec:related-work}}
 
@@
 \item
+  database GUIs:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Sieuferd \citep{bakke2016}
+  \item
+    Airtable \citep{2020a}
+  \item
+    other similar tools?
+  \item
+    Liu \& Jagadish: rules for a spreadsheet algebra for database
+    queries \citep{liu2009}
+  \item
+    mention DIRECT MANIPULATION
+  \end{itemize}
+\item
   browser extensions
 \item
-  instrumental interaction, Scotty, Webstrates
+  instrumental interaction, Scotty \citep{eagan2011}, Webstrates
 \item
@@
 \item
+  visual customization tools:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Mac Automator
+  \item
+    Zapier
+  \item
+    IFTTT
+  \item
+    These are weakly direct manipulation. But usually the thing you
+    directly manipulate in these UIs is the \emph{script}: not the
+    actual \emph{data} you want to operate on. VERY DIFFERENT.
+  \end{itemize}
+\item
   desktop customization: Applescript, VBA, COM
@@
   spreadsheets
+\item
+  ScrAPIr
 \end{itemize}
@@
 %% the bibliography file.
-%% \bibliographystyle{ACM-Reference-Format}
-%%\bibliography{references2.bib}
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{wildcard-onward-bibtex.bib}
 

### 2020-05-19 compile-latex.sh

@@ -13,7 +13,7 @@ pandoc \
   paper.md
 
 pdflatex paper.tex
-# bibtex paper
-# pdflatex -interaction=batchmode paper.tex
-# pdflatex -interaction=batchmode paper.tex
+bibtex paper
+pdflatex -interaction=batchmode paper.tex
+pdflatex -interaction=batchmode paper.tex
 open paper.pdf

### 2020-05-20 paper.tex
edit intro

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Table-Driven Customization of Web Applications}
+\title{Customizing Software by Direct Manipulation of Structured Data}
 
@@
 
-For decades, we have known about an easier model: \emph{direct
+We have known for decades about an alternate approach: \emph{direct
 manipulation} \citep{shneiderman1983}, where ``visibility of the object
 of interest'' replaces ``complex command language syntax''. Direct
-manipulation is widely used in GUIs today, but rarely applied for
-software customization. In this work, we ask: what would it look like to
-build a software customization interface that deeply relies on direct
-manipulation?
+manipulation is the de facto standard in GUIs today, but when it comes
+time to customize software, it is rarely to be found. In this work, we
+ask: what would it look like to build a software customization interface
+that deeply relies on direct manipulation? We take inspiration from
+spreadsheets and visual database query interfaces
+\citep{2020a, bakke2016}, which enable users to construct queries, edit
+data, and compute derived values through direct manipulation. These
+tools have successfully enabled millions of end users to compute with
+data more easily than ever before.
 
-We take particular inspiration from spreadsheets and visual database
-query interfaces \citep{2020a, bakke2016}. These tools enable users to
-construct queries, edit data, and compute derived values through direct
-manipulation of the data itself, with intermediate feedback available at
-every step. These tools have successfully enabled millions of end users
-to view, edit, and compute with data.
-
-We have developed a technique called \emph{table-driven customization}
-which applies ideas from visual query interfaces in the context of
-software customization. We provide the end user with a table view where
-they can directly manipulate the internal state of an application.
-Changes in the structured table view result in changes to the original
-user interface of the application, enabling the user to customize an
-application with immediate feedback.
+In this paper we present a technique called \emph{table-driven
+customization} which applies these ideas from visual query interfaces in
+the context of software customization. We provide the end user with a
+table view where they can directly see and manipulate the internal
+structured data of an application. The table view and the original UI
+serve as coordinated \emph{multiple representations} of the same
+underlying state. Changes in the table view result in immediate
+corresponding changes to the original user interface of the application,
+enabling the user to customize an application with live feedback.
 
@@
 customization in the context of the Web. Wildcard works with existing
-web applications by using web scraping techniques to extract relevant
-application state into a table and to reify changes back into the
-original application UI. In Section~\ref{sec:examples} we present
-examples of using Wildcard to implement useful customizations across
-many real websites, ranging from sorting lists of data to adding whole
-new features to applications.
+applications by using web scraping techniques to extract relevant state
+into a table, and to reify changes back into the original application
+UI. In Section~\ref{sec:examples} we present a many examples of using
+Wildcard to implement useful customizations across real websites,
+ranging from sorting lists of data to adding whole new features to
+applications.
 
 In Section~\ref{sec:architecture} we explain the architecture of the
-Wildcard system, focusing on the key abstraction of the \emph{table
-adapter} which manages a single table of data. We explain why some of
-these abstractions are general enough to support many customizations
-beyond the ones we've built so far in Wildcard---for example, we explore
-how applications themselves could be re-architected to expose tabular
-data to end users, rather than relying on web scraping techniques.
+Wildcard system, focusing on \emph{table adapters} which define the
+semantics of how a data table is connected to some underlying state. We
+explain how the abstractions of table-driven customization are general
+enough to support many customizations beyond the ones we've built so far
+in Wildcard. We also explore how applications themselves could be
+re-architected to expose tabular data to end users, rather than relying
+on web scraping techniques.
 
@@
 \item
-  \emph{Data-oriented programming}: End users should be able to
-  customize an application by examining and modifying its data, rather
-  than by issuing imperative commands to an API
+  \emph{Customization via direct manipulation}: End users should be able
+  to customize an application by examining and modifying its data,
+  rather than by issuing imperative commands to an API.
 \item
-  \emph{Black-box + semantic}: Typically, ``black-box'' customization
-  tools, which don't rely on official extension APIs, resort to offering
-  low-level APIs for customization. Instead, we suggest a third-party
-  community of semantic wrappers, offering end users a more convenient
-  semantic API for customization, even if it wasn't built in by the
-  original developer.
+  \emph{Third-party semantic wrappers}: Typically, ``black-box''
+  customization tools, which don't rely on official extension APIs,
+  resort to offering low-level APIs for customization. Instead, we
+  suggest a third-party community of semantic wrappers, offering end
+  users a more convenient semantic API for customization, even if it
+  wasn't built in by the original developer.
 \item
@@
 
-\emph{This section still todo. Borrow heavily from Convivial paper, but
-extend with newer results}
+To concretely illustrate how table-driven customization works, here are
+several examples of using the Wildcard browser extension, which is
+implemented in Typescript and works with Chrome, Firefox, and Edge.
 
-\begin{itemize}
-\tightlist
-\item
-  Just use airbnb + expedia from Convivial paper?
-\item
-  instacart + hacker news?
-\item
-  summarize/mention all 12 sites we've used it with
-\item
-  real usage:
+\begin{figure*}
+\hypertarget{fig:airbnb-demo}{%
+\centering
+\includegraphics{media/airbnb-demo-300dpi.png}
+\caption{Using Wildcard to augment the Airbnb search page for booking
+accommodations}\label{fig:airbnb-demo}
+}
+\end{figure*}
 
-  \begin{itemize}
-  \tightlist
-  \item
-    1 externally contributed adapter
-  \item
-    flux adapter in actual use by Sergio
-  \end{itemize}
-\end{itemize}
+\hypertarget{augmenting-search-results}{%
+\subsection{Augmenting search results}\label{augmenting-search-results}}
+
+In 2012, the travel site Airbnb removed the ability to sort
+accommodation searches by price. Users could still filter by price
+range, but could no longer view the cheapest listings first. Many users
+complained that the change seemed hostile to users. ``It's so
+frustrating! What is the logic behind not having this function?'' said
+one user on the
+\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
+support forum}. Alas, the feature remains missing to this day.
+
+Using Wildcard, the user can fix this omission, while leaving the page's
+design and the rest of its functionality unchanged.{
+Figure~\ref{fig:airbnb-demo} shows an overview of augmenting the Airbnb
+site.} First, the user opens up the Wildcard panel, which shows a table
+corresponding to the search results in the page. As they click around in
+the table, the corresponding row in the page is highlighted to indicate
+the mapping between the views.
+
+Then, the user clicks on the price column header to sort the spreadsheet
+and the Airbnb UI by price{ (Figure~\ref{fig:airbnb-demo}, Note A)}.
+They also filter to listings with a user rating above 4.5 (another
+feature missing in the original Airbnb UI).
+
+After manipulating the data, the user closes the table view and
+continues using the website. Because the application's UI usually has a
+nicer visual design than a spreadsheet, Wildcard does not aim to replace
+it---at any time, the user can use either the UI, the spreadsheet, or
+both together.
+
+Many websites that show lists of data also offer actions on rows in the
+table, like adding an item to a shopping cart. Wildcard has the ability
+to make these ``row actions'' available in the data table through the
+site adapter. In the Airbnb UI, saving multiple listings to a Favorites
+list requires tediously clicking through them one by one. Using Wildcard
+row actions, the user can select multiple rows and favorite all of them
+with a single click{ (Figure~\ref{fig:airbnb-demo}, Note B)}. Similarly,
+the user can open the detailed pages for many listings at once.
+
+Next, the user wants to jot down some notes about each listing. To do
+this, they type some notes into an additional column in each row, and
+the notes appear inside the listings in the original UI{
+(Figure~\ref{fig:airbnb-demo}, Note C)}. The annotations are saved in
+the browser and associated with the backend ID of the listing, so they
+will appear in future browser sessions that display the same listing.
+
+Wildcard also includes a formula language that enables more
+sophisticated customizations. When traveling without a car, it's useful
+to evaluate potential places to stay based on how walkable the
+surroundings are. Using a formula, the user can integrate Airbnb with
+Walkscore, an API that rates the walkability of any location on a 1-100
+scale. When the user calls the \texttt{walkscore} formula with the
+listing's latitude and longitude as arguments, it returns the walk score
+for that location and shows it as the cell value. Because the cell's
+contents are injected into the page, the score also appears in the UI{
+(Figure~\ref{fig:airbnb-demo}, Note D)}.
+
+\hypertarget{snoozing-todos}{%
+\subsection{Snoozing todos}\label{snoozing-todos}}
+
+\begin{figure*}
+\hypertarget{fig:airbnb-demo}{%
+\centering
+\includegraphics{media/todomvc-demo-300dpi.png}
+\caption{Using Wildcard to add a "snooze" feature to the TodoMVC todo list app}\label{fig:todomvc-demo}
+}
+\end{figure*}
+
+In addition to fetching data from other sources, Wildcard formulas can
+also perform computations. In this example, the user would like to
+augment the TodoMVC todo list app with a ``snooze'' feature, which will
+temporarily hide a todo from the list until a certain date.{
+Figure~\ref{fig:todomvc-demo} shows an overview of this customization.}
+
+The user opens the table view, which shows the text and completed status
+of each todo. They start the customization by adding a new column to
+store the snooze date for each todo{ (Figure~\ref{fig:todomvc-demo},
+Note A)}.
+
+The next step is to hide snoozed todos. The user creates a
+\texttt{snoozed?} column, which uses a formula to compute whether a todo
+is snoozed---i.e., whether it has a snooze date in the future{
+(Figure~\ref{fig:todomvc-demo}, Note B)}. Then, they simply filter the
+table to hide the snoozed todos{ (Figure~\ref{fig:todomvc-demo}, Note
+C)}.
+
+Because the built-in \texttt{NOW()} function always returns the current
+datetime, snoozed todos will automatically appear once their snooze date
+arrives.
+
+Because this implementation of snoozing was built on the spreadsheet
+abstraction, it is completely decoupled from this particular todo list
+app. We envision that users could share these types of customizations as
+generic browser extensions, which could be applied to any site supported
+by Wildcard with no additional effort.
+
+\hypertarget{adding-a-custom-datepicker}{%
+\subsection{Adding a custom
+datepicker}\label{adding-a-custom-datepicker}}
+
+\begin{figure*}
+\hypertarget{fig:airbnb-demo}{%
+\centering
+\includegraphics{media/expedia-demo-300dpi.png}
+\caption{Using Wildcard to augment the Expedia page for booking a flight}\label{fig:expedia-demo}
+}
+\end{figure*}
+
+It might seem that Wildcard is only useful on websites that display
+lists of tabular data, but the table metaphor is flexible enough to
+represent many types of data. For example, a flight search form on
+Expedia can be represented as a single row, with a column corresponding
+to each input{ (Figure~\ref{fig:expedia-demo}, Note A)}.
+
+In some of the previous examples, the table cells were read-only
+(because users can't, for example, change the name or price of an Airbnb
+listing). In this case, the cells are writable, which means that changes
+in the table are reflected in the form inputs. This becomes especially
+useful when combined with GUI widgets that can edit the value of a table
+cell.
+
+Filling in dates for a flight search often requires a cumbersome
+workflow: open up a separate calendar app, find the dates for the trip,
+and then carefully copy them into the form. In Wildcard, the user can
+avoid this by using a datepicker widget that shows the user's personal
+calendar events{ (Figure~\ref{fig:expedia-demo}, Note B)}. The user can
+directly click on the correct date, and it gets inserted into both the
+spreadsheet and the original form.
 
@@
 
+(\emph{Todo: formalize this more? Would math notation help?})
+
 In order to promote uniformity within the system, we've defined a small
-abstract API, which all table adapters conform to. (Todo: formalize this
-more, make it more specific?)
+abstract API, which all table adapters conform to.
 
-\emph{Output: data table}:The main purpose of a table adapter is to
-return a table representing some data source. The table is
-live-updating: a table adapter can produce a new version of its data
-table at any point (e.g., updating the data table in response to a
-change in a website's DOM). When this happens the new data is pushed to
-the editor and the query results are refreshed.
+The main purpose of a table adapter is to return a table representing
+some data source. The table is live-updating: a table adapter can
+produce a new version of its data table at any point (e.g., updating the
+data table in response to a change in a website's DOM). When this
+happens the new data is pushed to the editor and the query results are
+refreshed.
 
@@
 
-\emph{Input: record edits}:The editor can request to a table adapter to
-make an edit to a record. The meaning of making an edit can vary
-depending on the adapter: in the local storage adapter, an annotation
-can be persisted into local storage; in the DOM scraping adapter, an
-edit can represent filling in a form field. An adapter can also mark
-values as read-only if it wouldn't be meaningful to edit them; for
-example, the DOM scraping adapter typically marks page content as
-read-only, except for editable form fields.
+The editor can request to a table adapter to make an edit to a record.
+The meaning of making an edit can vary depending on the adapter: in the
+local storage adapter, an annotation can be persisted into local
+storage; in the DOM scraping adapter, an edit can represent filling in a
+form field. An adapter can also mark values as read-only if it wouldn't
+be meaningful to edit them; for example, the DOM scraping adapter
+typically marks page content as read-only, except for editable form
+fields.
 
-\emph{Input: query information}:The query engine also sends each table
-adapter information about the entire cross-table query view being
-created by the user. This functionality is currently mainly used by the
-DOM scraping adapter to make changes to the original web page:
+The query engine also sends each table adapter information about the
+entire cross-table query view being created by the user. This
+functionality is currently mainly used by the DOM scraping adapter to
+make changes to the original web page:
 
@@
 
-Here we describe the three types of table adapters which we have
-implemented so far to power the customizations shown above. Then, to
-demonstrate the generality of the table adapter paradigm, we describe
-three more hypothetical types of table adapters which could extend the
-power of the Wildcard system.
+Here we describe in more detail the table adapters which we have
+implemented to power the customizations shown above.
 
-\hypertarget{existing-adapters}{%
-\subsubsection{Existing Adapters}\label{existing-adapters}}
+\textbf{DOM scraping adapters} are a key part of the Wildcard system.
+These adapters extract data from the DOM of a web page, and manipulate
+the DOM to re-order rows, edit form entries, and inject annotations. To
+create a DOM scraping adapter for a specific website, a programmer does
+not need to implement all this functionality from scratch. They only
+need to write a single scraping function which uses CSS selectors and
+DOM APIs to extract and return relevant elements. The Wildcard framework
+then wraps this function to implement the rest of the needed
+functionality.
 
-A \textbf{DOM scraping adapter} extracts data from the DOM of a web
-page, and manipulates the DOM to re-order rows, edit form entries, and
-inject annotations. A programmer only needs to write a single function
-which returns scraped data and pointers to relevant DOM elements; the
-Wildcard framework uses this function to implement the rest of the
-needed functionality.
+To determine the viability of writing DOM scraping adapters, we built
+adapters for 10 websites, including transactional sites like Amazon and
+Uber Eats, media consumption sites like Hacker News and Youtube, and a
+medical results entry platform managed by a member of our research
+group. The adapters ranged from 36 to 117 lines of code, averaging 68
+lines. In addition, an external developer unaffiliated with the project
+contributed one adapter, designed to sort the Github page listing a
+user's repositories, and described the experience as ``very
+straightforward.''
 
-(\emph{do we want to call DOM scraping adapter a ``live'' adapter
-because its data is also shown elsewhere\ldots?})
+Some of the challenges of writing a DOM scraping adapter are the same
+ones involved with writing normal web scraping code, but the more
+interactive nature of Wildcard introduces additional challenges. One
+challenge is triggering updates to the spreadsheet data in response to
+UI changes that happen after initial page load. Site adapters are
+responsible for recognizing these changes by observing the DOM. So far,
+we have been able to use event listeners and the MutationObserver API to
+successfully observe changes, but it may prove challenging to observe
+changes on some sites only through the DOM. Another challenge is
+persisting updates to the DOM---some websites use virtual DOM frameworks
+that can occasionally overwrite changes made by Wildcard. In practice,
+we've managed to create an implementation that works around these issues
+and results in a usable customization experience for all the websites
+we've tried.
+
+(\emph{todo: get into the Sorting Problem? i.e., once we sort the page,
+the original sort order is lost from the page. Maybe not
+necessary\ldots{}})
+
+\hypertarget{other-table-adapters}{%
+\subsection{Other table adapters}\label{other-table-adapters}}
 
@@
 
-we have intentionally designed the table adapter API to be general
-enough to encapsulate other types of data and additional functionality
-in the future. Here are three concrete examples of such possibilities:
+We have designed the table adapter API to be general enough to
+additional types of functionality in the future. Here are three concrete
+examples of hypothetical possibilities:
 
-\textbf{Integrated website adapters}: We have taken pains to design a
-customization system which is not limited to web scraping as the only
-means for integrating with existing sites, and can also accommodate
-first party developers adding support for table-driven customization
-directly into their own websites. A DOM Scraping table adapter could be
-swapped out for such an ``integrated website adapter,'' which implements
-the exact same interface as a scraping adapter but by directly accessing
-the internal state of the user interface, without needing any scraping
-logic.
+\textbf{Integrated website adapters}: A key benefit of the table adapter
+abstraction is that Wildcard is not coupled to web scraping as the only
+means for integrating with existing sites, but can also accommodate
+first party developers adding support directly into their own websites.
+A DOM Scraping table adapter could be swapped out for such an
+``integrated website adapter,'' which implements the exact same
+interface as a scraping adapter but by directly accessing the internal
+state of the user interface, without needing any scraping logic.
 
-While we have not yet created a fully operational integrated website
-adapter, we think it is possible to create such adapters for existing
-frontend web frameworks which result in minimal effort on the part of
-the application developers. For example, we have created an early
-prototype of a plugin for the Redux state management library, which uses
-the Model-View-Update pattern that represents the entire state of a user
-interface as a single centralized object. To configure an integrated
-website adapter for such an existing application, the user can specify a
-function projecting the centralized application state as a table, and
-handlers for how data edits to the table should affect the state.
+We think it is possible to create such adapters for existing frontend
+web frameworks which result in minimal effort on the part of the
+application developers. For example, we have created an initial
+prototype of a plugin for the Redux state management library. Redux
+implements the Model-View-Update pattern, which represents the entire
+state of a user interface as a single centralized object. To configure
+an integrated website adapter for such an existing application, the user
+can simply specify a function projecting the centralized application
+state as a table, and define handlers for how data edits to the table
+should affect the state.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  formulas: do they go here?
-\end{itemize}
+The query engine is also responsible for executing formulas. We have
+built a small language resembling the formula languages in visual query
+tools like SIEUFERD. As in those tools, and unlike in spreadsheets,
+formulas automatically apply to an entire column of data, and can
+reference columns rather than only specific cells.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Our editor is closer to visual SQL query engines like Sieuferd or
-  Airtable than a freeform spreadsheet.
-\item
-  cell editors
-\item
-  Note that there could be other entire table editors
-\end{itemize}
+In Wildcard we provide a basic table editor as the user interface on top
+of the query engine. It is built with the Handsontable Javascript
+library, which provides UI elements for viewing and editing a table, as
+well as basic query operations like sorting and filtering.
+
+\emph{todo: other table instruments}
 
@@
 
-\hypertarget{data-oriented-programming}{%
-\subsection{Data-oriented programming}\label{data-oriented-programming}}
+\hypertarget{customization-by-direct-manipulation}{%
+\subsection{Customization by direct
+manipulation}\label{customization-by-direct-manipulation}}
 
@@
 
-\hypertarget{black-box-semantic}{%
-\subsection{Black-box + semantic}\label{black-box-semantic}}
+Add here:
+
+\begin{itemize}
+\tightlist
+\item
+  multiple representations
+\item
+  a discussion of
+\end{itemize}
+
+\hypertarget{third-party-semantic-wrappers}{%
+\subsection{Third-party semantic
+wrappers}\label{third-party-semantic-wrappers}}
 
@@
 
-\textbf{Black-box customization tools} enable customization without
+\textbf{Third-party customization tools} enable customization without
 using official extension APIs, enabling a broader range of
@@
 manipulating the DOM, without needing explicit extension APIs to be
-built in. However, black-box customization comes with a corresponding
+built in. However, third-party customization comes with a corresponding
 cost: these tools can typically only operate at a low level of
@@
 
-(Footnote this paragraph?) This categorization is slightly
+(\emph{Footnote this paragraph?}) This categorization is slightly
 oversimplified, and there are existing customization ecosystems that
@@
 
-\emph{borrow from convivial paper's discussion of in-place}
+\emph{borrow from convivial paper's discussion of in-place toolchain}
 
@@
 
-Find a way to organize all this:
-
 \begin{itemize}
@@
 \item
-  our own workshop paper. since then\ldots{}
+  our own workshop paper \citep{litt2020}. since then\ldots{}
 
@@
 \item
-  database GUIs:
+  customization tools
 
@@
   \item
-    Sieuferd \citep{bakke2016}
-  \item
-    Airtable \citep{2020a}
-  \item
-    other similar tools?
-  \item
-    Liu \& Jagadish: rules for a spreadsheet algebra for database
-    queries \citep{liu2009}
-  \item
-    mention DIRECT MANIPULATION
-  \end{itemize}
-\item
     browser extensions
   \item
-  instrumental interaction, Scotty \citep{eagan2011}, Webstrates
+    ScrAPIr
+  \item
+    Thresher
+  \item
+    Sifter
   \item
     customization research tools: Chickenfoot, Coscripter
-
-  \begin{itemize}
-  \tightlist
   \item
@@
     show a structured data view, whereas Chickenfoot shows nothing
-  \end{itemize}
   \item
@@
     desktop customization: Applescript, VBA, COM
+  \end{itemize}
+\item
+  database GUIs:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Sieuferd \citep{bakke2016}
+  \item
+    Airtable \citep{2020a}
+  \item
+    other similar tools?
+  \item
+    Liu \& Jagadish: rules for a spreadsheet algebra for database
+    queries \citep{liu2009}
+  \item
+    Shneiderman paper
+  \item
+    spreadsheets
+  \end{itemize}
+\item
+  instrumental interaction, Scotty \citep{eagan2011}, Webstrates
 \item
@@
 \item
-  database GUIs
+  Self, Smalltalk
 \item
-  spreadsheets
+  multiple representations
 \item
-  ScrAPIr
+  Semantic Web
 \end{itemize}

### 2020-05-20 paper.tex
better architecture intro

@@
 
-  We describe Wildcard, a browser extension which demonstrates
+  We describe Wildcard, a browser extension which implements
   table-driven customization in the context of web applications. Through
   concrete examples, we demonstrate that this paradigm can be used to
-  create useful customizations for a variety of real applications. We
-  share reflections from our usage of the Wildcard system, including its
+  create useful customizations for real applications. We share
+  reflections from our usage of the Wildcard system, including its
   strengths and limitations relative to other customization approaches.
-  We further explore how new software architectures might help
-  application developers promote this style of end-user customization.
+  We further explore how new software architectures might encourage
+  application developers to promote this style of end-user
+  customization.
 \end{abstract}
@@
 programming constructs to express their ideas, and the object of
-interest is fundamentally a \emph{script}, a sequence of commands.
+interest is a \emph{script}, a sequence of commands.
 
@@
 manipulation is the de facto standard in GUIs today, but when it comes
-time to customize software, it is rarely to be found. In this work, we
-ask: what would it look like to build a software customization interface
-that deeply relies on direct manipulation? We take inspiration from
+to customizing software, it is rarely to be found. In this work, we ask:
+what would it look like to build a software customization interface that
+deeply relies on direct manipulation? We take inspiration from
 spreadsheets and visual database query interfaces
-\citep{2020a, bakke2016}, which enable users to construct queries, edit
-data, and compute derived values through direct manipulation. These
-tools have successfully enabled millions of end users to compute with
-data more easily than ever before.
+\citep{2020a, bakke2016}, which have successfully enabled end users to
+construct queries, edit data, and compute derived values through direct
+manipulation of structured data.
 
@@
 customization} which applies these ideas from visual query interfaces in
-the context of software customization. We provide the end user with a
-table view where they can directly see and manipulate the internal
-structured data of an application. The table view and the original UI
-serve as coordinated \emph{multiple representations} of the same
-underlying state. Changes in the table view result in immediate
-corresponding changes to the original user interface of the application,
-enabling the user to customize an application with live feedback.
+the context of software customization. We augment an application's UI
+with a table view, where the user can see and manipulate the structured
+data inside an application. Changes in the table view result in
+immediate corresponding changes to the original user interface of the
+application, enabling the user to customize an application with live
+feedback. The table view and the original UI serve as coordinated
+\emph{multiple representations} of the same underlying state.
 
@@
 browser extension called Wildcard which implements table-driven
-customization in the context of the Web. Wildcard works with existing
-applications by using web scraping techniques to extract relevant state
-into a table, and to reify changes back into the original application
-UI. In Section~\ref{sec:examples} we present a many examples of using
-Wildcard to implement useful customizations across real websites,
-ranging from sorting lists of data to adding whole new features to
-applications.
+customization in the context of the Web. Wildcard is compatible with
+existing unmodified applications because it uses web scraping techniques
+to extract relevant state into a table and to reify changes in the
+original application UI. In Section~\ref{sec:examples} we present a many
+examples of using Wildcard to implement useful customizations across
+real websites, ranging from sorting lists of data to adding whole new
+features to applications.
 
-In Section~\ref{sec:architecture} we explain the architecture of the
-Wildcard system, focusing on \emph{table adapters} which define the
-semantics of how a data table is connected to some underlying state. We
-explain how the abstractions of table-driven customization are general
-enough to support many customizations beyond the ones we've built so far
-in Wildcard. We also explore how applications themselves could be
-re-architected to expose tabular data to end users, rather than relying
-on web scraping techniques.
+In Section~\ref{sec:architecture} explain the \emph{table adapter}
+abstraction, which allows many different types of underlying state to be
+bidirectionally mapped to a table of data. We explore how table adapters
+are general enough to support many customizations beyond the ones we've
+built so far in Wildcard; we also discuss how applications could be
+re-architected to expose tabular data directly to end users, rather than
+relying on web scraping techniques.
+
+In Section~\ref{sec:prototype} we discuss in more detail the
+implementation of the Wildcard browser extension. We reflect on the
+experience of using the tool to customize websites, and some of the
+challenges from the perspective of programmers and end users.
 
 Table-driven customization embodies several design principles for
-building customizable software systems, described in
-Section~\ref{sec:design-principles}:
+building customizable software systems, discussed in
+Section~\ref{sec:discussion}:
 
@@
 \item
-  \emph{Customization via direct manipulation}: End users should be able
+  \emph{Customization by direct manipulation}: End users should be able
   to customize an application by examining and modifying its data,
-  rather than by issuing imperative commands to an API.
+  rather than by writing imperative scripts.
 \item
-  \emph{Third-party semantic wrappers}: Typically, ``black-box''
-  customization tools, which don't rely on official extension APIs,
-  resort to offering low-level APIs for customization. Instead, we
-  suggest a third-party community of semantic wrappers, offering end
-  users a more convenient semantic API for customization, even if it
-  wasn't built in by the original developer.
+  \emph{Third-party semantic wrappers}: Typically, customization tools
+  which don't rely on official extension APIs resort to offering
+  low-level APIs for customization. Instead, we suggest a
+  community-maintained library of semantic wrappers around existing
+  applications, offering end users a more convenient semantic API for
+  customization.
 \item
-  \emph{Gentle slope}: Customization tools should offer a ``gentle
-  slope'' from normal use to customization. We explore how our approach
-  offers end users opportunities to start customizing software with
-  almost no effort, which we hope can make customization a viable part
-  of everyday software usage rather than a rare event.
+  \emph{Gentle slope}: Customization tools should ideally offer a
+  ``gentle slope'' from normal use to customization. We explore how our
+  approach meets this goal by making some software customizations as
+  easy as normal use of the software.
 \end{itemize}
 
-Table-driven customization relates to existing work in many areas. Our
-goals overlap with many software customization tools, in the browser and
-on the desktop. Our methods overlap with direct manipulation interfaces
-for working with data, including visual database query systems and
+Table-driven customization relates to existing work in many areas. In
+particular, our goals overlap with many software customization tools,
+and our methods overlap with direct manipulation interfaces for working
+with structured data, including visual database query systems and
 spreadsheets. We explore these connections and more in
@@
 
-To concretely illustrate how table-driven customization works, here are
-several examples of using the Wildcard browser extension, which is
-implemented in Typescript and works with Chrome, Firefox, and Edge.
+To concretely illustrate the end user experience of table-driven
+customization, here are several real examples of using the Wildcard
+browser extension to customize websites.
 
@@
 
+\begin{itemize}
+\tightlist
+\item
+  basic idea: end user editing a table. but these edits have effects.
+\item
+  table adapter: component that presents an interface similar to a SQL
+  table. but inside, it can do anything to produce that data, and can do
+  anything with incoming edits.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    out: table
+  \item
+    in: table
+  \item
+    in: other tables
+  \end{itemize}
+\end{itemize}
+
+\begin{figure}
+\hypertarget{fig:table-adapter}{%
+\centering
+\includegraphics[width=\columnwidth]{media/table-adapter.eps}
+\caption{The table adapter architecture}\label{fig:table-adapter}
+}
+\end{figure}
+
 The basic data unit in table-driven customization is a Table: an ordered
@@
 
-(\emph{Todo: drive this explanation via concrete Instacart example. Add
-a figure.}) For example, one table adapter manages scraping data from
-the DOM of a site, while another manages extracting data from AJAX
-requests. Each table adapter produces a table of data, and the two are
-joined together by a shared ID column to produce what looks like a
-single table of data extracted from the site. Similarly, when the user
-makes annotations on a site, these annotations are handled by a third
-table adapter, which manages a data table stored in the browser's local
+(\emph{Todo: drive this explanation via concrete example. Add a
+figure.}) For example, one table adapter manages scraping data from the
+DOM of a site, while another manages extracting data from AJAX requests.
+Each table adapter produces a table of data, and the two are joined
+together by a shared ID column to produce what looks like a single table
+of data extracted from the site. Similarly, when the user makes
+annotations on a site, these annotations are handled by a third table
+adapter, which manages a data table stored in the browser's local
 storage.
@@
 functionality.
-
-To determine the viability of writing DOM scraping adapters, we built
-adapters for 10 websites, including transactional sites like Amazon and
-Uber Eats, media consumption sites like Hacker News and Youtube, and a
-medical results entry platform managed by a member of our research
-group. The adapters ranged from 36 to 117 lines of code, averaging 68
-lines. In addition, an external developer unaffiliated with the project
-contributed one adapter, designed to sort the Github page listing a
-user's repositories, and described the experience as ``very
-straightforward.''
-
-Some of the challenges of writing a DOM scraping adapter are the same
-ones involved with writing normal web scraping code, but the more
-interactive nature of Wildcard introduces additional challenges. One
-challenge is triggering updates to the spreadsheet data in response to
-UI changes that happen after initial page load. Site adapters are
-responsible for recognizing these changes by observing the DOM. So far,
-we have been able to use event listeners and the MutationObserver API to
-successfully observe changes, but it may prove challenging to observe
-changes on some sites only through the DOM. Another challenge is
-persisting updates to the DOM---some websites use virtual DOM frameworks
-that can occasionally overwrite changes made by Wildcard. In practice,
-we've managed to create an implementation that works around these issues
-and results in a usable customization experience for all the websites
-we've tried.
 
@@
 
-\hypertarget{sec:design-principles}{%
-\section{Design principles}\label{sec:design-principles}}
+\hypertarget{sec:evaluation}{%
+\section{Evaluation}\label{sec:evaluation}}
+
+To determine the viability of writing DOM scraping adapters, we built
+adapters for 10 websites, including transactional sites like Amazon and
+Uber Eats, media consumption sites like Hacker News and Youtube, and a
+medical results entry platform managed by a member of our research
+group. The adapters ranged from 36 to 117 lines of code, averaging 68
+lines. In addition, an external developer unaffiliated with the project
+contributed one adapter, designed to sort the Github page listing a
+user's repositories, and described the experience as ``very
+straightforward.''
+
+Some of the challenges of writing a DOM scraping adapter are the same
+ones involved with writing normal web scraping code, but the more
+interactive nature of Wildcard introduces additional challenges. One
+challenge is triggering updates to the spreadsheet data in response to
+UI changes that happen after initial page load. Site adapters are
+responsible for recognizing these changes by observing the DOM. So far,
+we have been able to use event listeners and the MutationObserver API to
+successfully observe changes, but it may prove challenging to observe
+changes on some sites only through the DOM. Another challenge is
+persisting updates to the DOM---some websites use virtual DOM frameworks
+that can occasionally overwrite changes made by Wildcard. In practice,
+we've managed to create an implementation that works around these issues
+and results in a usable customization experience for all the websites
+we've tried.
+
+\hypertarget{sec:discussion}{%
+\section{Discussion}\label{sec:discussion}}
 
@@
 \item
-  multiple representations
+  multiple representations \citep{ainsworth1999}
 \item
-  a discussion of
+  address Tchernavskij's point about contrasting with instrumental
+  interaction
 \end{itemize}
@@
 Software customization tools typically fall into one of two categories:
-black-box, or semantic.
+third-party, or semantic.
 
@@
 
-\emph{borrow from convivial paper's discussion of in-place toolchain}
+Originally came from \citep{maclean1990}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  our own workshop paper \citep{litt2020}. since then\ldots{}
+This paper extends a workshop paper by Litt and Jackson \citep{litt2020}
+which presented an earlier prototype of the Wildcard system. We build on
+that work in this paper by rearchitecting the internal implementation of
+the system around the table adapter abstraction, evaluating it on many
+more websites, and by more fully characterizing the behavior and design
+of the system.
+
+\hypertarget{customization-tools}{%
+\subsection{Customization tools}\label{customization-tools}}
 
@@
 \tightlist
-  \item
-    fundamentally rearchitected around table adapters
-  \item
-    evaluated on many more websites
-  \item
-    more fully describing how the system works
-  \end{itemize}
 \item
@@
   \end{itemize}
+\end{itemize}
+
+\hypertarget{visual-database-query-tools}{%
+\subsection{Visual database query
+tools}\label{visual-database-query-tools}}
+
+\begin{itemize}
+\tightlist
 \item
@@
   \end{itemize}
+\end{itemize}
+
+\hypertarget{other}{%
+\subsection{Other}\label{other}}
+
+\begin{itemize}
+\tightlist
 \item

### 2020-05-20 paper.tex
latest draft

@@
 what would it look like to build a software customization interface that
-deeply relies on direct manipulation? We take inspiration from
-spreadsheets and visual database query interfaces
-\citep{2020a, bakke2016}, which have successfully enabled end users to
-construct queries, edit data, and compute derived values through direct
-manipulation of structured data.
+relies on direct manipulation? We take inspiration from spreadsheets and
+visual database query interfaces \citep{2020a, bakke2016}, which have
+successfully enabled end users to run queries and computations through
+direct manipulation of data.
 
@@
 application, enabling the user to customize an application with live
-feedback. The table view and the original UI serve as coordinated
-\emph{multiple representations} of the same underlying state.
+feedback.
 
-To demonstrate the viability of this approach, we have developed a
-browser extension called Wildcard which implements table-driven
-customization in the context of the Web. Wildcard is compatible with
-existing unmodified applications because it uses web scraping techniques
-to extract relevant state into a table and to reify changes in the
-original application UI. In Section~\ref{sec:examples} we present a many
-examples of using Wildcard to implement useful customizations across
-real websites, ranging from sorting lists of data to adding whole new
-features to applications.
+We have developed a browser extension called Wildcard which uses web
+scraping techniques to implement table-driven customization for existing
+Web applications. In Section~\ref{sec:examples} we concretely introduce
+the key ideas of table-driven customization by presenting several
+examples of real customizations implemented in Wildcard.
 
-In Section~\ref{sec:architecture} explain the \emph{table adapter}
+In Section~\ref{sec:architecture} we explain the architecture of
+table-driven customization. We focus on the \emph{table adapter}
 abstraction, which allows many different types of underlying state to be
-bidirectionally mapped to a table of data. We explore how table adapters
-are general enough to support many customizations beyond the ones we've
-built so far in Wildcard; we also discuss how applications could be
-re-architected to expose tabular data directly to end users, rather than
-relying on web scraping techniques.
+bidirectionally mapped to a table of data. We describe certain types of
+table adapters we've built in Wildcard; we also describe future adapters
+that could be supported in the general paradigm.
 
-In Section~\ref{sec:prototype} we discuss in more detail the
-implementation of the Wildcard browser extension. We reflect on the
-experience of using the tool to customize websites, and some of the
-challenges from the perspective of programmers and end users.
+We have used Wildcard to build real customizations for 11 different
+websites. In Section~\ref{sec:evaluation} we present reflections from
+this process: customizations we were able to build, limitations we
+encountered, and reflections on the ease of integrating scraping logic
+with real websites.
 
@@
 building customizable software systems, discussed in
-Section~\ref{sec:discussion}:
+Section~\ref{sec:themes}:
 
@@
 \section{Examples}\label{sec:examples}}
+
+\emph{todo: maybe redo this as a separate example to avoid self
+plagiarism?}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  basic idea: end user editing a table. but these edits have effects.
-\item
-  table adapter: component that presents an interface similar to a SQL
-  table. but inside, it can do anything to produce that data, and can do
-  anything with incoming edits.
-
-  \begin{itemize}
-  \tightlist
-  \item
-    out: table
-  \item
-    in: table
-  \item
-    in: other tables
-  \end{itemize}
-\end{itemize}
-
 \begin{figure}
@@
 
-The basic data unit in table-driven customization is a Table: an ordered
-list of tuples, each of which has an ID and some attributes. The user
-always sees a single table in the table editor, but this table is the
-result of combining data across one or more underlying tables, much like
-viewing the result of a SQL query across multiple database tables. Each
-individual table's data is managed by a ``table adapter'' component, and
-a query engine coordinates across the various table adapters.
+Figure~\ref{fig:table-adapter} summarizes the overall architecture of
+table-driven customization, using a simplified version of the Airbnb
+example above. In this example, the name of each listing is scraped from
+the web page DOM, the latitude and longitude of each listing is scraped
+from AJAX responses, and user annotations are loaded from the brower's
+local storage.
 
-(\emph{Todo: drive this explanation via concrete example. Add a
-figure.}) For example, one table adapter manages scraping data from the
-DOM of a site, while another manages extracting data from AJAX requests.
-Each table adapter produces a table of data, and the two are joined
-together by a shared ID column to produce what looks like a single table
-of data extracted from the site. Similarly, when the user makes
-annotations on a site, these annotations are handled by a third table
-adapter, which manages a data table stored in the browser's local
-storage.
+First, the three different data sources are each bidirectionally mapped
+to a table interface by a \textbf{table adapter}. The table adapter
+defines how to map a particular type of data to a table, and what
+effects edits should have on the original data source. In some cases,
+the mapping logic is straightforward: the local storage adapter stores a
+table of data, so the mapping to the table abstraction is trivial. In
+other cases, the mapping is more involved: the DOM scraping adapter
+implements web scraping logic to produce a table of data from the web
+page, and edits to the table result in DOM manipulations like reordering
+rows of data on the page.
 
-\hypertarget{table-adapter-api}{%
-\subsection{Table adapter API}\label{table-adapter-api}}
+Ultimately, we provide the user with a single combined table view that
+they can edit, rather than three separate tables. The \textbf{query
+engine} is responsible for combining the tables of data into this single
+view, and routing the user's edits back to the individual table
+adapters. In this example, the query engine has joined the three tables
+together by a shared ID column, and sorted the result by the name
+column.
 
-(\emph{Todo: formalize this more? Would math notation help?})
-
-In order to promote uniformity within the system, we've defined a small
-abstract API, which all table adapters conform to.
-
-The main purpose of a table adapter is to return a table representing
-some data source. The table is live-updating: a table adapter can
-produce a new version of its data table at any point (e.g., updating the
-data table in response to a change in a website's DOM). When this
-happens the new data is pushed to the editor and the query results are
-refreshed.
-
-Table editors also accept various incoming commands.
-
-The editor can request to a table adapter to make an edit to a record.
-The meaning of making an edit can vary depending on the adapter: in the
-local storage adapter, an annotation can be persisted into local
-storage; in the DOM scraping adapter, an edit can represent filling in a
-form field. An adapter can also mark values as read-only if it wouldn't
-be meaningful to edit them; for example, the DOM scraping adapter
-typically marks page content as read-only, except for editable form
-fields.
-
-The query engine also sends each table adapter information about the
-entire cross-table query view being created by the user. This
-functionality is currently mainly used by the DOM scraping adapter to
-make changes to the original web page:
-
-\begin{itemize}
-\tightlist
-\item
-  observes the sort order of the query view to re-order rows in the page
-\item
-  observes the content of other joined tables to render annotations in
-  the page
-\item
-  observes the currently selected row in the UI, to highlight the
-  corresponding row in the page
-\end{itemize}
+We now examine each component of the system in more detail.
 
@@
 
+A key idea in table-driven customization is that different data sources
+can all be mapped to the generic abstraction of an editable table. In a
+relational database, the table matches the underlying storage format,
+but in table-driven customization, the table is merely an
+\emph{interface layer}. The data shown in the table is a projection of
+some underlying state, and edits to the table can have arbitrary effects
+on the underlying state.
+
+Externally, a table adapter must satisfy an abstract interface. The
+first two parts of the interface largely resemble the interface exposed
+by a table in a relational database:
+
+\textbf{Returning a table}: A table adapter exposes a table of data: an
+ordered list of records. Each record contains associates attributes and
+values. Tables have a typed schema, so the same attributes are shared
+across all records. A table adapter can update the contents of a table
+at any time, in response to changes in the underlying state (e.g., a DOM
+scraping adapter can update the table when the page body changes). New
+data is pushed out to other components of the system, and the query view
+reactively updates in response.
+
+\textbf{Making edits}: The query engine can request to a table adapter
+to make an edit to a record. The meaning of making an edit can vary
+depending on the adapter: in the local storage adapter, an annotation
+can be persisted into local storage; in the DOM scraping adapter, an
+edit can represent filling in a form field. An adapter can also mark
+values as read-only if it wouldn't be meaningful to edit them; for
+example, the DOM scraping adapter typically marks page content as
+read-only, except for editable form fields.
+
+The query engine also sends additional information about the combined
+query view to each table adapter. These functions are currently used to
+provide the DOM scraping adapter with sufficient information to
+manipulate the original application UI as the user manipulates the table
+view.
+
+\textbf{Sorting/filtering}: When the user sorts or filters the query
+view, an ordered list of visible IDs is sent to each table adapter. The
+DOM scraping adapter uses this information to modify the list of visible
+rows in the web page.
+
+\textbf{Information from other tables}: The query engine sends each
+table adapter the entire combined table of data being rendered to the
+user. The DOM scraping adapter uses this to add annotations to the page
+by looking for additional data columns joined to scraped rows, and
+rendering them in the page.
+
+\textbf{Currently selected record}: The query engine notifies each table
+adapter about which record is currently selected by the user in the
+table UI. The DOM scraping adapter uses this information to highlight
+the row in the page that corresponds to the selected record in the
+table.
+
+\hypertarget{types-of-table-adapters}{%
+\subsubsection{Types of table adapters}\label{types-of-table-adapters}}
+
 Here we describe in more detail the table adapters which we have
-implemented to power the customizations shown above.
+implemented in Wildcard to power the customizations shown in
+Section~\ref{sec:examples}.
 
-\textbf{DOM scraping adapters} are a key part of the Wildcard system.
-These adapters extract data from the DOM of a web page, and manipulate
-the DOM to re-order rows, edit form entries, and inject annotations. To
-create a DOM scraping adapter for a specific website, a programmer does
-not need to implement all this functionality from scratch. They only
-need to write a single scraping function which uses CSS selectors and
-DOM APIs to extract and return relevant elements. The Wildcard framework
-then wraps this function to implement the rest of the needed
-functionality.
+\textbf{DOM scraping adapters} are the essential component that enables
+Wildcard to interface with an existing website UI. In addition to the
+standard web scraping problem of extracting a table of data from the
+DOM, a scraping adapter must also manipulate the DOM to re-order rows,
+edit form entries, and inject annotations as the table is edited.
 
-(\emph{todo: get into the Sorting Problem? i.e., once we sort the page,
-the original sort order is lost from the page. Maybe not
-necessary\ldots{}})
-
-\hypertarget{other-table-adapters}{%
-\subsection{Other table adapters}\label{other-table-adapters}}
+Because each website has unique content, we rely on programmers to
+create a DOM scraping adapter for each individual website to make it
+available for customization in Wildcard. To make this approach viable,
+we have built library functions that make it as easy as possible to
+implement a scraping adapter, only requiring the programmer to implement
+the minimal site-specific parts. The programmer writes a single scraping
+function which uses CSS selectors and DOM APIs to extract and return
+relevant elements from the page. The library functions then wrap this
+function to implement the rest of the needed functionality; for example,
+when the table is sorted, we find the DOM elements corresponding to the
+rows in the table, remove them from the page, and then reinsert them in
+the new order.
 
 An \textbf{AJAX scraping adapter} intercepts AJAX requests made by a web
-page, and extracts information from those requests. A programmer writes
-a function which specifies how to extract data from an AJAX request, and
-the framework handles the details of actually intercepting requests and
-calling the programmer-defined function.
+page, and extracts information from those requests. When available, this
+tends to be a helpful technique because the data is already in a
+structured form, and often includes information not shown in the UI. As
+with DOM scraping adapters, we have made it as easy as possible for
+programmers to create site-specific AJAX scraping adapters. A programmer
+writes a function which specifies how to extract data from an AJAX
+request, and the framework handles the details of actually intercepting
+requests and calling the programmer-defined function.\footnote{So far we
+  have only implemented AJAX scraping in the Firefox version of
+  Wildcard, since Firefox has convenient APIs for intercepting requests.
+  It appears possible to implement in Chrome as well, but we have not
+  finished our implementation.}
 
 The \textbf{local storage adapter} simply stores a table of data in the
-browser.
+browser. As shown in the example use cases, this can be useful for
+privately persisting annotations, without needing to upload them to a
+web service.
 
@@
 
-We have designed the table adapter API to be general enough to
-additional types of functionality in the future. Here are three concrete
-examples of hypothetical possibilities:
+We have designed the table adapter API to be general enough to support
+other types of useful adapters in the future. Here are three examples of
+future possibilities:
 
@@
 first party developers adding support directly into their own websites.
-A DOM Scraping table adapter could be swapped out for such an
-``integrated website adapter,'' which implements the exact same
-interface as a scraping adapter but by directly accessing the internal
-state of the user interface, without needing any scraping logic.
+A DOM Scraping table adapter could be swapped out for an ``integrated
+website adapter'' which directly accesses the internal state of the
+application, without needing to extract it from the UI.
 
-We think it is possible to create such adapters for existing frontend
-web frameworks which result in minimal effort on the part of the
-application developers. For example, we have created an initial
-prototype of a plugin for the Redux state management library. Redux
-implements the Model-View-Update pattern, which represents the entire
-state of a user interface as a single centralized object. To configure
-an integrated website adapter for such an existing application, the user
-can simply specify a function projecting the centralized application
-state as a table, and define handlers for how data edits to the table
-should affect the state.
+With the advent of rich frontend web frameworks, structured application
+state is now often available in the web client. We suspect it is
+possible to create plugins for frontend frameworks which expose
+structured state to Wildcard with minimal effort from the application
+developers. For example, we have created an initial prototype of a
+plugin for the Redux state management library, which represents the
+state of a user interface as a single object. To configure an integrated
+website adapter for such an existing application, the user simply
+specifies two functions. The first one projects the centralized
+application state into a table, and define handlers for how data edits
+to the table should affect the state.
 
@@
 extended to share live synchronized data with other users. This could be
-achieved through a centralized web server (perhaps an existing service
-like Google Sheets capable of storing tabular data), or through P2P
-connections which might improve the privacy guarantees available for the
-shared annotations.
+achieved through a centralized web server, or through P2P connections
+which might provide stronger privacy guarantees.
 
@@
 latitude and longitude, and the table adapter could dynamically perform
-API queries to populate a result table. (\emph{todo: why is this better
-than spreadsheet formulas? I don't think it offers any more room for
-batching performance optimization because we've already framed formulas
-as whole-column, not row specific})
+API queries to populate a result table. (\emph{todo: make a stronger
+case for why this is useful or better than formulas})
+
+(\emph{note: another ordering for the adapters could be by function.
+First, DOM, AJAX, and Redux, which all integrate with the original app.
+Then, Local Storage, Shared Storage, API, which are all about adding
+functionality to the site. This might make more logical sense, but it
+risks confusing what's already been done and what's a future
+possibility. The current order seems good enough?})
 
@@
 
-\hypertarget{sec:evaluation}{%
-\section{Evaluation}\label{sec:evaluation}}
-
-To determine the viability of writing DOM scraping adapters, we built
-adapters for 10 websites, including transactional sites like Amazon and
-Uber Eats, media consumption sites like Hacker News and Youtube, and a
-medical results entry platform managed by a member of our research
-group. The adapters ranged from 36 to 117 lines of code, averaging 68
-lines. In addition, an external developer unaffiliated with the project
-contributed one adapter, designed to sort the Github page listing a
-user's repositories, and described the experience as ``very
-straightforward.''
-
-Some of the challenges of writing a DOM scraping adapter are the same
-ones involved with writing normal web scraping code, but the more
-interactive nature of Wildcard introduces additional challenges. One
-challenge is triggering updates to the spreadsheet data in response to
-UI changes that happen after initial page load. Site adapters are
-responsible for recognizing these changes by observing the DOM. So far,
-we have been able to use event listeners and the MutationObserver API to
-successfully observe changes, but it may prove challenging to observe
-changes on some sites only through the DOM. Another challenge is
-persisting updates to the DOM---some websites use virtual DOM frameworks
-that can occasionally overwrite changes made by Wildcard. In practice,
-we've managed to create an implementation that works around these issues
-and results in a usable customization experience for all the websites
-we've tried.
-
-\hypertarget{sec:discussion}{%
-\section{Discussion}\label{sec:discussion}}
+\hypertarget{sec:themes}{%
+\section{Key themes}\label{sec:themes}}
 
@@
 
-There have been many attempts at simplifying aspects of traditional
-programming to enable non-programmer end users to perform customization.
-End user scripting languages (AppleScript, VBA, Chickenfoot, CoScripter)
-have innovated at the syntax level, aiming to improve ease-of-use.
-Visual programming environments (Mac Automator, Zapier) go further by
-eliminating text syntax entirely, but maintain the same computational
-model. Programming-by-demonstration and macro recording environments
-(Applescript, Helena, WebVCR) remove some of the initial programming
-burden by starting with concrete demonstrations, but typically require
-programming later in the process.
-
-However, all of these approaches share something in common: \textbf{an
-imperative programming model}. They all involve a list of commands
-executed in sequence, with control flow constructs like conditionals and
-loops, and mutable variables. This makes some intuitive sense: many
-customizations seem most naturally expressed as a series of commands,
-each of which has a side effect on the UI: first enter this value, then
-click this button, etc. The result is that end users eventually need to
-become familiar with all of these traditional imperative programming
-constructs, even if only in diluted form. (todo: bolster the case that
-this is a barrier to learning)
-
-Imperative programming is not the only possible model. Spreadsheets have
-demonstrated another computational model: data editing via direct
-manipulation, combined with pure functional expressions for computation,
-with automatic reactive updates provided by the runtime. We call this a
-data-oriented model, because the user is focused on editing data values
-and producing the correct derived data values, rather than encoding a
-sequence of commands. The success of spreadsheets with end user
-programmers suggests that finding a way to apply data-oriented
-programming model to software customization might enable more people to
-modify their software. However, it's not immediately obvious how the
-spreadsheet paradigm applies to customization.
-
-A key insight from our work is that, rather than representing
-customization as a series of commands issued to an existing application,
-we can present a projected view of the application's internal state, and
-let the user directly edit that projection as a means of customizing the
-application. In some sense, this is similar to the DOM inspector, which
-provides a user with an alternate representation of the UI that they can
-directly manipulate (e.g., by selecting an element and deleting it from
-the DOM tree). But rather than provide a low-level view of the user
-interface elements, we instead provide a higher-level view of the
-semantic state of the application.
-
-Add here:
+\begin{itemize}
+\tightlist
+\item
+  reprise intro--explain direct manipulation
+\item
+  the key is that the DATA is being direct manipulated, not a script
+\item
+  it's essentially an alternate UI for an existing app, not a scripting
+  env
+\item
+  argue for the benefit of multiple representations
+  \citep{ainsworth1999}
 
@@
   \item
-  multiple representations \citep{ainsworth1999}
+    abstraction
+  \item
+    different affordances
+  \end{itemize}
 \item
@@
   interaction
+\end{itemize}
+
+\hypertarget{gentle-slope}{%
+\subsection{Gentle slope}\label{gentle-slope}}
+
+Term originally came from \citep{maclean1990}
+
+\begin{itemize}
+\tightlist
+\item
+  \citep{maclean1990} says ``some customizations should be as easy as
+  using the app'', we have achieved that
+\item
+  here's where ``automation vs customization'' comes in: automation
+  frames it as entering a whole new scripting environment; we frame it
+  more as just an alternate UI that you can use.
+\item
+  backend APIs: terrible
+\item
+  the best would be right inline with the GUI (cite Scotty, instrumental
+  work) but this has its own problems
+\item
+  we settle for a compromise: an alternate structured view. (explicitly
+  contrast this clearly)
+\item
+  the CLI GUI thing
 \end{itemize}
@@
 
-(\emph{Footnote this paragraph?}) This categorization is slightly
-oversimplified, and there are existing customization ecosystems that
-span both categories. For example, part of the success of browser
-extensions stems from the fact that the DOM encourages the use of
-standardized semantic UI elements; even if an app developer doesn't
-anticipate customization, merely using of semantic and accessible HTML
-creates a collateral benefit of easier extension. Another example of a
-middle ground is AppleScript, which provides system-wide low-level APIs
-for manipulating GUIs as a means of black-box customization, while also
-enabling developers to optionally add application-specific APIs to add
-support for semantic customization.
-
 With Wildcard, we use a hybrid approach that takes the best of both
@@
 implemented as a black-box customization, but externally provides a
-semantic interface to the application. End users get an ergonomic and
-simplified customization experience, but without the need to depend on a
-first-party application developer exposing extension APIs.
+semantic interface to the application. These wrappers are added to a
+shared repository, available to all users of the system. When an end
+user is using a site that already has an adapter, they get a semantic
+customization experience that avoids low-level details.
 
 One way to view this approach is as introducing a new abstraction
-barrier into black-box extension. Typically, a black box customization
-script combines two responsibilities: 1) mapping the low-level details
-of a user interface to semantic constructs (e.g., using CSS selectors to
-find certain page elements), and 2) the actual logic of the specific
-customization. (todo: could easily show examples of this from browser
-extensions, Chickenfoot, etc) Even though the mapping logic is often
-more generic than the specific customization (e.g., finding a given
-input element is independent of what text to insert into that element),
-the intertwining of these two responsibilities in a single script makes
-it very difficult to share the mapping logic across scripts. With
-Wildcard we propose a decoupling of these two layers: a
+barrier into third-party extension. Typically, a third party
+customization script combines two responsibilities: 1) mapping the
+low-level details of a user interface to semantic constructs (e.g.,
+using CSS selectors to find certain page elements), and 2) the actual
+logic of the specific customization. (todo: could easily show examples
+of this from browser extensions, Chickenfoot, etc) Even though the
+mapping logic is often more generic than the specific customization, the
+intertwining of these two responsibilities in a single script makes it
+very difficult to share the mapping logic across scripts.
+
+With Wildcard we propose a decoupling of these two layers: a
 community-maintained mapping layer, shared across many specific
@@
 
-\hypertarget{gentle-slope}{%
-\subsection{Gentle slope}\label{gentle-slope}}
+\hypertarget{sec:evaluation}{%
+\section{Evaluation}\label{sec:evaluation}}
 
-Originally came from \citep{maclean1990}
+We have used Wildcard to build customizations for many real websites. So
+far, most usage has come from members of the project team, although we
+have had a few external users try the system. Here we offer our
+reflections on using the system, focused on two key questions:
 
@@
 \item
-  here's where ``automation vs customization'' comes in: automation
-  frames it as entering a whole new scripting environment; we frame it
-  more as just an alternate UI that you can use.
+  What is the range of useful customizations possible in this paradigm?
 \item
-  this is the heart of the gentle slope
-\item
-  backend APIs: terrible
-\item
-  the best would be right inline with the GUI (cite Scotty, instrumental
-  work) but this has its own problems
-\item
-  we settle for a compromise: an alternate structured view. (explicitly
-  contrast this clearly)
-\item
-  the CLI GUI thing
+  How feasible is it to build DOM scraping adapters for real websites,
+  in practice?
 \end{itemize}
+
+\hypertarget{range-of-customizations}{%
+\subsection{Range of customizations}\label{range-of-customizations}}
+
+\emph{todo}
+
+\hypertarget{viability-of-dom-scraping}{%
+\subsection{Viability of DOM scraping}\label{viability-of-dom-scraping}}
+
+In order for our idea of third-party customization through Wildcard to
+succeed, it is important that creating usable adapters for existing
+websites takes minimal effort.
+
+To test the viability of writing DOM scraping adapters, we built
+adapters for 10 websites, including transactional sites like Amazon and
+Uber Eats, media consumption sites like Hacker News and Youtube, and a
+medical results entry platform managed by a member of our research
+group. The adapters ranged from 36 to 117 lines of code, averaging 68
+lines. In addition, an external developer unaffiliated with the project
+contributed one adapter, designed to sort the Github page listing a
+user's repositories, and described the experience as ``very
+straightforward.''
+
+Some of the challenges of writing a DOM scraping adapter are the same
+ones involved with writing normal web scraping code, but the more
+interactive nature of Wildcard introduces additional challenges. One
+challenge is triggering updates to the spreadsheet data in response to
+UI changes that happen after initial page load. Site adapters are
+responsible for recognizing these changes by observing the DOM. So far,
+we have been able to use event listeners and the MutationObserver API to
+successfully observe changes, but it may prove challenging to observe
+changes on some sites only through the DOM. Another challenge is
+persisting updates to the DOM---some websites use virtual DOM frameworks
+that can occasionally overwrite changes made by Wildcard. In practice,
+we've managed to create an implementation that works around these issues
+and results in a usable customization experience for all the websites
+we've tried.
 

### 2020-05-21 paper.tex
dm section

@@
 
-Table-driven customization embodies several design principles for
-building customizable software systems, discussed in
-Section~\ref{sec:themes}:
+In Section~\ref{sec:themes}, we discuss some key themes from our work:
 
@@
 \item
-  \emph{Third-party semantic wrappers}: Typically, customization tools
-  which don't rely on official extension APIs resort to offering
-  low-level APIs for customization. Instead, we suggest a
-  community-maintained library of semantic wrappers around existing
-  applications, offering end users a more convenient semantic API for
-  customization.
-\item
-  \emph{Gentle slope}: Customization tools should ideally offer a
-  ``gentle slope'' from normal use to customization. We explore how our
-  approach meets this goal by making some software customizations as
-  easy as normal use of the software.
+  \emph{Third-party semantic wrappers}: Typically, tools that don't rely
+  on official extension APIs resort to offering low-level APIs for
+  customization. Instead, we suggest a community-maintained library of
+  semantic wrappers around existing applications, enabling end users to
+  work with domain objects rather than low-level representations.
 \end{itemize}
@@
 directly click on the correct date, and it gets inserted into both the
-spreadsheet and the original form.
+spreadsheet and the original form.n
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  reprise intro--explain direct manipulation
-\item
-  the key is that the DATA is being direct manipulated, not a script
-\item
-  it's essentially an alternate UI for an existing app, not a scripting
-  env
-\item
-  argue for the benefit of multiple representations
-  \citep{ainsworth1999}
+\hypertarget{the-programming-cliff}{%
+\subsubsection{The programming cliff}\label{the-programming-cliff}}
 
-  \begin{itemize}
-  \tightlist
-  \item
-    abstraction
-  \item
-    different affordances
-  \end{itemize}
-\item
-  address Tchernavskij's point about contrasting with instrumental
-  interaction
-\end{itemize}
+Hutchins, Hollan and Norman \citep{hutchins1985} describe a direct
+manipulation interface as one that uses a model-world metaphor, rather
+than a conversation metaphor. Rather than conversing with the system
+about an assumed world, ``the world is explicitly represented'' and the
+user can ``{[}act{]} upon the objects of the task domain themselves.''
 
-\hypertarget{gentle-slope}{%
-\subsection{Gentle slope}\label{gentle-slope}}
+Most GUIs today employ direct manipulation, but software customization
+tools typically use an imperative programming model, which implements
+the conversational metaphor rather than direct manipulation. For
+example, in Applescript \citep{cook2007}, the scripting language for
+customizing Mac OS applications, here is how a user retrieves a list of
+of calendar names from the Calendar application:
 
-Term originally came from \citep{maclean1990}
+\begin{verbatim}
+tell application "Calendar"
+  name of calendars
+end tell
+\end{verbatim}
 
-\begin{itemize}
-\tightlist
-\item
-  \citep{maclean1990} says ``some customizations should be as easy as
-  using the app'', we have achieved that
-\item
-  here's where ``automation vs customization'' comes in: automation
-  frames it as entering a whole new scripting environment; we frame it
-  more as just an alternate UI that you can use.
-\item
-  backend APIs: terrible
-\item
-  the best would be right inline with the GUI (cite Scotty, instrumental
-  work) but this has its own problems
-\item
-  we settle for a compromise: an alternate structured view. (explicitly
-  contrast this clearly)
-\item
-  the CLI GUI thing
-\end{itemize}
+Some customization environments (Mac Automator, Zapier) forego text
+syntax and enable the user to connect programs and construct automations
+by dragging and dropping commands. These environments still do not
+constitute direct manipulation, though: the objects being manipulated
+are in the domain of programming, not the domain of the task at hand.
+
+It is a perfectly unreasonable choice to choose imperative programming
+as the model for building customizations. Imperative programming
+provides a high ceiling for possible customizations, and the model of a
+sequence of commands is a natural fit for automation workflows which
+simulate a series of steps taken by the user. However, there is a
+serious drawback to this approach.
+
+MacLean et al \citep{maclean1990} describe an ideal for user-tailorable
+systems: a ``gentle slope'' from using to customizing, where small
+incremental increases in skill lead to corresponding increments of
+customization power. Requiring users to learn programming for
+customization creates an abrupt ``cliff,'' where a significant learning
+investment is required even to implement simple customizations. Another
+goal of MacLean et al is that ``it should be as easy to change the
+environment as it is to use it,'' at least for some subset of changes.
+In scripting languages, even user-friendly ones like Applescript or
+Chickenfoot, the experience of customization does not remotely resemble
+the experience of use, so these systems can't meet this goal.
+
+\hypertarget{a-gentler-slope}{%
+\subsubsection{A gentler slope}\label{a-gentler-slope}}
+
+With table-driven customization we aim to provide a gentler slope, by
+using direct manipulation for software customization. The data shown in
+the table view is the domain data from the original application. The
+user makes changes to the data by selecting areas of interest in the
+table, e.g.~sorting/filtering by clicking the relevant column header, or
+adding annotation by clicking on the relevant row. These interactions
+are common in GUI applications, and Wildcard therefore meets MacLean et
+al's goal: some one-click customizations are as easy as using the
+original application.
+
+One aspect of directness which we have chosen not to maintain in
+Wildcard is enabling customization in the context of the original user
+interface, as explored by some other tools like Scotty
+\citep{eagan2011}. We have found that augmenting the original UI with an
+additional structured representation has numerous benefits. It provides
+a consistent experience across all applications, and clearly shows the
+user what structured data is available to work with.
+
+Ainsworth et al's provide a helpful taxonomy of the value of multiple
+representations \citep{ainsworth1999}. In their terms, Wildcard plays a
+\emph{complementary role} by supporting a different set of tasks from
+the original application, while displaying \emph{shared information}. We
+also suspect that Wildcard helps \emph{constructs deeper understanding}
+by \emph{subtraction}. By stripping away details and only showing the
+essential data in an interface, Wildcard helps users think of new ways
+of using that data, outside the specific constraints of the original
+application.
+
+It is important to note that there are many specific customizations that
+can be achieved in systems like Chickenfoot \citep{bolin2005} that
+cannot be reproduced in Wildcard. We consider this an acceptable
+tradeoff in exchange for achieving a gentler slope of customization, and
+we show in Section~\ref{sec:examples} and Section~\ref{sec:evaluation}
+that our model can still implement many useful customizations in
+practice. Also, there is sometimes a way to reframe an imperative script
+in terms of our direct manipulation model; for example, a script that
+iterates through rows in a page adding some additional information could
+be reproduced using a formula in Wildcard.
 
@@
 
-\textbf{Third-party customization tools} enable customization without
+\emph{Third-party customization tools} enable customization without
 using official extension APIs, enabling a broader range of
@@
 
-\textbf{Semantic customization tools} use explicit extension APIs
-provided by the application developer. Examples of this include
-accessing a backend web API, or writing a customization in Applescript
-that uses an application-specific API. The main benefit is that this
-allows the extension author to work with meaningful concepts in the
-application domain---``create a new calendar event'' rather than ``click
-the button that says new event.''---which makes customizations easier to
-build and more robust. However, this style limits the range of
-extensions that can be built, to only those which the official plugin
-API supports.
+\emph{Semantic customization tools} use explicit extension APIs provided
+by the application developer. Examples of this include accessing a
+backend web API, or writing a customization in Applescript that uses an
+application-specific API. The main benefit is that this allows the
+extension author to work with meaningful concepts in the application
+domain---``create a new calendar event'' rather than ``click the button
+that says new event.''---which makes customizations easier to build and
+more robust. However, this style limits the range of extensions that can
+be built, to only those which the official plugin API supports.
 
@@
 using CSS selectors to find certain page elements), and 2) the actual
-logic of the specific customization. (todo: could easily show examples
-of this from browser extensions, Chickenfoot, etc) Even though the
-mapping logic is often more generic than the specific customization, the
-intertwining of these two responsibilities in a single script makes it
-very difficult to share the mapping logic across scripts.
+logic of the specific customization. Even though the mapping logic is
+often more generic than the specific customization, the intertwining of
+these two responsibilities in a single script makes it very difficult to
+share the mapping logic across scripts.
 
@@
 persisting updates to the DOM---some websites use virtual DOM frameworks
-that can occasionally overwrite changes made by Wildcard. In practice,
-we've managed to create an implementation that works around these issues
-and results in a usable customization experience for all the websites
-we've tried.
+that can occasionally overwrite changes made by Wildcard. Despite these
+challenges, we've managed to create an implementation that works around
+these issues in practice, and results in a usable customization
+experience for all the websites we've tried so far.
 
@@
 This paper extends a workshop paper by Litt and Jackson \citep{litt2020}
-which presented an earlier prototype of the Wildcard system. We build on
-that work in this paper by rearchitecting the internal implementation of
-the system around the table adapter abstraction, evaluating it on many
-more websites, and by more fully characterizing the behavior and design
-of the system.
+which presented an earlier prototype version of the Wildcard system. We
+build on their work in this paper by rearchitecting the internal
+implementation of the system around the table adapter abstraction,
+evaluating it on many more websites, and by more fully characterizing
+the behavior and design of the system.
 
@@
 \begin{itemize}
-\tightlist
+\item
+  Buttons \citep{maclean1990}: the original DM customization tool
+\item
+  ScrAPIr: direct manipulation for APIs!
 \item
@@
   \begin{itemize}
-  \tightlist
   \item
     browser extensions
-  \item
-    ScrAPIr
   \item

### 2020-05-21 paper.tex
compile

@@
 
-\hypertarget{the-programming-cliff}{%
-\subsubsection{The programming cliff}\label{the-programming-cliff}}
-
 Hutchins, Hollan and Norman \citep{hutchins1985} describe a direct
@@
 
-It is a perfectly unreasonable choice to choose imperative programming
-as the model for building customizations. Imperative programming
-provides a high ceiling for possible customizations, and the model of a
-sequence of commands is a natural fit for automation workflows which
-simulate a series of steps taken by the user. However, there is a
-serious drawback to this approach.
+It is reasonable to choose imperative programming as the model for
+building customizations. Turing-Complete programming provides a high
+ceiling for possible customizations, and the a sequence of commands is a
+natural fit for automations which simulate a series of steps taken by
+the user.
 
-MacLean et al \citep{maclean1990} describe an ideal for user-tailorable
-systems: a ``gentle slope'' from using to customizing, where small
-incremental increases in skill lead to corresponding increments of
-customization power. Requiring users to learn programming for
-customization creates an abrupt ``cliff,'' where a significant learning
-investment is required even to implement simple customizations. Another
-goal of MacLean et al is that ``it should be as easy to change the
-environment as it is to use it,'' at least for some subset of changes.
-In scripting languages, even user-friendly ones like Applescript or
-Chickenfoot, the experience of customization does not remotely resemble
-the experience of use, so these systems can't meet this goal.
-
-\hypertarget{a-gentler-slope}{%
-\subsubsection{A gentler slope}\label{a-gentler-slope}}
+However, there is a serious drawback to this approach. MacLean et al
+\citep{maclean1990} describe an ideal for user-tailorable systems: a
+``gentle slope'' from using to customizing, where small incremental
+increases in skill lead to corresponding increments of customization
+power. Requiring users to learn programming for customization creates an
+abrupt ``cliff,'' where a significant learning investment is required
+even to implement simple customizations. Another goal of MacLean et al
+is that ``it should be as easy to change the environment as it is to use
+it,'' at least for some subset of changes. In scripting languages, even
+user-friendly ones like Applescript or Chickenfoot, the experience of
+customization does not remotely resemble the experience of use, so these
+systems can't meet this goal.
 

### 2020-05-21 paper.tex
dnj edits

@@
 \begin{abstract}
-  In this paper we propose \emph{table-driven customization,} a new
-  paradigm for enabling end users to customize software applications
-  without doing traditional programming\emph{.} Users directly
-  manipulate a tabular view of the structured data inside the
-  application, rather than writing imperative scripts as in most
-  customization tools. We extend this simple model with a spreadsheet
-  formula language and custom data editing widgets, which provide
-  sufficient expressivity to implement many useful customizations.
+  In this paper we show how the behavior of a software application can
+  be extended and adapted by direct manipulation, using table-driven
+  customization, a new paradigm that allows end users to customize
+  applications without writing traditional code.
 
-  We describe Wildcard, a browser extension which implements
-  table-driven customization in the context of web applications. Through
-  concrete examples, we demonstrate that this paradigm can be used to
-  create useful customizations for real applications. We share
-  reflections from our usage of the Wildcard system, including its
-  strengths and limitations relative to other customization approaches.
-  We further explore how new software architectures might encourage
-  application developers to promote this style of end-user
-  customization.
+  Instead, users directly manipulate a tabular view of the structured
+  data inside the application---rather than writing imperative scripts
+  (as in most customization tools). This simple model also accommodates
+  a spreadsheet formula language and custom data editing widgets, which
+  provide sufficient expressivity to implement many useful
+  customizations.
+
+  We illustrate the approach with Wildcard, a browser extension that
+  implements table-driven customization in the context of web
+  applications. Through concrete examples, we show that our paradigm can
+  be used to create useful customizations for real applications. We
+  share reflections from experience using the Wildcard system, on both
+  its strengths and limitations relative to other customization
+  approaches. Finally, we explore how this paradigm might lead to new
+  software architectures that encourage this form of customization.
 \end{abstract}
@@
 
-There have been many attempts at empowering end users to customize their
-software by offering them a simplified programming tool. Some scripting
-languages (AppleScript, Chickenfoot) have a friendly syntax that
-resembles natural language. Visual programming tools (Mac Automator,
-Zapier) eliminate text syntax entirely. Macro recorders (Applescript,
-Helena, WebVCR) remove some of the initial programming burden by letting
-a user start by concrete demonstrations.
+Most attempts at empowering end users to customize their software offer
+a simplified version of programming. Some scripting languages
+(AppleScript, Chickenfoot) have a friendly syntax that resembles natural
+language. Visual programming tools (Mac Automator, Zapier) eliminate
+text syntax entirely. Macro recorders (Applescript, Helena, WebVCR)
+remove some of the initial programming burden by letting a user start
+with concrete demonstrations.
 
-These approaches have many differences, but they all share something in
-common: an imperative programming model, with mutable variables,
-conditionals and loops. End users ultimately must use these traditional
-programming constructs to express their ideas, and the object of
-interest is a \emph{script}, a sequence of commands.
+Despite their many differences, these approaches all share something in
+common: an imperative programming model, with statement sequencing,
+mutable variables and loops. End users express their ideas in
+scripts---sequences of commands---which, name aside, are not very
+different from conventional code.
 
@@
 of interest'' replaces ``complex command language syntax''. Direct
-manipulation is the de facto standard in GUIs today, but when it comes
-to customizing software, it is rarely to be found. In this work, we ask:
-what would it look like to build a software customization interface that
-relies on direct manipulation? We take inspiration from spreadsheets and
-visual database query interfaces \citep{2020a, bakke2016}, which have
-successfully enabled end users to run queries and computations through
-direct manipulation of data.
+manipulation is the \emph{de facto} standard in GUIs today, but when it
+comes to customizing software, it is rarely to be found. In this work,
+we ask: what would it look like to build a software customization
+interface that relies on direct manipulation? We take inspiration from
+spreadsheets and visual database query interfaces
+\citep{2020a, bakke2016}, which have successfully enabled end users to
+run queries and computations through direct manipulation of data.
 
 In this paper we present a technique called \emph{table-driven
-customization} which applies these ideas from visual query interfaces in
-the context of software customization. We augment an application's UI
-with a table view, where the user can see and manipulate the structured
-data inside an application. Changes in the table view result in
-immediate corresponding changes to the original user interface of the
-application, enabling the user to customize an application with live
-feedback.
+customization}, which applies ideas from visual query interfaces in the
+context of software customization. An application's UI is augmented with
+a table view, where the user can see and manipulate the application's
+internal data . Changes in the table view result in immediate
+corresponding changes to the original user interface of the application,
+enabling the user to customize an application with live feedback.
 
@@
 scraping techniques to implement table-driven customization for existing
-Web applications. In Section~\ref{sec:examples} we concretely introduce
-the key ideas of table-driven customization by presenting several
-examples of real customizations implemented in Wildcard.
+Web applications. In Section~\ref{sec:examples}, we introduce the key
+ideas of table-driven customization by presenting several examples of
+real customizations implemented in Wildcard.
 
-In Section~\ref{sec:architecture} we explain the architecture of
+In Section~\ref{sec:architecture}, we explain the architecture of
 table-driven customization. We focus on the \emph{table adapter}
-abstraction, which allows many different types of underlying state to be
-bidirectionally mapped to a table of data. We describe certain types of
-table adapters we've built in Wildcard; we also describe future adapters
-that could be supported in the general paradigm.
+abstraction, which allows many different types of underlying data to be
+bidirectionally mapped to a table. We describe several types of table
+adapters we've built in Wildcard, and also describe future adapters that
+are supported by the general paradigm.
 
 We have used Wildcard to build real customizations for 11 different
-websites. In Section~\ref{sec:evaluation} we present reflections from
-this process: customizations we were able to build, limitations we
-encountered, and reflections on the ease of integrating scraping logic
-with real websites.
+websites. In Section~\ref{sec:evaluation}, we present reflections from
+this process, outlining the kinds of customizations we were able to
+build, limitations we encountered, and reflections on the ease of
+integrating scraping logic with real websites.
 
@@
   \emph{Customization by direct manipulation}: End users should be able
-  to customize an application by examining and modifying its data,
-  rather than by writing imperative scripts.
+  to customize an application by directly examining and modifying its
+  data, rather than by writing imperative scripts.
 \item
-  \emph{Third-party semantic wrappers}: Typically, tools that don't rely
-  on official extension APIs resort to offering low-level APIs for
-  customization. Instead, we suggest a community-maintained library of
-  semantic wrappers around existing applications, enabling end users to
-  work with domain objects rather than low-level representations.
+  \emph{Wrapping applications for customization}: Typically, tools that
+  don't rely on official extension APIs resort to offering low-level
+  APIs for customization. Instead, we propose a community-maintained
+  library of semantic wrappers around existing applications, enabling
+  end users to work with domain objects rather than low-level
+  representations.
 \end{itemize}
@@
 implements web scraping logic to produce a table of data from the web
-page, and edits to the table result in DOM manipulations like reordering
-rows of data on the page.
+page, and turns edits to the table into DOM manipulations like
+reordering rows of data on the page.
 
-Ultimately, we provide the user with a single combined table view that
-they can edit, rather than three separate tables. The \textbf{query
-engine} is responsible for combining the tables of data into this single
-view, and routing the user's edits back to the individual table
-adapters. In this example, the query engine has joined the three tables
-together by a shared ID column, and sorted the result by the name
-column.
+The three separate tables are then combined into a single table for the
+end user to view and edit. The \textbf{query engine} is responsible for
+creating this combined view, and routing the user's edits back to the
+individual table adapters. In this example, the query engine has joined
+the three tables together by a shared ID column, and sorted the result
+by the name column.
 
@@
 
-A key idea in table-driven customization is that different data sources
-can all be mapped to the generic abstraction of an editable table. In a
-relational database, the table matches the underlying storage format,
-but in table-driven customization, the table is merely an
+A key idea in table-driven customization is that a wide variety of data
+sources can be mapped to the generic abstraction of an editable table.
+In a relational database, the table matches the underlying storage
+format, but in table-driven customization, the table is merely an
 \emph{interface layer}. The data shown in the table is a projection of
-some underlying state, and edits to the table can have arbitrary effects
+some underlying state, and edits to the table can have complex effects
 on the underlying state.
@@
 Externally, a table adapter must satisfy an abstract interface. The
-first two parts of the interface largely resemble the interface exposed
-by a table in a relational database:
+first two parts of the interface resemble the interface exposed by a
+table in a relational database:
 
 \textbf{Returning a table}: A table adapter exposes a table of data: an
-ordered list of records. Each record contains associates attributes and
-values. Tables have a typed schema, so the same attributes are shared
-across all records. A table adapter can update the contents of a table
-at any time, in response to changes in the underlying state (e.g., a DOM
-scraping adapter can update the table when the page body changes). New
-data is pushed out to other components of the system, and the query view
-reactively updates in response.
+ordered list of records. Each record carries a unique identifier and
+associates named attributes with values. Tables have a typed schema, so
+the same attributes are shared across all records. A table adapter can
+update the contents of a table at any time, in response to changes in
+the underlying state (e.g., a DOM scraping adapter can update the table
+when the page body changes). When data changes, the query view is
+reactively updated in response.
 
-\textbf{Making edits}: The query engine can request to a table adapter
-to make an edit to a record. The meaning of making an edit can vary
-depending on the adapter: in the local storage adapter, an annotation
-can be persisted into local storage; in the DOM scraping adapter, an
-edit can represent filling in a form field. An adapter can also mark
-values as read-only if it wouldn't be meaningful to edit them; for
-example, the DOM scraping adapter typically marks page content as
+\textbf{Making edits}: The query engine can issue a request to a table
+adapter to make an edit to a record. The meaning of making an edit can
+vary depending on the adapter: in the local storage adapter, an
+annotation may be persisted into local storage; in the DOM scraping
+adapter, an edit may represent filling in a form field. An adapter can
+also mark values as read-only if it wouldn't be meaningful to edit them;
+for example, the DOM scraping adapter typically marks page content as
 read-only, except for editable form fields.
@@
 
-Here we describe in more detail the table adapters which we have
+Here we describe in more detail the table adapters that we have
 implemented in Wildcard to power the customizations shown in
@@
 standard web scraping problem of extracting a table of data from the
-DOM, a scraping adapter must also manipulate the DOM to re-order rows,
+DOM, a scraping adapter must also manipulate the DOM to reorder rows,
 edit form entries, and inject annotations as the table is edited.
@@
 available for customization in Wildcard. To make this approach viable,
-we have built library functions that make it as easy as possible to
-implement a scraping adapter, only requiring the programmer to implement
-the minimal site-specific parts. The programmer writes a single scraping
-function which uses CSS selectors and DOM APIs to extract and return
-relevant elements from the page. The library functions then wrap this
-function to implement the rest of the needed functionality; for example,
-when the table is sorted, we find the DOM elements corresponding to the
-rows in the table, remove them from the page, and then reinsert them in
-the new order.
+we have built a library of functions that make it easy to implement a
+scraping adapter, requiring the programmer only to implement the minimal
+site-specific parts. The programmer writes a single Javascript function
+which can use standard scraping techniques like CSS selectors and DOM
+APIs to extract relevant elements from the page. The library functions
+then wrap this function to implement the rest of the needed
+functionality; for example, when the table is sorted, we find the DOM
+elements corresponding to the rows in the table, remove them from the
+page, and then reinsert them in the new order.
 
@@
 structured form, and often includes information not shown in the UI. As
-with DOM scraping adapters, we have made it as easy as possible for
-programmers to create site-specific AJAX scraping adapters. A programmer
-writes a function which specifies how to extract data from an AJAX
-request, and the framework handles the details of actually intercepting
-requests and calling the programmer-defined function.\footnote{So far we
-  have only implemented AJAX scraping in the Firefox version of
-  Wildcard, since Firefox has convenient APIs for intercepting requests.
-  It appears possible to implement in Chrome as well, but we have not
-  finished our implementation.}
+with DOM scraping adapters, we have made it easy for programmers to
+create site-specific AJAX scraping adapters. A programmer writes a
+function that specifies how to extract data from an AJAX request, and
+the framework handles the details of intercepting requests and calling
+the programmer-defined function.\footnote{So far we have only
+  implemented AJAX scraping in the Firefox version of Wildcard, since
+  Firefox has convenient APIs for intercepting requests. It appears
+  possible to implement in Chrome as well, but we have not finished our
+  implementation.}
 
@@
 browser. As shown in the example use cases, this can be useful for
-privately persisting annotations, without needing to upload them to a
-web service.
+persisting annotations in a private way, without uploading them to a web
+service.
 
@@
 We have designed the table adapter API to be general enough to support
-other types of useful adapters in the future. Here are three examples of
-future possibilities:
+other types of useful adapters in the future. Here are two such
+possibilities:
 
@@
 first party developers adding support directly into their own websites.
-A DOM Scraping table adapter could be swapped out for an ``integrated
-website adapter'' which directly accesses the internal state of the
-application, without needing to extract it from the UI.
+An ``integrated website adapter'' installed by the developer could
+directly accesses the internal state of the application, providing the
+same functionality as a DOM scraping adapter but in a more robust way.
 
@@
 state is now often available in the web client. We suspect it is
-possible to create plugins for frontend frameworks which expose
-structured state to Wildcard with minimal effort from the application
-developers. For example, we have created an initial prototype of a
-plugin for the Redux state management library, which represents the
-state of a user interface as a single object. To configure an integrated
-website adapter for such an existing application, the user simply
-specifies two functions. The first one projects the centralized
-application state into a table, and define handlers for how data edits
-to the table should affect the state.
+possible to create plugins for frontend frameworks that expose this
+state to Wildcard with only minimal effort from the application
+developers. To test this hypothesis, we have created an early prototype
+of a Wildcard adapter for the Redux state management library, and used
+it with the TodoMVC todo list application. To install the adapter, the
+programmer only needs to make a few small changes to their app:
+
+\emph{Exposing a table}: In Redux, app developers already represent the
+state of a user interface as a single object. To expose a table of data,
+the developer simply writes a function that maps the state object to a
+table. \emph{Editing data values}: The developer defines how edits to
+records in the table should map to edits in the state object.
+\emph{Sorting/filtering the UI}: The developer inserts an additional
+data processing step before the state object is rendered in the UI, so
+that Wildcard can apply sorting and filtering. \emph{Annotations}: The
+developer adds additional React components managed by Wildcard into
+their UI tree, which render additional data as annotations.
+
+In certain types of software, customizability is often touted as a key
+selling point. An integrated website adapter would provide a way for
+developers to integrate with an existing ecosystem of formulas and
+customization tools, without needing to build all that functionality
+from scratch.
 
 \textbf{Shared storage adapter}: It would be useful to share user
-annotations between people and across devices---for example,
-collaboratively taking notes with friends on a list of options for
-places to stay on Airbnb. The existing Local Storage Adapter could be
-extended to share live synchronized data with other users. This could be
-achieved through a centralized web server, or through P2P connections
-which might provide stronger privacy guarantees.
-
-\textbf{Third party API data adapter}: Currently, the main mechanism for
-including data from web APIs in Wildcard is using spreadsheet formulas.
-However, a web API could also be wrapped to expose a table API that
-would dynamically create tables in response to queries. For example,
-when fetching walkability scores for many GPS locations, the query
-engine could request a table of walkability scores for various pairs of
-latitude and longitude, and the table adapter could dynamically perform
-API queries to populate a result table. (\emph{todo: make a stronger
-case for why this is useful or better than formulas})
-
-(\emph{note: another ordering for the adapters could be by function.
-First, DOM, AJAX, and Redux, which all integrate with the original app.
-Then, Local Storage, Shared Storage, API, which are all about adding
-functionality to the site. This might make more logical sense, but it
-risks confusing what's already been done and what's a future
-possibility. The current order seems good enough?})
+annotations among users and across devices---for example,
+collaboratively taking notes with friends on options for places to stay
+on Airbnb. The existing Local Storage Adapter could be extended to share
+live synchronized data with other users. This could be achieved through
+a centralized web server, or through P2P connections that might provide
+stronger privacy guarantees.
 
@@
 
-First, every query involves a primary DOM scraping table adapter which
-associates records in the result with elements in the application's user
-interface. At minimum, the primary table adapter needs to return record
-IDs and have the ability to manipulate the application's UI. It can also
-optionally return data about each record.
+Queries are processed in three steps. First, the query invokes a primary
+DOM scraping table adapter that associates records in the result with
+elements in the application's user interface. At minimum, the primary
+table adapter needs to return record IDs and have the ability to
+manipulate the application's UI. It can also optionally return data
+about each record. Next, additional tables (AJAX data, local storage
+data) are left-joined by ID. Finally, the result table can be sorted and
+filtered by any column.
 
-Next, additional tables (AJAX data, local storage data) are left joined
-by ID. (\emph{todo: discuss IDs here?}) Finally, the result table can be
-sorted and filtered by any column.
-
-One way to think of this model is a tiny constrained subset of the SQL
-query model. We've found that this simple model has proven sufficient
-for meeting the needs of customization in practice, and minimizes the
-complexity of supporting more general and arbitrary queries. But because
-it fits into the general SQL paradigm, it could theoretically be
-extended to support more types of queries.
+This query model can be viewed as a tiny (and rather constrained) subset
+of the SQL query model. Despite its simplicity, this simple model has
+proven sufficient for meeting the needs of customization in practice,
+and minimizes the complexity of supporting more general and arbitrary
+queries. But because it fits into the general paradigm of relational
+queries, it could theoretically be extended to support a wider range of
+queries.
 
 The query engine is also responsible for executing formulas. We have
-built a small language resembling the formula languages in visual query
+built a small formula language resembling the ones used in visual query
 tools like SIEUFERD. As in those tools, and unlike in spreadsheets,
-formulas automatically apply to an entire column of data, and can
-reference columns rather than only specific cells.
+formulas automatically apply across an entire column of data, and
+reference other column names instead of values on specific rows. For
+example, the user would write \texttt{=price\ *\ 2} to create a column
+with a value derived from a price attribute. This is more convenient
+than needing to copy-paste a formula across an entire column as in
+spreadsheets, and none of our customization use cases have uncovered a
+need for writing different formulas on different rows of the table.
 
@@
 
-Ainsworth et al's provide a helpful taxonomy of the value of multiple
+Ainsworth et al provide a helpful taxonomy of the value of multiple
 representations \citep{ainsworth1999}. In their terms, Wildcard plays a
@@
 iterates through rows in a page adding some additional information could
-be reproduced using a formula in Wildcard.
+be reproduced using a formula in Wildcard. (\emph{todo: mention the fact
+that the data abstraction step limits possibilities})
 
-\hypertarget{third-party-semantic-wrappers}{%
-\subsection{Third-party semantic
-wrappers}\label{third-party-semantic-wrappers}}
+\hypertarget{wrapping-applications-for-customization}{%
+\subsection{Wrapping applications for
+customization}\label{wrapping-applications-for-customization}}
 
@@
 
-\emph{Third-party customization tools} enable customization without
-using official extension APIs, enabling a broader range of
-customizations on top of more applications. For example, web browser
-extensions have demonstrated the utility of customizing websites through
-manipulating the DOM, without needing explicit extension APIs to be
-built in. However, third-party customization comes with a corresponding
-cost: these tools can typically only operate at a low level of
-abstraction, e.g.~manipulating user interface elements. This makes it
-harder for end users to write scripts, and makes the resulting scripts
-more brittle. (todo: support this claim more, provide examples)
+\emph{Ad hoc customization tools} enable customization without using
+official extension APIs, enabling a broader range of customizations on
+top of more applications. For example, web browser extensions have
+demonstrated the utility of customizing websites through manipulating
+the DOM, without needing explicit extension APIs to be built in.
+However, ad hoc customization comes with a corresponding cost: these
+tools can typically only operate at a low level of abstraction,
+e.g.~manipulating user interface elements. This makes it harder for end
+users to write scripts, and makes the resulting scripts more brittle.
 
-\emph{Semantic customization tools} use explicit extension APIs provided
-by the application developer. Examples of this include accessing a
-backend web API, or writing a customization in Applescript that uses an
-application-specific API. The main benefit is that this allows the
-extension author to work with meaningful concepts in the application
-domain---``create a new calendar event'' rather than ``click the button
-that says new event.''---which makes customizations easier to build and
-more robust. However, this style limits the range of extensions that can
-be built, to only those which the official plugin API supports.
+\emph{Anticipated customization tools}, in contrast, use explicit
+extension APIs provided by the application developer. Examples of this
+include accessing a backend web API, or writing a customization in
+Applescript that uses an application-specific API. The main benefit is
+that this allows the extension author to work with meaningful concepts
+in the application domain---``create a new calendar event'' rather than
+``click the button that says new event.''---which makes customizations
+easier to build and more robust. However, this style limits the range of
+extensions that can be built to those that the official plugin API
+supports.
 
-With Wildcard, we use a hybrid approach that takes the best of both
-worlds. Programmers implement an API wrapper that is internally
-implemented as a black-box customization, but externally provides a
-semantic interface to the application. These wrappers are added to a
-shared repository, available to all users of the system. When an end
-user is using a site that already has an adapter, they get a semantic
-customization experience that avoids low-level details.
+With Wildcard, we use a hybrid approach that aims to provide the best of
+both worlds. Programmers implement an API wrapper that is internally
+implemented as an ad hoc customization, but externally provides a
+high-level interface to the application, abstracting away the details of
+the user interface. These wrappers are added to a shared repository,
+available to all users of the system. When an end user is using a site
+that already has an adapter, they get a semantic customization
+experience that avoids low-level details.
 
@@
 low-level details of a user interface to semantic constructs (e.g.,
-using CSS selectors to find certain page elements), and 2) the actual
-logic of the specific customization. Even though the mapping logic is
-often more generic than the specific customization, the intertwining of
-these two responsibilities in a single script makes it very difficult to
-share the mapping logic across scripts.
+using CSS selectors to find certain page elements), and 2) handling the
+actual logic of the specific customization. Even though the mapping
+logic is often more generic than the specific customization, the
+intertwining of these two responsibilities in a single script makes it
+very difficult to share the mapping logic across scripts.
 
-With Wildcard we propose a decoupling of these two layers: a
-community-maintained mapping layer, shared across many specific
-customizations by individual users. This architecture has been
-successfully used by projects like Gmail.js, an open source project that
-creates a convenient API for browser extensions to interface with the
-Gmail web email client.
+With Wildcard we propose a decoupling of these two layers: a repository
+of shared wrappers maintained by programmers, and a separate repository
+of specific customizations built on top of these wrappers. This general
+architecture has been successfully demonstrated by projects like
+Gmail.js, an open source project that creates a convenient API for
+browser extensions to interface with the Gmail web email client.
+
+(\emph{todo: mention automated wrapper induction here or somewhere
+else?})
 
@@
 \item
+  web scraping / wrapper induction
+\item
   instrumental interaction, Scotty \citep{eagan2011}, Webstrates

### 2020-05-21 paper.tex
more edits from dnj

@@
 \documentclass[sigplan,10pt,anonymous,review]{acmart}
+
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\newenvironment{Shaded}{}{}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
 
@@
 
-It is reasonable to choose imperative programming as the model for
-building customizations. Turing-Complete programming provides a high
-ceiling for possible customizations, and the a sequence of commands is a
-natural fit for automations which simulate a series of steps taken by
-the user.
+Imperative programming is a reasonable choice as the model for building
+customizations. Turing-complete programming provides a high ceiling for
+possible customizations, and a sequence of commands is a natural fit for
+automations which simulate a series of steps taken by the user.
 
@@
 
-We have used Wildcard to build customizations for many real websites. So
-far, most usage has come from members of the project team, although we
-have had a few external users try the system. Here we offer our
-reflections on using the system, focused on two key questions:
+We used Wildcard to customize 11 websites, including transactional sites
+like Amazon and Uber Eats, media consumption sites like Hacker News and
+Youtube, and a medical results entry platform managed by a member of our
+research group. (\emph{todo: summarize all of these in a table}) So far,
+most usage has come from members of the project team, although we have
+had a few external users try the system.
+
+Here we offer our reflections on using the system, focused on two key
+questions:
 
@@
 
-\emph{todo}
-
 \hypertarget{viability-of-dom-scraping}{%
@@
 
-In order for our idea of third-party customization through Wildcard to
-succeed, it is important that creating usable adapters for existing
-websites takes minimal effort.
+In order for third-party customization through Wildcard to succeed, it
+is important that creating usable adapters for existing websites takes
+minimal effort.
 
-To test the viability of writing DOM scraping adapters, we built
-adapters for 10 websites, including transactional sites like Amazon and
-Uber Eats, media consumption sites like Hacker News and Youtube, and a
-medical results entry platform managed by a member of our research
-group. The adapters ranged from 36 to 117 lines of code, averaging 68
-lines. In addition, an external developer unaffiliated with the project
-contributed one adapter, designed to sort the Github page listing a
-user's repositories, and described the experience as ``very
-straightforward.''
+Most of our adapters were created by members of our team. However, an
+external developer unaffiliated with the project contributed one
+adapter, designed to sort the Github page listing a user's repositories.
+They described the experience as ``very straightforward.''
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\StringTok{\textquotesingle{}use strict\textquotesingle{}}\OperatorTok{;}
+\ImportTok{import} \OperatorTok{\{}\NormalTok{ urlMatches}\OperatorTok{,}\NormalTok{ extractNumber }\OperatorTok{\}} \ImportTok{from} \StringTok{"../utils"}\OperatorTok{;}
+\ImportTok{import} \OperatorTok{\{}\NormalTok{ createDomScrapingAdapter }\OperatorTok{\}} \ImportTok{from} \StringTok{"./domScrapingBase"}
+
+\KeywordTok{const}\NormalTok{ GithubAdapter }\OperatorTok{=} \AttributeTok{createDomScrapingAdapter}\NormalTok{(}\OperatorTok{\{}
+  \DataTypeTok{name}\OperatorTok{:} \StringTok{"Github"}\OperatorTok{,}
+  \DataTypeTok{enabled}\OperatorTok{:}\NormalTok{ () }\KeywordTok{=>} \AttributeTok{urlMatches}\NormalTok{(}\SpecialStringTok{/github.com.}\SpecialCharTok{*}\SpecialStringTok{tab=repositories/}\NormalTok{)}\OperatorTok{,}
+  \DataTypeTok{attributes}\OperatorTok{:}\NormalTok{ [}
+    \OperatorTok{\{} \DataTypeTok{name}\OperatorTok{:} \StringTok{"name"}   \OperatorTok{,} \DataTypeTok{type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{\},}
+    \OperatorTok{\{} \DataTypeTok{name}\OperatorTok{:} \StringTok{"stars"}  \OperatorTok{,} \DataTypeTok{type}\OperatorTok{:} \StringTok{"numeric"}\OperatorTok{\},}
+    \OperatorTok{\{} \DataTypeTok{name}\OperatorTok{:} \StringTok{"forks"}  \OperatorTok{,} \DataTypeTok{type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{\},}
+    \OperatorTok{\{} \DataTypeTok{name}\OperatorTok{:} \StringTok{"updated"}\OperatorTok{,} \DataTypeTok{type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{\},}
+    \CommentTok{// }\AlertTok{TODO}\CommentTok{ datetime type would be nice? not everything has ISO formatted strings}
+\NormalTok{  ]}\OperatorTok{,}
+  \AttributeTok{scrapePage}\NormalTok{() }\OperatorTok{\{}
+    \ControlFlowTok{return} \VariableTok{Array}\NormalTok{.}\AttributeTok{from}\NormalTok{(}\VariableTok{document}\NormalTok{.}\AttributeTok{querySelectorAll}\NormalTok{(}\StringTok{"li.source"}\NormalTok{)).}\AttributeTok{map}\NormalTok{(el }\KeywordTok{=>} \OperatorTok{\{}
+      \KeywordTok{let}\NormalTok{ name\_el }\OperatorTok{=} \VariableTok{el}\NormalTok{.}\AttributeTok{querySelector}\NormalTok{(}\StringTok{\textquotesingle{}a[itemprop="name codeRepository"]\textquotesingle{}}\NormalTok{)}
+      \KeywordTok{let}\NormalTok{ name }\OperatorTok{=} \VariableTok{name\_el}\NormalTok{.}\VariableTok{textContent}\NormalTok{.}\AttributeTok{trim}\NormalTok{()}
+
+      \KeywordTok{let}\NormalTok{ stars\_el }\OperatorTok{=} \VariableTok{el}\NormalTok{.}\AttributeTok{querySelector}\NormalTok{(}\StringTok{\textquotesingle{}*[href*="/stargazers"\textquotesingle{}}\NormalTok{)}
+      \KeywordTok{let}\NormalTok{ stars }\OperatorTok{=} \AttributeTok{extractNumber}\NormalTok{(stars\_el}\OperatorTok{,} \DecValTok{0}\NormalTok{)}
+
+      \KeywordTok{let}\NormalTok{ forks\_el }\OperatorTok{=} \VariableTok{el}\NormalTok{.}\AttributeTok{querySelector}\NormalTok{(}\StringTok{\textquotesingle{}*[href*="/network/members"]\textquotesingle{}}\NormalTok{)}
+      \KeywordTok{let}\NormalTok{ forks }\OperatorTok{=} \AttributeTok{extractNumber}\NormalTok{(forks\_el}\OperatorTok{,} \DecValTok{0}\NormalTok{)}
+
+      \KeywordTok{let}\NormalTok{ lang\_el }\OperatorTok{=} \VariableTok{el}\NormalTok{.}\AttributeTok{querySelector}\NormalTok{(}\StringTok{\textquotesingle{}*[itemprop="programmingLanguage"]\textquotesingle{}}\NormalTok{)}
+      \CommentTok{// some repos don\textquotesingle{}t have language set}
+      \KeywordTok{let}\NormalTok{ lang }\OperatorTok{=}\NormalTok{ lang\_el }\OperatorTok{==} \KeywordTok{null} \OperatorTok{?} \KeywordTok{null}\NormalTok{ : }\VariableTok{lang\_el}\NormalTok{.}\VariableTok{textContent}\NormalTok{.}\AttributeTok{trim}\NormalTok{()}
+
+      \KeywordTok{let}\NormalTok{ updated\_el }\OperatorTok{=} \VariableTok{el}\NormalTok{.}\AttributeTok{querySelector}\NormalTok{(}\StringTok{\textquotesingle{}relative{-}time\textquotesingle{}}\NormalTok{)}
+      \KeywordTok{let}\NormalTok{ updated }\OperatorTok{=} \VariableTok{updated\_el}\NormalTok{.}\AttributeTok{getAttribute}\NormalTok{(}\StringTok{\textquotesingle{}datetime\textquotesingle{}}\NormalTok{)}
+
+      \ControlFlowTok{return} \OperatorTok{\{}
+          \DataTypeTok{id}\OperatorTok{:}\NormalTok{ name}\OperatorTok{,}
+          \DataTypeTok{rowElements}\OperatorTok{:}\NormalTok{ [el]}\OperatorTok{,}
+          \DataTypeTok{dataValues}\OperatorTok{:} \OperatorTok{\{}
+              \DataTypeTok{name}\OperatorTok{:}\NormalTok{ name}\OperatorTok{,}
+              \DataTypeTok{stars}\OperatorTok{:}\NormalTok{ stars}\OperatorTok{,}
+              \DataTypeTok{forks}\OperatorTok{:}\NormalTok{ forks}\OperatorTok{,}
+              \DataTypeTok{updated}\OperatorTok{:}\NormalTok{ updated}\OperatorTok{,}
+          \OperatorTok{\},}
+      \OperatorTok{\}}
+    \OperatorTok{\}}\NormalTok{)}
+  \OperatorTok{\},}
+  \DataTypeTok{onRowSelected}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \OperatorTok{\{}
+      \VariableTok{row}\NormalTok{.}\VariableTok{rowElements}\NormalTok{.}\AttributeTok{forEach}\NormalTok{(el }\KeywordTok{=>} \OperatorTok{\{}
+          \ControlFlowTok{if}\NormalTok{ (}\VariableTok{el}\NormalTok{.}\AttributeTok{style}\NormalTok{) }\OperatorTok{\{}
+              \VariableTok{el}\NormalTok{.}\AttributeTok{style}\NormalTok{[}\StringTok{"background{-}color"}\NormalTok{] }\OperatorTok{=} \StringTok{"\#def3ff"}
+          \OperatorTok{\}}
+      \OperatorTok{\}}\NormalTok{)}\OperatorTok{;}
+      \VariableTok{row}\NormalTok{.}\AttributeTok{rowElements}\NormalTok{[}\DecValTok{0}\NormalTok{].}\AttributeTok{scrollIntoView}\NormalTok{(}\OperatorTok{\{} \DataTypeTok{behavior}\OperatorTok{:} \StringTok{"smooth"}\OperatorTok{,} \DataTypeTok{block}\OperatorTok{:} \StringTok{"center"} \OperatorTok{\}}\NormalTok{)}
+  \OperatorTok{\},}
+  \DataTypeTok{onRowUnselected}\OperatorTok{:}\NormalTok{ (row) }\KeywordTok{=>} \OperatorTok{\{}
+      \VariableTok{row}\NormalTok{.}\VariableTok{rowElements}\NormalTok{.}\AttributeTok{forEach}\NormalTok{(el }\KeywordTok{=>} \OperatorTok{\{}
+          \ControlFlowTok{if}\NormalTok{(}\VariableTok{el}\NormalTok{.}\AttributeTok{style}\NormalTok{) }\OperatorTok{\{}
+              \VariableTok{el}\NormalTok{.}\AttributeTok{style}\NormalTok{[}\StringTok{"background{-}color"}\NormalTok{] }\OperatorTok{=} \VerbatimStringTok{\textasciigrave{}\textasciigrave{}}
+          \OperatorTok{\}}
+      \OperatorTok{\}}\NormalTok{)}
+  \OperatorTok{\},}
+\OperatorTok{\}}\NormalTok{)}
+
+\ImportTok{export} \ImportTok{default}\NormalTok{ GithubAdapter}\OperatorTok{;}
+\end{Highlighting}
+\end{Shaded}
+
+The adapters for our test sites ranged from 36 to 117 lines of code,
+averaging 68 lines.
 

### 2020-05-21 paper.tex
related work

--- 2020-05-21 paper.tex
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Customizing Software by Direct Manipulation of Structured Data}
+\title{Customizing Software by Direct Manipulation of Tabular Data}
 
@@
 
+\begin{table*}[]
+\begin{tabular}{llrl}
+\hline
+\textbf{Website} & \textbf{Description} & \textbf{LOC} & \textbf{Example customizations}                                                              \\ \hline
+Airbnb           & Travel               & 73                                       & Add Walkability Scores to listings. Sort listings by price.                           \\
+Amazon           & Online shopping      & 99                                       & Sort used book sellers by total price, including delivery fees                                \\
+Blogger          & Blogging             & 36                                       & Use alternate text editor to edit blog posts                                                 \\
+Expedia          & Travel               & 41                                       & Use alternate datepicker to enter travel dates                                               \\
+Flux             & Data portal          & 67                                       & Use Wildcard as a faster table editor for editing lab results                                \\
+Github           & Code repository      & 62                                       & Sort a user's code repositories by stars to find popular work                                \\
+Hacker News      & News                 & 69                                       & Add read times to links. Filter out links that have been read. \\
+Instacart        & Grocery delivery     & 48                                       & Sort groceries by price and category. Take notes on items.                                   \\
+Uber Eats        & Food delivery        & 117                                      & Sort/filter restaurants by estimated delivery ETA and price.                                 \\
+Weather.com  & Weather              & 51                                       & Sort/filter hourly weather to find nice times of day.                                        \\
+Youtube          & Videos               & 80                                       & Sort/filter videos by length, to find short videos to watch.                                 \\ \hline
+\end{tabular}
+\caption{A list of table-driven customizations that we have implemented using Wildcard.}
+\label{tab:websites}
+\end{table*}
+
+\hypertarget{sec:evaluation}{%
+\section{Evaluation}\label{sec:evaluation}}
+
+To evaluate table-driven customization in practice, we built the
+Wildcard browser extension, which implements table-driven customization
+in the context of existing websites. It is built in Typescript, and
+works across three major browsers: Chrome, Firefox, and Edge.
+
+We have used Wildcard to customize 11 websites, including transactional
+sites like Amazon and Uber Eats, and media consumption sites like Hacker
+News and Youtube. Table 1 summarizes all these customizations, including
+the number of lines of code in the adapter configuration for each site.
+
+So far, most usage of Wildcard has come from members of the project
+team. Here we offer our reflections on using the system, focused on two
+key questions:
+
+\begin{itemize}
+\tightlist
+\item
+  How broad is the range of possible customizations in this paradigm?
+\item
+  How feasible is it to build DOM scraping adapters for real websites,
+  in practice?
+\end{itemize}
+
+\hypertarget{range-of-customizations}{%
+\subsection{Range of customizations}\label{range-of-customizations}}
+
+We have found that table-driven customization can serve a broad range of
+useful purposes. Here we expand on some examples that illuminate various
+aspects of using the system in practice.
+
+\hypertarget{sortingfiltering}{%
+\subsubsection{Sorting/filtering}\label{sortingfiltering}}
+
+It might seem that most websites already have adequate sorting and
+filtering functionality, but we have found it surprisingly helpful to
+add new sorting/filtering functionality to websites using Wildcard.
+
+Sometimes, websites have opaque algorithms which they use to rank
+results, which presumably maximize profit but don't offer much control
+to the user. For example, Airbnb doesn't allow users to sort listings by
+price, and Youtube doesn't allow users to sort the homepage by watch
+time. Wildcard enables users to take back some control over this
+process.
+
+In other cases we have found that a lack of sorting options seems less
+intentional, and simply a missing feature. For example, the Instacart
+grocery delivery service doesn't allow users to sort their grocery cart
+by price or category. Our guess is that they simply haven't gotten
+around to implementing the feature yet.
+
+In the current implementation of Wildcard, users can only sort and
+filter entries which are shown on the current page, which means that
+users are not entirely liberated from the suggestions of the opaque
+algorithm. We could work around this restriction in the future by
+fetching content across multiple pages, but we've also found that
+sorting/filtering a single page of a paginated list is often an
+acceptable outcome, and even a preferable one. For example, it's more
+useful to sort 30 recommended Youtube videos than to try to sort all
+videos on Youtube; similarly it's convenient to view the Hacker News
+front page sorted by read time, rather than all Hacker News articles.
+
+\hypertarget{annotating}{%
+\subsubsection{Annotating}\label{annotating}}
+
+Many other web annotation systems focus on annotating text or arbitrary
+webpage content, but Wildcard only allows for annotating structured
+objects extracted by an adapter, which results in quite a different set
+of use cases.
+
+Annotating with Wildcard has proven most useful when taking notes on a
+list of possible options: e.g., evaluating possible Airbnb locations to
+rent. We have also used it with Instacart's online grocery cart, for
+jotting down notes like ``should we get more milk?''
+
+\hypertarget{formulas}{%
+\subsubsection{Formulas}\label{formulas}}
+
+Formulas are the most powerful part of the Wildcard system. So far, we
+have only built a small number of functions into the language, but by
+adding more functions, the fundamental computing model could support a
+very broad range of useful computations, as shown by the success of
+spreadsheets.
+
+We have found that formulas are useful for fetching data from Web APIs.
+We've used them to augment Airbnb listings with walkability scores, and
+augment Hacker News articles with estimated read times. One challenge of
+the current language design is that supporting each web API requires
+adding a new function to the language using code, because web APIs
+typically return complex JSON data structures which don't make sense to
+display in a single table cell. Adding functions for handling JSON data
+to the formula language might help resolve this.
+
+We have also found instances where simple data manipulation is useful:
+adding together different subtotals into a total price, or manipulating
+the results of a formula expression with arithmetic or concatenating
+string labels, as shown in the example in Section~\ref{sec:examples}.
+(\emph{todo: this makes sense iff we switch the examples sectino to HN})
+
+\hypertarget{cell-editors}{%
+\subsubsection{Cell editors}\label{cell-editors}}
+
+(\emph{note: written with the assumption that Expedia gets cut from the
+Examples section})
+
+``Cell editors'' are UI widgets that expose a custom editing UI for a
+single cell of the table view. A programmer building a cell editor only
+needs to integrate it with the table viewer; actually propagating values
+into the website UI is handled by the site-specific DOM adapter. As a
+result, creating a new cell editor can be as easy as just creating glue
+code between the table editor and an existing UI widget library.
+
+One benefit of cell editors is that a user can use their personal
+information in a web UI without uploading it to the website. To explore
+this idea, we created a cell editor based on the FullCalendar Javascript
+plugin, which can load data from a Google Calendar. This makes it
+convenient to enter dates into a website based on the user's personal
+calendar information.
+
+Another benefit of cell editors is that a user can choose their
+preferred widget for editing some type of information. We built a cell
+editor based on the CKEditor rich text editor. As an example of using
+this editor, we integrated it with the Blogger website and used it to
+edit blog posts, in place of the Blogger website's built-in editor.
+
+\hypertarget{limitations}{%
+\subsubsection{Limitations}\label{limitations}}
+
+There are many useful customizations which are not possible in the
+table-driven customization paradigm. Wildcard can only make
+customizations that use the available data exposed in the table. UI
+modifications are limited in scope; for example, deleting arbitrary
+buttons isn't possible. At one point, we wanted to build an automation
+to repeatedly load a grocery delivery website to check for open delivery
+slots, but it wasn't clear how to achieve this. Some of these
+limitations are specific to the current implementation of the Wildcard
+extension, but some are more general to the entire idea of table-driven
+customization.
+
+We consider this an acceptable outcome---our goal is simply to make many
+useful customizations available to end users with a low threshold, not
+to span all possible customizations. Section~\ref{sec:dm} discusses this
+point further.
+
+One benefit of Wildcard is \emph{predictability}: once we built a site
+adapter for a website, it was generally obvious to us what types of
+customizations were and weren't possible, and the UI guided us towards
+building customizations that matched the system's model. For example,
+it's clear that it's not possible to customize a website using data not
+available in the table.
+
+\hypertarget{viability-of-dom-scraping}{%
+\subsection{Viability of DOM scraping}\label{viability-of-dom-scraping}}
+
+Our second evaluation area is less about the conceptual approach of
+table-driven customization, and more about our specific implementation
+of customizing existing web applications. In order for third-party
+customization through Wildcard to succeed, it is important that creating
+usable adapters for existing websites takes minimal effort.
+
+Most of our DOM scraping adapters were created by members of our team.
+However, an external developer unaffiliated with the project contributed
+one adapter, designed to sort the Github page listing a user's
+repositories. They described the experience as ``very straightforward.''
+
+The adapters for our test sites ranged from 36 to 117 lines of code,
+averaging 68 lines; Table 1 shows the full lines of code for each
+adapter.
+
+Some of the challenges of writing a DOM scraping adapter are the same
+ones as with writing normal web scraping code, but the more interactive
+nature of Wildcard introduces additional challenges. rve changes, but it
+may prove challenging to observe changes on some sites only through the
+DOM. One challenge is triggering updates to the spreadsheet data in
+response to UI changes that happen after initial page load. Site
+adapters are responsible for recognizing these changes by observing the
+DOM. So far, we have been able to use event listeners and the
+MutationObserver API to successfully obse
+
+Another challenge is persisting updates to the DOM---some websites use
+virtual DOM frameworks that can occasionally overwrite changes made by
+Wildcard.
+
+So far, we've managed to work around these issues for all the websites
+we've tried, but we don't claim that any website can be customized
+through DOM scraping. As web frontend code gets increasingly complex
+(and starts to move beyond the DOM to other technologies like Shadow DOM
+or even WebGL for rendering), it may become increasingly difficult to
+customize websites from the outside without first-party support.
+
 \hypertarget{sec:themes}{%
@@
 
-\hypertarget{customization-by-direct-manipulation}{%
-\subsection{Customization by direct
-manipulation}\label{customization-by-direct-manipulation}}
+\hypertarget{sec:dm}{%
+\subsection{Customization by direct manipulation}\label{sec:dm}}
 
@@
 iterates through rows in a page adding some additional information could
-be reproduced using a formula in Wildcard. (\emph{todo: mention the fact
-that the data abstraction step limits possibilities})
+be reproduced using a formula in Wildcard.
 
@@
 
-\hypertarget{sec:evaluation}{%
-\section{Evaluation}\label{sec:evaluation}}
-
-We have used Wildcard to build customizations for many real websites. So
-far, most usage has come from members of the project team, although we
-have had a few external users try the system. Here we offer our
-reflections on using the system, focused on two key questions:
-
-\begin{itemize}
-\tightlist
-\item
-  What is the range of useful customizations possible in this paradigm?
-\item
-  How feasible is it to build DOM scraping adapters for real websites,
-  in practice?
-\end{itemize}
-
-\hypertarget{range-of-customizations}{%
-\subsection{Range of customizations}\label{range-of-customizations}}
-
-\emph{todo}
-
-\hypertarget{viability-of-dom-scraping}{%
-\subsection{Viability of DOM scraping}\label{viability-of-dom-scraping}}
-
-In order for our idea of third-party customization through Wildcard to
-succeed, it is important that creating usable adapters for existing
-websites takes minimal effort.
-
-To test the viability of writing DOM scraping adapters, we built
-adapters for 10 websites, including transactional sites like Amazon and
-Uber Eats, media consumption sites like Hacker News and Youtube, and a
-medical results entry platform managed by a member of our research
-group. The adapters ranged from 36 to 117 lines of code, averaging 68
-lines. In addition, an external developer unaffiliated with the project
-contributed one adapter, designed to sort the Github page listing a
-user's repositories, and described the experience as ``very
-straightforward.''
-
-Some of the challenges of writing a DOM scraping adapter are the same
-ones involved with writing normal web scraping code, but the more
-interactive nature of Wildcard introduces additional challenges. One
-challenge is triggering updates to the spreadsheet data in response to
-UI changes that happen after initial page load. Site adapters are
-responsible for recognizing these changes by observing the DOM. So far,
-we have been able to use event listeners and the MutationObserver API to
-successfully observe changes, but it may prove challenging to observe
-changes on some sites only through the DOM. Another challenge is
-persisting updates to the DOM---some websites use virtual DOM frameworks
-that can occasionally overwrite changes made by Wildcard. Despite these
-challenges, we've managed to create an implementation that works around
-these issues in practice, and results in a usable customization
-experience for all the websites we've tried so far.
-
 \hypertarget{sec:related-work}{%
@@
 
-This paper extends a workshop paper by Litt and Jackson \citep{litt2020}
-which presented an earlier prototype version of the Wildcard system. We
-build on their work in this paper by rearchitecting the internal
-implementation of the system around the table adapter abstraction,
-evaluating it on many more websites, and by more fully characterizing
-the behavior and design of the system.
+This paper extends work reported in a workshop paper by Litt and Jackson
+\citep{litt2020} which presented an earlier prototype version of the
+Wildcard system. We build on their work in this paper by rearchitecting
+the internal implementation of the system around the table adapter
+abstraction, evaluating it on many more websites, and by more fully
+characterizing the behavior and design of the system.
 

### 2020-05-21 paper.tex
related work

@@
 a simplified version of programming. Some scripting languages
-(AppleScript, Chickenfoot) have a friendly syntax that resembles natural
-language. Visual programming tools (Mac Automator, Zapier) eliminate
-text syntax entirely. Macro recorders (Applescript, Helena, WebVCR)
+\citep{bolin2005, cook2007} have a friendly syntax that resembles
+natural language. Some visual customization tools eliminate text syntax
+entirely. Macro recorders \citep{cook2007, chasins2018, anupam2000}
 remove some of the initial programming burden by letting a user start
@@
 
-Some customization environments (Mac Automator, Zapier) forego text
-syntax and enable the user to connect programs and construct automations
-by dragging and dropping commands. These environments still do not
-constitute direct manipulation, though: the objects being manipulated
-are in the domain of programming, not the domain of the task at hand.
+Some customization environments (Automator for Mac, Shortcuts for iOS,
+Zapier for web APIs) forego text syntax and enable the user to connect
+programs and construct automations by dragging and dropping commands.
+These environments still do not constitute direct manipulation, though:
+the objects being manipulated are in the domain of programming, not the
+domain of the task at hand.
 
@@
 This paper extends work reported in a workshop paper by Litt and Jackson
-\citep{litt2020} which presented an earlier prototype version of the
-Wildcard system. We build on their work in this paper by rearchitecting
-the internal implementation of the system around the table adapter
-abstraction, evaluating it on many more websites, and by more fully
-characterizing the behavior and design of the system.
+\citep{litt2020} which presented an early prototype version of Wildcard.
+We have substantially extended their work in this paper by creating the
+table adapter abstraction and reimplementing the system around that
+abstraction, evaluating the system more fully on many more websites, and
+by characterizing the design of the system in more detail.
+
+Table-driven customization relates to two broad areas of related work.
+Our problem statement is related to software customization tools, and
+our solution approach is related to spreadsheets and other direct
+manipulation interfaces.
 
@@
 
-\begin{itemize}
-\item
-  Buttons \citep{maclean1990}: the original DM customization tool
-\item
-  ScrAPIr: direct manipulation for APIs!
-\item
-  customization tools
-
-  \begin{itemize}
-  \item
-    browser extensions
-  \item
-    Thresher
-  \item
-    Sifter
-  \item
-    customization research tools: Chickenfoot, Coscripter
-  \item
-    Wildcard, like Chickenfoot, wants to hide HTML from users. But we
-    show a structured data view, whereas Chickenfoot shows nothing
-  \item
-    visual customization tools:
-
-    \begin{itemize}
-    \tightlist
-    \item
-      Mac Automator
-    \item
-      Zapier
-    \item
-      IFTTT
-    \item
-      These are weakly direct manipulation. But usually the thing you
-      directly manipulate in these UIs is the \emph{script}: not the
-      actual \emph{data} you want to operate on. VERY DIFFERENT.
-    \end{itemize}
-  \item
-    desktop customization: Applescript, VBA, COM
-  \end{itemize}
-\end{itemize}
-
-\hypertarget{visual-database-query-tools}{%
-\subsection{Visual database query
-tools}\label{visual-database-query-tools}}
-
-\begin{itemize}
-\tightlist
-\item
-  database GUIs:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Sieuferd \citep{bakke2016}
-  \item
-    Airtable \citep{2020a}
-  \item
-    other similar tools?
-  \item
-    Liu \& Jagadish: rules for a spreadsheet algebra for database
-    queries \citep{liu2009}
-  \item
-    Shneiderman paper
-  \item
-    spreadsheets
-  \end{itemize}
-\end{itemize}
-
-\hypertarget{other}{%
-\subsection{Other}\label{other}}
-
-\begin{itemize}
-\tightlist
-\item
-  web scraping / wrapper induction
-\item
-  instrumental interaction, Scotty \citep{eagan2011}, Webstrates
-\item
-  browser dev tools
-\item
-  Self, Smalltalk
-\item
-  multiple representations
-\item
-  Semantic Web
-\end{itemize}
+Table-driven customization is most closely related to other tools that
+aim to empower end users to customize software without doing any
+programming.
+
+This lineage goes back at least to the Buttons system by MacLean et al
+\citep{maclean1990}, where Xerox Lisp users could share buttons that
+performed various ``tailoring'' actions on the system. The authors
+proposed the ``gentle slope'' idea which has greatly influenced our
+approach to table-driven customization (as discussed in
+Section~\ref{sec:dm}). The authors also point out the importance of a
+``tailoring culture'' where people with different skillsets collaborate
+to produce useful customizations; in their system, Lisp programmers
+create buttons that others can use, modify, and rearrange. This division
+of labor corresponds to our idea of semantic wrappers, where end user
+customization is supported by programmer-created building blocks.
+
+Some recent web customization tools aim to enable end users to modify
+web interfaces without programming.
+
+Sifter \citep{huynh2006} enables end users to sort and filter lists of
+data based on web scraping, much like Wildcard's sorting features. The
+main difference between the systems is that table-driven customization
+has many other use cases besides sorting and filtering. Also, Sifter
+involves end users in a semi-automated data extraction process, rather
+than having programmers create wrappers. This provides coverage of more
+websites, but at the expense of complicating the end user experience. We
+might explore such techniques in Wildcard in the future, but we think
+that it's valuable for end users to have a customization experience free
+of web scraping details on supported websites. Finally Sifter implements
+scraping across multiple pages, a valuable feature for sorting and
+filtering which isn't present in Wildcard.
+
+Thresher \citep{hogue2005} enables end users to create wrappers which
+map website content to Semantic Web schemas, making customizations
+available using the schema information. We share the general idea of
+wrappers with Thresher, but we have chosen to map to a single generic
+data type, rather than more specific schemas, which increases the range
+of websites and data supported by the system.
+
+There are many software customization tools that offer simplified
+versions of programming for end users. Chickenfoot \citep{bolin2005} and
+Coscripter \citep{leshed2008} offer friendly syntax for writing web
+automation scripts; Applescript \citep{cook2007} attempts a similar goal
+for desktop customization. There are visual programming environments for
+customization which don't involve writing any text: as a few examples,
+\href{https://support.apple.com/guide/automator/welcome/mac}{Automator}
+for Mac and
+\href{https://apps.apple.com/us/app/shortcuts/id915249334}{Shortcuts}
+for iOS are modern options for customizing Apple products, and
+\href{https://zapier.com/}{Zapier} enables users to connect different
+web applications together visually. As mentioned previously, these tools
+all require writing imperative programs in some form; we take a
+fundamentally different approach in table-driven customization.
+
+\hypertarget{direct-manipulation-programming-interfaces.}{%
+\subsection{Direct manipulation programming
+interfaces.}\label{direct-manipulation-programming-interfaces.}}
+
+Another relevant area of literature is direct manipulation interfaces
+for interacting with structured data. We take much inspiration from
+these tools in our work, but apply them in a different domain:
+customizing existing software applications, rather than interacting with
+databases or constructing software from scratch.
+
+The most closely related work is in systems that offer spreadsheet-like
+querying of relational data, as proposed by Liu and Jagadish
+\citep{liu2009}. SIEUFERD by Bakke and Karger \citep{bakke2016} is an
+example; their paper presents a survey of various other similar tools.
+Our work is particularly influenced by the authors' observation that
+direct manipulation requires that the user manipulate the \emph{results}
+of a database query rather than the query itself, and that the user must
+see intermediate results at every step of constructing a computation.
+SIEUFERD's interface supports a far wider range of possible queries than
+Wildcard, but the basic ideas of the user interface are similar.
+
+\href{https://airtable.com/}{Airtable} is another example of a modern
+commercial product that offers spreadsheet-like interaction with a
+relational database.
+
+Our work is also inspired by the many projects which have explored using
+spreadsheets as a foundation for building software applications,
+including Object Spreadsheets \citep{mccutchen2016}, Quilt
+\citep{benson2014}, Gneiss \citep{chang2014}, Marmite \citep{wong2007},
+\href{https://www.glideapps.com/}{Glide}. These projects share the idea
+that a spreadsheet is a convenient interface for editing the data
+underlying a GUI application. We share that idea, but apply it to
+software customization, rather than building software from scratch.
+
+Another system in this space is ScrAPIr, by Alrashed et
+al.~\citep{alrashed2020}, which enables end users to access backend web
+APIs without programming. ScrAPIr shares a high level goal of end user
+empowerment, and also shares the idea of wrappers, by creating a shared
+library of wrappers around existing web APIs. However, unlike Wildcard,
+ScrAPIr targets explicit APIs exposed by developers, and also doesn't
+aim to extend the frontend interfaces of web applications.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+\emph{todo}
 

### 2020-05-22 paper.tex
add more dnj edits
new example section

@@
 Web applications. In Section~\ref{sec:examples}, we introduce the key
-ideas of table-driven customization by presenting several examples of
-real customizations implemented in Wildcard.
+ideas of table-driven customization through an example scenario.
 
@@
 \hypertarget{sec:examples}{%
-\section{Examples}\label{sec:examples}}
-
-\emph{todo: maybe redo this as a separate example to avoid self
-plagiarism?}
+\section{Example Scenario}\label{sec:examples}}
 
 To concretely illustrate the end user experience of table-driven
-customization, here are several real examples of using the Wildcard
-browser extension to customize websites.
+customization, here is a real example of using the Wildcard browser
+extension to customize a website.
+
+\begin{itemize}
+\tightlist
+\item
+  Hacker News
+\item
+  open the table
+\item
+  sort by comments
+\item
+  take notes on links
+\item
+  filter out visited sites with a formula
+\item
+  add read times with a formula
+\end{itemize}
 
 \begin{figure*}
-\hypertarget{fig:airbnb-demo}{%
+\hypertarget{fig:hacker-news}{%
 \centering
-\includegraphics{media/airbnb-demo-300dpi.png}
-\caption{Using Wildcard to augment the Airbnb search page for booking
-accommodations}\label{fig:airbnb-demo}
+\includegraphics[width=\textwidth]{media/hacker-news.png}
+\caption{Customizing Hacker News by interacting with a table view}\label{fig:hacker-news}
 }
 \end{figure*}
-
-\hypertarget{augmenting-search-results}{%
-\subsection{Augmenting search results}\label{augmenting-search-results}}
-
-In 2012, the travel site Airbnb removed the ability to sort
-accommodation searches by price. Users could still filter by price
-range, but could no longer view the cheapest listings first. Many users
-complained that the change seemed hostile to users. ``It's so
-frustrating! What is the logic behind not having this function?'' said
-one user on the
-\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
-support forum}. Alas, the feature remains missing to this day.
-
-Using Wildcard, the user can fix this omission, while leaving the page's
-design and the rest of its functionality unchanged.{
-Figure~\ref{fig:airbnb-demo} shows an overview of augmenting the Airbnb
-site.} First, the user opens up the Wildcard panel, which shows a table
-corresponding to the search results in the page. As they click around in
-the table, the corresponding row in the page is highlighted to indicate
-the mapping between the views.
-
-Then, the user clicks on the price column header to sort the spreadsheet
-and the Airbnb UI by price{ (Figure~\ref{fig:airbnb-demo}, Note A)}.
-They also filter to listings with a user rating above 4.5 (another
-feature missing in the original Airbnb UI).
-
-After manipulating the data, the user closes the table view and
-continues using the website. Because the application's UI usually has a
-nicer visual design than a spreadsheet, Wildcard does not aim to replace
-it---at any time, the user can use either the UI, the spreadsheet, or
-both together.
-
-Many websites that show lists of data also offer actions on rows in the
-table, like adding an item to a shopping cart. Wildcard has the ability
-to make these ``row actions'' available in the data table through the
-site adapter. In the Airbnb UI, saving multiple listings to a Favorites
-list requires tediously clicking through them one by one. Using Wildcard
-row actions, the user can select multiple rows and favorite all of them
-with a single click{ (Figure~\ref{fig:airbnb-demo}, Note B)}. Similarly,
-the user can open the detailed pages for many listings at once.
-
-Next, the user wants to jot down some notes about each listing. To do
-this, they type some notes into an additional column in each row, and
-the notes appear inside the listings in the original UI{
-(Figure~\ref{fig:airbnb-demo}, Note C)}. The annotations are saved in
-the browser and associated with the backend ID of the listing, so they
-will appear in future browser sessions that display the same listing.
-
-Wildcard also includes a formula language that enables more
-sophisticated customizations. When traveling without a car, it's useful
-to evaluate potential places to stay based on how walkable the
-surroundings are. Using a formula, the user can integrate Airbnb with
-Walkscore, an API that rates the walkability of any location on a 1-100
-scale. When the user calls the \texttt{walkscore} formula with the
-listing's latitude and longitude as arguments, it returns the walk score
-for that location and shows it as the cell value. Because the cell's
-contents are injected into the page, the score also appears in the UI{
-(Figure~\ref{fig:airbnb-demo}, Note D)}.
-
-\hypertarget{snoozing-todos}{%
-\subsection{Snoozing todos}\label{snoozing-todos}}
-
-\begin{figure*}
-\hypertarget{fig:airbnb-demo}{%
-\centering
-\includegraphics{media/todomvc-demo-300dpi.png}
-\caption{Using Wildcard to add a "snooze" feature to the TodoMVC todo list app}\label{fig:todomvc-demo}
-}
-\end{figure*}
-
-In addition to fetching data from other sources, Wildcard formulas can
-also perform computations. In this example, the user would like to
-augment the TodoMVC todo list app with a ``snooze'' feature, which will
-temporarily hide a todo from the list until a certain date.{
-Figure~\ref{fig:todomvc-demo} shows an overview of this customization.}
-
-The user opens the table view, which shows the text and completed status
-of each todo. They start the customization by adding a new column to
-store the snooze date for each todo{ (Figure~\ref{fig:todomvc-demo},
-Note A)}.
-
-The next step is to hide snoozed todos. The user creates a
-\texttt{snoozed?} column, which uses a formula to compute whether a todo
-is snoozed---i.e., whether it has a snooze date in the future{
-(Figure~\ref{fig:todomvc-demo}, Note B)}. Then, they simply filter the
-table to hide the snoozed todos{ (Figure~\ref{fig:todomvc-demo}, Note
-C)}.
-
-Because the built-in \texttt{NOW()} function always returns the current
-datetime, snoozed todos will automatically appear once their snooze date
-arrives.
-
-Because this implementation of snoozing was built on the spreadsheet
-abstraction, it is completely decoupled from this particular todo list
-app. We envision that users could share these types of customizations as
-generic browser extensions, which could be applied to any site supported
-by Wildcard with no additional effort.
-
-\hypertarget{adding-a-custom-datepicker}{%
-\subsection{Adding a custom
-datepicker}\label{adding-a-custom-datepicker}}
-
-\begin{figure*}
-\hypertarget{fig:airbnb-demo}{%
-\centering
-\includegraphics{media/expedia-demo-300dpi.png}
-\caption{Using Wildcard to augment the Expedia page for booking a flight}\label{fig:expedia-demo}
-}
-\end{figure*}
-
-It might seem that Wildcard is only useful on websites that display
-lists of tabular data, but the table metaphor is flexible enough to
-represent many types of data. For example, a flight search form on
-Expedia can be represented as a single row, with a column corresponding
-to each input{ (Figure~\ref{fig:expedia-demo}, Note A)}.
-
-In some of the previous examples, the table cells were read-only
-(because users can't, for example, change the name or price of an Airbnb
-listing). In this case, the cells are writable, which means that changes
-in the table are reflected in the form inputs. This becomes especially
-useful when combined with GUI widgets that can edit the value of a table
-cell.
-
-Filling in dates for a flight search often requires a cumbersome
-workflow: open up a separate calendar app, find the dates for the trip,
-and then carefully copy them into the form. In Wildcard, the user can
-avoid this by using a datepicker widget that shows the user's personal
-calendar events{ (Figure~\ref{fig:expedia-demo}, Note B)}. The user can
-directly click on the correct date, and it gets inserted into both the
-spreadsheet and the original form.n
 
@@
 Wildcard browser extension, which implements table-driven customization
-in the context of existing websites. It is built in Typescript, and
-works across three major browsers: Chrome, Firefox, and Edge.
+in the context of existing websites. It is implemented in Typescript,
+and works across three major browsers: Chrome, Firefox, and Edge.
 
@@
 sites like Amazon and Uber Eats, and media consumption sites like Hacker
-News and Youtube. Table 1 summarizes all these customizations, including
-the number of lines of code in the adapter configuration for each site.
+News and Youtube. Table 1 summarizes these customizations, showing the
+number of lines of code in the adapter configuration for each site.
 
@@
 \item
-  How feasible is it to build DOM scraping adapters for real websites,
-  in practice?
+  How feasible is it to build DOM scraping adapters for real websites?
 \end{itemize}
@@
 We have found that table-driven customization can serve a broad range of
-useful purposes. Here we expand on some examples that illuminate various
-aspects of using the system in practice.
+useful purposes. Here we expand on some archetypal examples that
+illuminate various aspects of using the system in practice.
 
@@
 In other cases we have found that a lack of sorting options seems less
-intentional, and simply a missing feature. For example, the Instacart
-grocery delivery service doesn't allow users to sort their grocery cart
-by price or category. Our guess is that they simply haven't gotten
-around to implementing the feature yet.
+intentional, and simply represents a missing feature. For example, the
+Instacart grocery delivery service doesn't allow users to sort their
+grocery cart by price or category. Our guess is that they simply haven't
+gotten around to implementing the feature yet.
 
 In the current implementation of Wildcard, users can only sort and
-filter entries which are shown on the current page, which means that
+filter entries that are shown on the current page, which means that
 users are not entirely liberated from the suggestions of the opaque
@@
 sorting/filtering a single page of a paginated list is often an
-acceptable outcome, and even a preferable one. For example, it's more
-useful to sort 30 recommended Youtube videos than to try to sort all
-videos on Youtube; similarly it's convenient to view the Hacker News
-front page sorted by read time, rather than all Hacker News articles.
+acceptable outcome (and sometimes even a preferable one). For example,
+it's more useful to sort (by length, say) 30 recommended Youtube videos
+than to try to sort all videos on Youtube, or to sort (by read time,
+say) the articles on the Hacker News front page than to sort all
+non-archived articles.
 
@@
 Many other web annotation systems focus on annotating text or arbitrary
-webpage content, but Wildcard only allows for annotating structured
-objects extracted by an adapter, which results in quite a different set
-of use cases.
+webpage content, but Wildcard limits annotations to structured objects
+extracted by an adapter, resulting in quite a different set of use
+cases.
 
 Annotating with Wildcard has proven most useful when taking notes on a
-list of possible options: e.g., evaluating possible Airbnb locations to
-rent. We have also used it with Instacart's online grocery cart, for
+list of possible options (e.g., evaluating possible Airbnb locations to
+rent). We have also used it with Instacart's online grocery cart, for
 jotting down notes like ``should we get more milk?''
@@
 
-Formulas are the most powerful part of the Wildcard system. So far, we
-have only built a small number of functions into the language, but by
-adding more functions, the fundamental computing model could support a
-very broad range of useful computations, as shown by the success of
-spreadsheets.
-
-We have found that formulas are useful for fetching data from Web APIs.
-We've used them to augment Airbnb listings with walkability scores, and
+Formulas are the most powerful part of the Wildcard system. So far, our
+language supports only a small number of predefined functions. Adding
+more will allow a very broad range of useful computations, as shown by
+the success of spreadsheets.
+
+Formulas are especially useful for fetching data from Web APIs. We've
+used them to augment Airbnb listings with walkability scores, and to
 augment Hacker News articles with estimated read times. One challenge of
-the current language design is that supporting each web API requires
-adding a new function to the language using code, because web APIs
-typically return complex JSON data structures which don't make sense to
-display in a single table cell. Adding functions for handling JSON data
-to the formula language might help resolve this.
+the current language design is that supporting a new web API requires
+writing conventional code to add a new function to the language, because
+web APIs typically return complex JSON data structures that can't be
+easily displayed in a single table cell. Adding functions for handling
+JSON data to the formula language might make it possible to access new
+APIs without requiring an entire new function in the formula system.
 
@@
 ``Cell editors'' are UI widgets that expose a custom editing UI for a
-single cell of the table view. A programmer building a cell editor only
-needs to integrate it with the table viewer; actually propagating values
-into the website UI is handled by the site-specific DOM adapter. As a
-result, creating a new cell editor can be as easy as just creating glue
-code between the table editor and an existing UI widget library.
-
-One benefit of cell editors is that a user can use their personal
-information in a web UI without uploading it to the website. To explore
-this idea, we created a cell editor based on the FullCalendar Javascript
-plugin, which can load data from a Google Calendar. This makes it
-convenient to enter dates into a website based on the user's personal
-calendar information.
+single cell of the table view. A programmer building a cell editor need
+only integrate it with the table viewer; propagating values into the
+website UI is handled by the site-specific DOM adapter. As a result,
+creating a new cell editor can be as easy as just writing some glue code
+between the table editor and an existing UI widget library.
+
+An important benefit of cell editors is that they allow users to
+incorporate their personal information within the web UI but without
+uploading their personal data to the web server. To explore this idea,
+we created a cell editor based on the FullCalendar Javascript plugin,
+which can load data from a Google Calendar. This makes it convenient to
+enter dates into a website based on the user's personal calendar
+information.
 
@@
 preferred widget for editing some type of information. We built a cell
-editor based on the CKEditor rich text editor. As an example of using
-this editor, we integrated it with the Blogger website and used it to
-edit blog posts, in place of the Blogger website's built-in editor.
+editor based on the CKEditor rich text editor, and demonstrated its use
+by integrating it with Google's Blogger website and using it to edit
+blog posts, in place of Blogger's built-in editor.
 
@@
 
-There are many useful customizations which are not possible in the
-table-driven customization paradigm. Wildcard can only make
+There are many useful customizations that are not currently possible in
+the table-driven customization paradigm. Wildcard can only make
 customizations that use the available data exposed in the table. UI
-modifications are limited in scope; for example, deleting arbitrary
-buttons isn't possible. At one point, we wanted to build an automation
-to repeatedly load a grocery delivery website to check for open delivery
+modifications are limited in scope; deleting arbitrary buttons isn't
+possible, for example. At one point, we wanted to build an automation to
+repeatedly load a grocery delivery website to check for open delivery
 slots, but it wasn't clear how to achieve this. Some of these
 limitations are specific to the current implementation of the Wildcard
-extension, but some are more general to the entire idea of table-driven
-customization.
-
-We consider this an acceptable outcome---our goal is simply to make many
-useful customizations available to end users with a low threshold, not
-to span all possible customizations. Section~\ref{sec:dm} discusses this
-point further.
-
-One benefit of Wildcard is \emph{predictability}: once we built a site
-adapter for a website, it was generally obvious to us what types of
+extension, but some are more fundamental to the entire idea of
+table-driven customization.
+
+We consider this an acceptable outcome; our goal is to make many useful
+customizations available to end users with a low threshold, and not to
+span all possible customizations. Section 5.1 discusses this point
+further.
+
+One benefit of Wildcard is predictability: once we built a site adapter
+for a website, it was generally obvious to us what types of
 customizations were and weren't possible, and the UI guided us towards
-building customizations that matched the system's model. For example,
-it's clear that it's not possible to customize a website using data not
-available in the table.
+building customizations that matched the system's model. If data is not
+available in the table, for example, then it's abundantly clear that
+customizations relying on that data will not be expressible.
 
@@
 
-Our second evaluation area is less about the conceptual approach of
-table-driven customization, and more about our specific implementation
-of customizing existing web applications. In order for third-party
+Our second evaluation area relates less to the conceptual approach of
+table-driven customization, and more to the specific implementation of
+customizing existing web applications. In order for third-party
 customization through Wildcard to succeed, it is important that creating
@@
 
-Hutchins, Hollan and Norman \citep{hutchins1985} describe a direct
-manipulation interface as one that uses a model-world metaphor, rather
-than a conversation metaphor. Rather than conversing with the system
-about an assumed world, ``the world is explicitly represented'' and the
-user can ``{[}act{]} upon the objects of the task domain themselves.''
-
-Most GUIs today employ direct manipulation, but software customization
-tools typically use an imperative programming model, which implements
-the conversational metaphor rather than direct manipulation. For
-example, in Applescript \citep{cook2007}, the scripting language for
-customizing Mac OS applications, here is how a user retrieves a list of
-of calendar names from the Calendar application:
+Hutchins, Hollan and Norman \citep{hutchins1985} define a direct
+manipulation interface as one that uses a model-world metaphor rather
+than a conversation metaphor. Instead of presenting an ``assumed'' but
+not directly visible world that the user converses with, ``the world is
+explicitly represented'' and the user can ``{[}act{]} upon the objects
+of the task domain themselves.''
+
+Although most GUIs today employ direct manipulation, software
+customization tools typically use an imperative programming model, which
+implements the conversational metaphor rather than direct manipulation.
+Here, for example, is how a user retrieves a list of of calendar names
+from the Calendar application in Applescript \citep{cook2007}, the
+scripting language for customizing Mac OS applications:
 
@@
 
-Some customization environments (Automator for Mac, Shortcuts for iOS,
-Zapier for web APIs) forego text syntax and enable the user to connect
-programs and construct automations by dragging and dropping commands.
-These environments still do not constitute direct manipulation, though:
-the objects being manipulated are in the domain of programming, not the
-domain of the task at hand.
+Some customization environments like Mac Automator and Zapier forego
+textual syntax and let the user connect programs and construct
+automations by dragging and dropping icons representing commands. These
+environments still do not constitute direct manipulation, though: the
+objects being manipulated are in the domain of programming, not in the
+domain of the task at hand. Imperative programming is a reasonable
+choice as the model for building customizations. Turing-complete
+programming provides a high ceiling for possible customizations, and a
+sequence of commands is a natural fit for automations that simulate a
+series of steps taken by the user.
 
@@
 
-However, there is a serious drawback to this approach. MacLean et al
+There is, however, a serious drawback to this approach. MacLean et al
 \citep{maclean1990} describe an ideal for user-tailorable systems: a
@@
 increases in skill lead to corresponding increments of customization
-power. Requiring users to learn programming for customization creates an
-abrupt ``cliff,'' where a significant learning investment is required
-even to implement simple customizations. Another goal of MacLean et al
-is that ``it should be as easy to change the environment as it is to use
-it,'' at least for some subset of changes. In scripting languages, even
-user-friendly ones like Applescript or Chickenfoot, the experience of
-customization does not remotely resemble the experience of use, so these
-systems can't meet this goal.
+power. Requiring users wanting to customize their applications to learn
+programming creates an abrupt ``cliff,'' exacting a significant
+investment in learning even to implement the simplest customizations.
+Another goal of MacLean et al is to make it ``as easy to change the
+environment as it is to use it''---at least for some subset of changes.
+But in scripting languages, even user-friendly ones like Applescript or
+Chickenfoot, the experience of customization does not remotely resemble
+the experience of use, so these systems can't meet this goal.
 
@@
 
-One aspect of directness which we have chosen not to maintain in
-Wildcard is enabling customization in the context of the original user
-interface, as explored by some other tools like Scotty
-\citep{eagan2011}. We have found that augmenting the original UI with an
-additional structured representation has numerous benefits. It provides
-a consistent experience across all applications, and clearly shows the
-user what structured data is available to work with.
+One aspect of directness that we have chosen not to maintain in Wildcard
+is enabling customization in the context of the original user interface,
+as explored by tools like Scotty \citep{eagan2011}. We have found that
+augmenting the original UI with a distinct, additional representation
+provides a consistent experience across all applications, and clearly
+shows the user what structured data is available to work with.
 
@@
 representations \citep{ainsworth1999}. In their terms, Wildcard plays a
-\emph{complementary role} by supporting a different set of tasks from
-the original application, while displaying \emph{shared information}. We
-also suspect that Wildcard helps \emph{constructs deeper understanding}
-by \emph{subtraction}. By stripping away details and only showing the
-essential data in an interface, Wildcard helps users think of new ways
-of using that data, outside the specific constraints of the original
-application.
+complementary role by supporting a different set of tasks from the
+original application, while displaying shared information. Wildcard may
+also help construct deeper understanding by subtraction. By stripping
+away details and only showing the essential data in an interface,
+Wildcard helps users think of new ways of using that data, outside the
+specific constraints of the original application.
 
@@
 cannot be reproduced in Wildcard. We consider this an acceptable
-tradeoff in exchange for achieving a gentler slope of customization, and
-we show in Section~\ref{sec:examples} and Section~\ref{sec:evaluation}
-that our model can still implement many useful customizations in
-practice. Also, there is sometimes a way to reframe an imperative script
-in terms of our direct manipulation model; for example, a script that
-iterates through rows in a page adding some additional information could
-be reproduced using a formula in Wildcard.
-
-\hypertarget{wrapping-applications-for-customization}{%
-\subsection{Wrapping applications for
-customization}\label{wrapping-applications-for-customization}}
+tradeoff in exchange for a gentler slope in customization, and we show
+in Section~\ref{sec:examples} and Section~\ref{sec:evaluation} that our
+model can still implement many useful customizations in practice. Also,
+there is sometimes a way to reframe an imperative script in terms of our
+direct manipulation model; for example, a script that iterates through
+rows in a page adding some additional information can be reproduced
+using a formula in Wildcard.
+
+\hypertarget{semantic-wrappers}{%
+\subsection{Semantic wrappers}\label{semantic-wrappers}}
 
@@
 
+\emph{todo: copy over more content from gdoc}
+
 With Wildcard, we use a hybrid approach that aims to provide the best of
@@
 Table-driven customization is most closely related to other tools that
-aim to empower end users to customize software without doing any
-programming.
+aim to empower end users to customize software without traditional
+coding.
 
@@
 Sifter \citep{huynh2006} enables end users to sort and filter lists of
-data based on web scraping, much like Wildcard's sorting features. The
-main difference between the systems is that table-driven customization
-has many other use cases besides sorting and filtering. Also, Sifter
-involves end users in a semi-automated data extraction process, rather
-than having programmers create wrappers. This provides coverage of more
-websites, but at the expense of complicating the end user experience. We
-might explore such techniques in Wildcard in the future, but we think
-that it's valuable for end users to have a customization experience free
-of web scraping details on supported websites. Finally Sifter implements
-scraping across multiple pages, a valuable feature for sorting and
-filtering which isn't present in Wildcard.
-
-Thresher \citep{hogue2005} enables end users to create wrappers which
-map website content to Semantic Web schemas, making customizations
-available using the schema information. We share the general idea of
-wrappers with Thresher, but we have chosen to map to a single generic
-data type, rather than more specific schemas, which increases the range
-of websites and data supported by the system.
-
-There are many software customization tools that offer simplified
-versions of programming for end users. Chickenfoot \citep{bolin2005} and
-Coscripter \citep{leshed2008} offer friendly syntax for writing web
-automation scripts; Applescript \citep{cook2007} attempts a similar goal
-for desktop customization. There are visual programming environments for
-customization which don't involve writing any text: as a few examples,
+data obtained by web scraping, much like Wildcard's sorting features.
+The main difference between the systems is that table-driven
+customization has many other use cases besides sorting and filtering.
+Also, Sifter involves end users in a semi-automated data extraction
+process, rather than having programmers create wrappers. This provides
+coverage of more websites, but at the expense of complicating the end
+user experience. We might explore such techniques in Wildcard in the
+future, but we think that it's valuable for end users to have a
+customization experience decoupled from the challenge of web scraping
+the underlying data. Sifter implements scraping across multiple pages, a
+valuable feature for sorting and filtering that isn't present in
+Wildcard.
+
+Thresher \citep{hogue2005} helps end users create wrappers that map
+website content to Semantic Web schemas, making customizations available
+using the schema information. Wildcard shares the general idea of
+wrappers with Thresher, but maps to a single generic data type, rather
+than more specific schemas, increasing the range of websites and data
+that Wildcard is able to support.
+
+There are many software customization tools that offer simplified forms
+of programming for end users. Chickenfoot \citep{bolin2005} and
+Coscripter \citep{leshed2008} offer user friendly syntax for writing web
+automation scripts; Applescript \citep{cook2007} has a similar goal for
+desktop customization. There are visual programming environments for
+customization that don't involve writing any text:
 \href{https://support.apple.com/guide/automator/welcome/mac}{Automator}
@@
 web applications together visually. As mentioned previously, these tools
-all require writing imperative programs in some form; we take a
-fundamentally different approach in table-driven customization.
+all require writing imperative programs in some form, in contrast to the
+more declarative and direct approach of table-driven customization.
 
@@
 
-Another relevant area of literature is direct manipulation interfaces
-for interacting with structured data. We take much inspiration from
-these tools in our work, but apply them in a different domain:
-customizing existing software applications, rather than interacting with
-databases or constructing software from scratch.
+Another relevant area involves direct manipulation interfaces for
+interacting with structured data. We take inspiration from these tools
+in our work, but apply them in a different domain: customizing existing
+software applications, rather than interacting with databases or
+constructing software from scratch.
 
@@
 \citep{liu2009}. SIEUFERD by Bakke and Karger \citep{bakke2016} is an
-example; their paper presents a survey of various other similar tools.
-Our work is particularly influenced by the authors' observation that
-direct manipulation requires that the user manipulate the \emph{results}
-of a database query rather than the query itself, and that the user must
-see intermediate results at every step of constructing a computation.
-SIEUFERD's interface supports a far wider range of possible queries than
-Wildcard, but the basic ideas of the user interface are similar.
+example; their paper presents a survey of other similar tools. Our work
+is particularly influenced by the authors' observation that direct
+manipulation requires that the user manipulate the results of a database
+query rather than the query itself, and that the user must see
+intermediate results at every step of constructing the query. SIEUFERD's
+interface supports a far wider range of queries than Wildcard, but the
+basic ideas of the user interface are similar.
 
@@
 
-Our work is also inspired by the many projects which have explored using
+Our work is also inspired by the many projects that have explored using
 spreadsheets as a foundation for building software applications,
@@
 al.~\citep{alrashed2020}, which enables end users to access backend web
-APIs without programming. ScrAPIr shares a high level goal of end user
-empowerment, and also shares the idea of wrappers, by creating a shared
-library of wrappers around existing web APIs. However, unlike Wildcard,
-ScrAPIr targets explicit APIs exposed by developers, and also doesn't
-aim to extend the frontend interfaces of web applications.
+APIs without programming. ScrAPIr shares our high level goal of end user
+empowerment, as well as the idea of wrappers, by creating a shared
+library of wrappers around existing web APIs. Unlike Wildcard, however,
+ScrAPIr targets explicit APIs exposed by developers. It also focused on
+backend services and doesn't aim to extend the frontend interfaces of
+web applications.
 

### 2020-05-22 paper.tex
full draft

@@
 
-Most attempts at empowering end users to customize their software offer
-a simplified version of programming. Some scripting languages
-\citep{bolin2005, cook2007} have a friendly syntax that resembles
-natural language. Some visual customization tools eliminate text syntax
-entirely. Macro recorders \citep{cook2007, chasins2018, anupam2000}
-remove some of the initial programming burden by letting a user start
-with concrete demonstrations.
+Many applications don't meet the precise needs of their users. End user
+customization systems can help improve the situation, by empowering
+non-programmers to modify their software in the way they would like.
+
+Most end user customization systems offer a simplified version of
+programming. Some scripting languages \citep{bolin2005, cook2007} have a
+friendly syntax that resembles natural language. Some visual
+customization tools eliminate text syntax entirely. Macro recorders
+\citep{cook2007, chasins2018, anupam2000} remove some of the initial
+programming burden by letting a user start with concrete demonstrations.
 
@@
 common: an imperative programming model, with statement sequencing,
-mutable variables and loops. End users express their ideas in
-scripts---sequences of commands---which, name aside, are not very
-different from conventional code.
+mutable variables and loops. End users express their ideas in scripts
+made up of sequences of commands.
 
@@
 scraping techniques to implement table-driven customization for existing
-Web applications. In Section~\ref{sec:examples}, we introduce the key
+Web applications. In Section~\ref{sec:example}, we introduce the key
 ideas of table-driven customization through an example scenario.
@@
 
-\hypertarget{sec:examples}{%
-\section{Example Scenario}\label{sec:examples}}
-
-To concretely illustrate the end user experience of table-driven
-customization, here is a real example of using the Wildcard browser
-extension to customize a website.
-
-\begin{itemize}
-\tightlist
-\item
-  Hacker News
-\item
-  open the table
-\item
-  sort by comments
-\item
-  take notes on links
-\item
-  filter out visited sites with a formula
-\item
-  add read times with a formula
-\end{itemize}
+\hypertarget{sec:example}{%
+\section{Example Scenario}\label{sec:example}}
+
+To illustrate the end user experience of table-driven customization, we
+consider a real example of using the Wildcard browser extension to
+customize Hacker News, a tech news aggregator. Figure 1 shows
+accompanying screenshots.
 
@@
 
+When the user opens Hacker News in a browser equipped with the Wildcard
+extension, they see a table at the bottom of the page, listing
+information about each link on the site: its title, URL, number of
+upvote points, etc. (Figure~\ref{fig:hacker-news}, Note A) The end user
+didn't need to do any work to produce this table; Wildcard contains a
+library of adapters created by programmers for extracting data from
+websites.
+
+The user's first goal might be to change the ranking of links on the
+homepage. Hacker News itself uses a ranking algorithm in which the
+position of an article depends not only on its point count (a measure of
+popularity), but also on how long it has been on the site. If the user
+hasn't been checking the site frequently, it'seasy to miss a popular
+article that has fallen down the page. Sorting the page solely by number
+of points would achieve a more stable ranking, with the most upvoted
+articles remaining at the top of the list until they are no longer on
+the home page.
+
+To achieve this sort order, the user simply clicks on the ``points''
+column header in the table. This sorts the table view by points, and the
+website UI also becomes sorted in the same order
+(Figure~\ref{fig:hacker-news}, Note B). This sort order persists across
+page loads, so every time the user loads Hacker News, they will see
+articles with this stable ordering.
+
+Next, the user might decide to attempt a more substantial customization:
+to show estimated read times next to each article, as additional context
+when deciding what to read. The table contains additional columns to the
+right of the original attributes for each link, where the user can enter
+spreadsheet-style formulas to compute derived values. The user enters a
+formula (Figure~\ref{fig:hacker-news}, Note C):
+
+\begin{verbatim}
+=ReadTimeInSeconds(link)
+\end{verbatim}
+
+This calls a built-in function that takes a URL as an argument and makes
+a request to a public web service that returns an estimated read time
+for the contents of that link. The result of evaluating the formula is
+simply a numerical value in seconds, which is not a particularly legible
+format. The user enters another formula in the next column
+(Figure~\ref{fig:hacker-news}, Note D):
+
+\begin{verbatim}
+=Concat(Round(user1/60), "min read")
+\end{verbatim}
+
+This converts seconds to minutes by dividing by 60 and rounding to the
+nearest integer, and then concatenates a label to the number. Finally,
+the user chooses to show the results of this new column in the original
+page (Figure~\ref{fig:hacker-news}, Note E). Each article on the page
+now shows an annotation with the estimated read time in minutes.
+
+The user might also manually add notes to the table without using
+formulas. In this case, the user jots down a few notes about articles
+they might want to read, and the notes appear in the page as well.
+
+The features presented in this scenario can support a broad range of
+other customizations. Section~\ref{sec:evaluation} explains many other
+use cases, but first we explain how the system works internally.
+
 \hypertarget{sec:architecture}{%
@@
 table-driven customization, using a simplified version of the Airbnb
-example above. In this example, the name of each listing is scraped from
-the web page DOM, the latitude and longitude of each listing is scraped
-from AJAX responses, and user annotations are loaded from the brower's
-local storage.
+example above. (\emph{todo: change this figure to hacker news}) In this
+example, the name of each listing is scraped from the web page DOM, the
+latitude and longitude of each listing is scraped from AJAX responses,
+and user annotations are loaded from the brower's local storage.
 
@@
 by looking for additional data columns joined to scraped rows, and
-rendering them in the page.
+rendering them in the page. (\emph{todo: explain joined})
 
@@
 the row in the page that corresponds to the selected record in the
-table.
+table. (\emph{todo: clarify})
 
@@
 implemented in Wildcard to power the customizations shown in
-Section~\ref{sec:examples}.
+Section~\ref{sec:example}.
 
@@
 the results of a formula expression with arithmetic or concatenating
-string labels, as shown in the example in Section~\ref{sec:examples}.
+string labels, as shown in the example in Section~\ref{sec:example}.
 (\emph{todo: this makes sense iff we switch the examples sectino to HN})
@@
 
-``Cell editors'' are UI widgets that expose a custom editing UI for a
-single cell of the table view. A programmer building a cell editor need
-only integrate it with the table viewer; propagating values into the
-website UI is handled by the site-specific DOM adapter. As a result,
-creating a new cell editor can be as easy as just writing some glue code
-between the table editor and an existing UI widget library.
+Cell editors are UI widgets that expose a custom editing UI for a single
+cell of the table view. A programmer building a cell editor need only
+integrate it with the table viewer; propagating values into the website
+UI is handled by the site-specific DOM adapter. As a result, creating a
+new cell editor can be as easy as just writing some glue code between
+the table editor and an existing UI widget library.
 
@@
 tradeoff in exchange for a gentler slope in customization, and we show
-in Section~\ref{sec:examples} and Section~\ref{sec:evaluation} that our
+in Section~\ref{sec:example} and Section~\ref{sec:evaluation} that our
 model can still implement many useful customizations in practice. Also,
@@
 Software customization tools typically fall into one of two categories:
-third-party, or semantic.
+ad hoc, or semantic.
 
@@
 
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-\emph{todo}
+\hypertarget{sec:conclusion}{%
+\section{Conclusion and Future Work}\label{sec:conclusion}}
+
+In this paper, we have presented table-driven customization, a new
+paradigm for customizing software by direct manipulation of the
+underlying structured data. We have demonstrated the paradigm using a
+browser extension, and have used it to create useful customizations for
+a variety of websites.
+
+Our primary goal for future work is to evaluate the system with a
+broader group of users. What barriers do end users face when using the
+system? What types of customizations do they choose to create? What
+formulas prove most helpful?
+
+Another area for future work is exploring whether it is possible to
+express a broader range of customizations by extending our table-editing
+paradigm. Are there ways to offer an increase in power and functional
+complexity, while retaining a programming model that is simpler for end
+users than conventional coding?
+
+As computing plays an ever greater role in our lives, it is increasingly
+important that end users have agency over the behavior of their
+software, rather than having every detail be dictated by companies whose
+incentives are not always aligned with the user's interests. We hope
+that table-driven customization can serve as one point, and maybe a
+guidepost, on our path from normal use to deep modification, in support
+of a more adaptable experience for all computer users.
 

### 2020-05-22 paper.tex
gs
3.1

@@
 %% The first command in your LaTeX source must be the \documentclass command.
-\documentclass[sigplan,10pt,anonymous,review]{acmart}
+\documentclass[sigplan,screen,10pt,anonymous,review]{acmart}
 
@@
 
+\author{Tyler Millis}
+\affiliation{%
+  \institution{Massachusetts Institute of Technology}
+  \city{Cambridge, MA}
+  \country{USA}
+}
+\email{tmillis@mit.edu}
+
+\author{Jessica Quaye}
+\affiliation{%
+  \institution{Massachusetts Institute of Technology}
+  \city{Cambridge, MA}
+  \country{USA}
+}
+\email{jquaye@mit.edu}
+
 %%
@@
   In this paper we show how the behavior of a software application can
-  be extended and adapted by direct manipulation, using table-driven
-  customization, a new paradigm that allows end users to customize
-  applications without writing traditional code.
-
-  Instead, users directly manipulate a tabular view of the structured
-  data inside the application---rather than writing imperative scripts
-  (as in most customization tools). This simple model also accommodates
-  a spreadsheet formula language and custom data editing widgets, which
+  be extended and adapted by direct manipulation. We introduce
+  \emph{table-driven customization}, a new paradigm where end users
+  directly manipulate a table view of the structured data inside an
+  application, rather than writing imperative scripts (as in most
+  customization tools). This simple model also accommodates a
+  spreadsheet formula language and custom data editing widgets, which
   provide sufficient expressivity to implement many useful
@@
   We illustrate the approach with Wildcard, a browser extension that
-  implements table-driven customization in the context of web
-  applications. Through concrete examples, we show that our paradigm can
-  be used to create useful customizations for real applications. We
-  share reflections from experience using the Wildcard system, on both
+  implements table-driven customization in the context of existing web
+  applications. Through concrete examples, we show that this paradigm
+  can support useful customizations for real websites. We share
+  reflections from our experiences using the Wildcard system, on both
   its strengths and limitations relative to other customization
@@
 programming. Some scripting languages \citep{bolin2005, cook2007} have a
-friendly syntax that resembles natural language. Some visual
+friendly syntax that resembles natural language. Other visual
 customization tools eliminate text syntax entirely. Macro recorders
@@
 programming burden by letting a user start with concrete demonstrations.
-
 Despite their many differences, these approaches all share something in
 common: an imperative programming model, with statement sequencing,
-mutable variables and loops. End users express their ideas in scripts
-made up of sequences of commands.
+mutable variables and loops.
 
@@
 manipulation is the \emph{de facto} standard in GUIs today, but when it
-comes to customizing software, it is rarely to be found. In this work,
-we ask: what would it look like to build a software customization
-interface that relies on direct manipulation? We take inspiration from
-spreadsheets and visual database query interfaces
-\citep{2020a, bakke2016}, which have successfully enabled end users to
-run queries and computations through direct manipulation of data.
+comes to customizing those GUIs, it is rarely to be found. As a result,
+switching from using software to customizing it poses a learning barrier
+for users not familiar with programming, and requires an abrupt shift in
+mental model even for those familiar with scripting.
+
+In this work, we ask: what would it look like to build a direct
+maniulation interface that lowers the threshold for starting to
+customize software? We take inspiration from spreadsheets and visual
+database query interfaces \citep{bakke2016, 2020a}, which have
+successfully enabled millions of end users to compute with data through
+direct manipulation.
 
 In this paper we present a technique called \emph{table-driven
-customization}, which applies ideas from visual query interfaces in the
-context of software customization. An application's UI is augmented with
-a table view, where the user can see and manipulate the application's
-internal data . Changes in the table view result in immediate
-corresponding changes to the original user interface of the application,
-enabling the user to customize an application with live feedback.
+customization} inspired by those tools. An application's UI is augmented
+with a table view, where the user can see and manipulate the
+application's internal data. Changes in the table view result in
+immediate corresponding changes to the original user interface, enabling
+the user to customize an application with live feedback.
+
+\begin{figure*}
+\hypertarget{fig:overview}{%
+\centering
+\includegraphics[width=\textwidth]{media/overview.eps}
+\caption{An overview of table-driven customization}\label{fig:overview}
+}
+\end{figure*}
 
@@
 scraping techniques to implement table-driven customization for existing
-Web applications. In Section~\ref{sec:example}, we introduce the key
-ideas of table-driven customization through an example scenario.
-
-In Section~\ref{sec:architecture}, we explain the architecture of
+Web applications. In Section~\ref{sec:example}, we illustrate the end
+user experience of table-driven customization through an example
+scenario using Wildcard.
+
+In Section~\ref{sec:architecture}, we explain the system architecture of
 table-driven customization. We focus on the \emph{table adapter}
@@
 bidirectionally mapped to a table. We describe several types of table
-adapters we've built in Wildcard, and also describe future adapters that
-are supported by the general paradigm.
-
-We have used Wildcard to build real customizations for 11 different
-websites. In Section~\ref{sec:evaluation}, we present reflections from
-this process, outlining the kinds of customizations we were able to
-build, limitations we encountered, and reflections on the ease of
-integrating scraping logic with real websites.
+adapters we've built in Wildcard, and also describe future adapters
+supported by the general paradigm.
+
+We have successfully used Wildcard to build customizations for 11
+different websites. In Section~\ref{sec:evaluation}, we present
+reflections from this process, outlining the kinds of customizations we
+were able to build, limitations we encountered, and the challenges of
+integrating scraping logic with websites.
 
@@
 \item
-  \emph{Customization by direct manipulation}: End users should be able
-  to customize an application by directly examining and modifying its
-  data, rather than by writing imperative scripts.
+  \emph{Customization by direct manipulation}: We agree with MacLean et
+  al \citep{maclean1990} that customization should follow a ``gentle
+  slope'', where users can gradually invest in more complex techniques
+  to achieve deeper customizations. We believe an important point on
+  this slope is the ability to customize an application by directly
+  seeing and changing its internal structured data, rather than by
+  writing imperative scripts.
 \item
@@
 To illustrate the end user experience of table-driven customization, we
-consider a real example of using the Wildcard browser extension to
-customize Hacker News, a tech news aggregator. Figure 1 shows
-accompanying screenshots.
+consider an example scenario of customizing Hacker News, a tech news
+aggregator. Figure~\ref{fig:hacker-news} shows accompanying screenshots.
 
@@
 
-When the user opens Hacker News in a browser equipped with the Wildcard
-extension, they see a table at the bottom of the page, listing
-information about each link on the site: its title, URL, number of
-upvote points, etc. (Figure~\ref{fig:hacker-news}, Note A) The end user
-didn't need to do any work to produce this table; Wildcard contains a
-library of adapters created by programmers for extracting data from
-websites.
-
-The user's first goal might be to change the ranking of links on the
-homepage. Hacker News itself uses a ranking algorithm in which the
-position of an article depends not only on its point count (a measure of
-popularity), but also on how long it has been on the site. If the user
-hasn't been checking the site frequently, it'seasy to miss a popular
-article that has fallen down the page. Sorting the page solely by number
-of points would achieve a more stable ranking, with the most upvoted
-articles remaining at the top of the list until they are no longer on
-the home page.
-
-To achieve this sort order, the user simply clicks on the ``points''
+\textbf{Opening the table}: When the user opens Hacker News in a browser
+equipped with the Wildcard extension, they see a table at the bottom of
+the page. It contains a row for each link on the homepage, listing
+information like the title, URL, submitter username, number of points,
+and number of comments (Figure~\ref{fig:hacker-news}, Note A). The end
+user didn't need to do any work to create this table, because a
+programmer previously created an adapter to extract data from this
+particular website, and contributed it to a shared library of adapters
+integrated into Wildcard.
+
+\textbf{Sorting by points}: First, the user decides to change the
+ranking of links on the homepage. Hacker News itself uses a ranking
+algorithm in which the position of an article depends not only on its
+point count (a measure of popularity), but also on how long it has been
+on the site. If the user hasn't been checking the site frequently, it's
+easy to miss a popular article that has fallen lower on the list.
+Sorting the page just by points would achieve a more stable ranking.
+
+To achieve this ordering, the user simply clicks on the ``points''
 column header in the table. This sorts the table view by points, and the
 website UI also becomes sorted in the same order
-(Figure~\ref{fig:hacker-news}, Note B). This sort order persists across
-page loads, so every time the user loads Hacker News, they will see
-articles with this stable ordering.
-
-Next, the user might decide to attempt a more substantial customization:
-to show estimated read times next to each article, as additional context
-when deciding what to read. The table contains additional columns to the
-right of the original attributes for each link, where the user can enter
-spreadsheet-style formulas to compute derived values. The user enters a
-formula (Figure~\ref{fig:hacker-news}, Note C):
+(Figure~\ref{fig:hacker-news}, Note B)---Wildcard has manipulated the
+website's DOM to synchronize it with the sort order of the table. This
+sort order is also persisted in the browser and reapplied automatically
+the next time the user loads the page, so they can always browse the
+latest set of links using this sort order.
+
+\textbf{Adding estimated read times}: Next, the user decides to attempt
+a more substantial customization: to add estimated read times to
+articles on Hacker News.
+
+The table contains additional columns to the right of the original
+attributes for each row, where the user can enter spreadsheet-style
+formulas to compute derived values. We provide generic names for these
+columns by default as spreadsheets do, so that users don't need to
+choose column names themselves. The user enters a formula into the
+\texttt{user1} column: (Figure~\ref{fig:hacker-news}, Note C):
 
@@
 
-This calls a built-in function that takes a URL as an argument and makes
-a request to a public web service that returns an estimated read time
-for the contents of that link. The result of evaluating the formula is
-simply a numerical value in seconds, which is not a particularly legible
-format. The user enters another formula in the next column
-(Figure~\ref{fig:hacker-news}, Note D):
+This calls a built-in function that takes a URL as an argument and uses
+a public web API to compute an estimated read time for the link's
+contents. The argument \texttt{link} refers to a column name in the
+table; for each row, the formula is evaluated with the link for that
+particular row. The result of evaluating this formula is a numerical
+value in seconds for each article.
+
+The user clicks the \texttt{user1} column header to sort the articles on
+the page in descending order of estimated read time, helping to
+prioritize reading deeper content. It would be useful to show these read
+times in the page too, but before doing that, the user wants to make the
+time estimates more readable. They enter another formula in the next
+column, \texttt{user2}:
 
@@
 
-This converts seconds to minutes by dividing by 60 and rounding to the
-nearest integer, and then concatenates a label to the number. Finally,
-the user chooses to show the results of this new column in the original
-page (Figure~\ref{fig:hacker-news}, Note E). Each article on the page
-now shows an annotation with the estimated read time in minutes.
-
-The user might also manually add notes to the table without using
-formulas. In this case, the user jots down a few notes about articles
-they might want to read, and the notes appear in the page as well.
-
-The features presented in this scenario can support a broad range of
-other customizations. Section~\ref{sec:evaluation} explains many other
-use cases, but first we explain how the system works internally.
+This formula converts seconds to minutes by dividing by 60 and rounding
+to the nearest integer, and then concatenates a label to the number,
+producing strings like ``21 min read''.
+
+Finally, the user clicks a menu option in the table header to display
+the contents of this new column in the original page
+(Figure~\ref{fig:hacker-news}, Note D). Each article on the page now
+shows an annotation with the estimated read time in minutes. (The format
+of how annotations appear for each item was determined by the programmer
+who created the Wildcard adapter for Hacker News.)
+
+\textbf{Adding notes to links}: The user can also manually add notes to
+the table, by simply entering values into the table without using
+formulas. In this case, the user jots down a few notes in another column
+about articles they might want to read, and the notes appear in the page
+next to the read times (Figure~\ref{fig:hacker-news}, Note D).
+
+\textbf{Filtering out visited links}: Another way to use formulas to
+customize Hacker News is to filter out articles the user has already
+read. (We omit this example from the figure for brevity.) The user can
+call a built-in function that returns a boolean depending on whether a
+URL is in the browser's history:
+
+\begin{verbatim}
+=Visited(link)
+\end{verbatim}
+
+They can then filter the table to only contain rows where this formula
+column contains \texttt{false}; visited rows are hidden both from the
+table view and the original page. This is an example of a customization
+that the original website could not implement, since websites don't have
+access to the browser history for privacy reasons. But by using
+Wildcard, the user can implement the customization locally, without
+needing to expose their entire browser history to Hacker News.
+
+This scenario has shown a few examples of how table-driven
+customizations can help a user improve their experience of a website.
+Section~\ref{sec:evaluation} explains many other use cases, but first we
+explain how the system works internally.
 
@@
 view, an ordered list of visible IDs is sent to each table adapter. The
-DOM scraping adapter uses this information to modify the list of visible
-rows in the web page.
+DOM scraping adapter uses this information to change the list of rows
+shown in the web page.
 

### 2020-05-22 paper.tex
actually finished 3.1

@@
 about articles they might want to read, and the notes appear in the page
-next to the read times (Figure~\ref{fig:hacker-news}, Note D).
+next to the read times (Figure~\ref{fig:hacker-news}, Note D). The
+annotations are also stored in the browser's local storage so they can
+be retrieved on future visits.
 
@@
 Figure~\ref{fig:table-adapter} summarizes the overall architecture of
-table-driven customization, using a simplified version of the Airbnb
-example above. (\emph{todo: change this figure to hacker news}) In this
-example, the name of each listing is scraped from the web page DOM, the
-latitude and longitude of each listing is scraped from AJAX responses,
-and user annotations are loaded from the brower's local storage.
-
-First, the three different data sources are each bidirectionally mapped
-to a table interface by a \textbf{table adapter}. The table adapter
-defines how to map a particular type of data to a table, and what
-effects edits should have on the original data source. In some cases,
-the mapping logic is straightforward: the local storage adapter stores a
-table of data, so the mapping to the table abstraction is trivial. In
-other cases, the mapping is more involved: the DOM scraping adapter
-implements web scraping logic to produce a table of data from the web
-page, and turns edits to the table into DOM manipulations like
+table-driven customization, using a simplified illustration of the
+Hacker News example scenario. In this example, the name and points value
+for each article is scraped from the web page DOM, and user annotations
+are loaded from the brower's local storage.
+
+First, the web page and the browser storage are each wrapped by a
+\textbf{table adapter}, which defines a bidirectional mapping between
+some underlying data source and a table.
+
+In addition to a \emph{read} mapping for how the underlying data should
+be represented as a table, it also defines a \emph{write} mapping for
+what effects edits should have on the original data source. In this
+case, the local storage adapter has a trivial mapping: it loads a table
+of data stored in the browser, and persists edits to that persisted
+state. The mapping logic of the DOM scraping adapter is much more
+involved. It implements web scraping logic to produce a table of data
+from the web page, and turns edits into DOM manipulations, such as
 reordering rows of data on the page.
 
-The three separate tables are then combined into a single table for the
-end user to view and edit. The \textbf{query engine} is responsible for
-creating this combined view, and routing the user's edits back to the
-individual table adapters. In this example, the query engine has joined
-the three tables together by a shared ID column, and sorted the result
-by the name column.
+The two tables are then combined into a single table for the end user to
+view and edit. The \textbf{query engine} is responsible for creating
+this combined view, and routing the user's edits back to the individual
+table adapters. In this example, the query engine has joined the two
+tables together by a shared ID column, and sorted the result by the
+points column.
 
@@
 A key idea in table-driven customization is that a wide variety of data
-sources can be mapped to the generic abstraction of an editable table.
-In a relational database, the table matches the underlying storage
-format, but in table-driven customization, the table is merely an
-\emph{interface layer}. The data shown in the table is a projection of
-some underlying state, and edits to the table can have complex effects
-on the underlying state.
-
-Externally, a table adapter must satisfy an abstract interface. The
-first two parts of the interface resemble the interface exposed by a
-table in a relational database:
-
-\textbf{Returning a table}: A table adapter exposes a table of data: an
+sources can be mapped to a generic table abstraction. In a relational
+database, the table matches the underlying storage format, but in
+table-driven customization, the table is merely an \emph{interface
+layer}. The data shown in the table is a projection of some underlying
+state, and edits to the table can have complex effects on the underlying
+state.
+
+The first two parts of the table adapter interface resemble a typical
+database table:
+
+\emph{Returning a table}: A table adapter exposes a table of data: an
 ordered list of records. Each record carries a unique identifier and
 associates named attributes with values. Tables have a typed schema, so
-the same attributes are shared across all records. A table adapter can
-update the contents of a table at any time, in response to changes in
-the underlying state (e.g., a DOM scraping adapter can update the table
-when the page body changes). When data changes, the query view is
-reactively updated in response.
-
-\textbf{Making edits}: The query engine can issue a request to a table
+the same attributes are shared across all records. The columns also
+carry some additional metadata in addition to their type, such as
+whether or not they are read-only or editable.
+
+A table adapter can update the contents of a table at any time in
+response to changes in the underlying state (e.g., a DOM scraping
+adapter can update the table when the page body changes). When data
+changes, the query view is reactively updated in response.
+
+\emph{Making edits}: The query engine can issue a request to a table
 adapter to make an edit to a record. The meaning of making an edit can
@@
 annotation may be persisted into local storage; in the DOM scraping
-adapter, an edit may represent filling in a form field. An adapter can
-also mark values as read-only if it wouldn't be meaningful to edit them;
-for example, the DOM scraping adapter typically marks page content as
-read-only, except for editable form fields.
-
-The query engine also sends additional information about the combined
-query view to each table adapter. These functions are currently used to
-provide the DOM scraping adapter with sufficient information to
-manipulate the original application UI as the user manipulates the table
-view.
-
-\textbf{Sorting/filtering}: When the user sorts or filters the query
-view, an ordered list of visible IDs is sent to each table adapter. The
-DOM scraping adapter uses this information to change the list of rows
-shown in the web page.
-
-\textbf{Information from other tables}: The query engine sends each
-table adapter the entire combined table of data being rendered to the
-user. The DOM scraping adapter uses this to add annotations to the page
-by looking for additional data columns joined to scraped rows, and
-rendering them in the page. (\emph{todo: explain joined})
-
-\textbf{Currently selected record}: The query engine notifies each table
-adapter about which record is currently selected by the user in the
-table UI. The DOM scraping adapter uses this information to highlight
-the row in the page that corresponds to the selected record in the
-table. (\emph{todo: clarify})
+adapter, an edit may represent filling in a form field.
+
+In addition, the query engine also sends additional information about
+the combined query view to each table adapter:
+
+\emph{Sorting/filtering}: When the user sorts or filters the query view,
+an ordered list of visible IDs is sent to each table adapter. The DOM
+scraping adapter uses this information to change the list of rows shown
+in the web page.
+
+\emph{Data from other tables}: The query engine provides each table
+adapter with the entire combined table shown to the user. The DOM
+scraping adapter uses this for injecting annotations---values in
+additional columns from other tables are added to the original web page.
+
+\emph{Currently selected record}: As the user clicks in different parts
+of the table view, the query engine communicates the record currently
+selected by the user. The DOM scraping adapter uses this information to
+highlight the row in the page that corresponds to the selected row in
+the table (and scroll to the row if it's not currently showing on the
+page), which helps clarify the connection between the table and the
+original UI.
 
@@
 
-Here we describe in more detail the table adapters that we have
-implemented in Wildcard to power the customizations shown in
-Section~\ref{sec:example}.
+In Wildcard, we have built three specific types of table adapters so
+far.
 
@@
 available for customization in Wildcard. To make this approach viable,
-we have built a library of functions that make it easy to implement a
-scraping adapter, requiring the programmer only to implement the minimal
-site-specific parts. The programmer writes a single Javascript function
-which can use standard scraping techniques like CSS selectors and DOM
-APIs to extract relevant elements from the page. The library functions
-then wrap this function to implement the rest of the needed
-functionality; for example, when the table is sorted, we find the DOM
-elements corresponding to the rows in the table, remove them from the
-page, and then reinsert them in the new order.
+we have built a generic DOM scraping adapter which a programmer can
+configure with the minimal site-specific parts.
+
+After specifying basic metadata like which URLs the adapter should apply
+to, the programmer only needs to implement a single Javascript function
+that scrapes relevant elements from the page. They are free to use any
+scraping techniques; in practice simple CSS selectors and DOM APIs
+usually suffice. he generic adapter then wraps this scraping function to
+implement the table adapter interface. For example, when the table view
+is sorted, the generic adapter takes the DOM elements corresponding to
+the rows in the table, removes all of them from the page, and then
+reinserts them in the new order.
+
+The programmer can optionally override other default behavior of the
+generic DOM scraping adapter, including the logic for when to re-run the
+scraping function in response to page changes, how to style injected
+annotations, and how to style the currently selected row in the table.
 
 An \textbf{AJAX scraping adapter} intercepts AJAX requests made by a web
-page, and extracts information from those requests. When available, this
-tends to be a helpful technique because the data is already in a
-structured form, and often includes information not shown in the UI. As
-with DOM scraping adapters, we have made it easy for programmers to
+page, and extracts information from those requests to add to the table.
+When available, this tends to be a helpful technique because the data is
+already in a structured form so it is easier to scrape, and it often
+includes valuable information not shown in the UI.
+
+As with DOM scraping adapters, we have made it easy for programmers to
 create site-specific AJAX scraping adapters. A programmer writes a
@@
 
+In order to join the tables produced by AJAX scraping and DOM scraping,
+a common set of identifiers is required across records in the two
+tables. Often there is a server-defined ID present both in the DOM and
+in AJAX responses; if not, the programmer can use some set of
+overlapping data (e.g.~an item name) as a shared ID.
+
 The \textbf{local storage adapter} simply stores a table of data in the
-browser. As shown in the example use cases, this can be useful for
-persisting annotations in a private way, without uploading them to a web
-service.
+browser. This is currently only used to persist annotations: the user
+makes edits in the table, they are stored privately in the browser, and
+can be loaded into the table on future pageloads.
 
@@
 
-\emph{Exposing a table}: In Redux, app developers already represent the
-state of a user interface as a single object. To expose a table of data,
-the developer simply writes a function that maps the state object to a
-table. \emph{Editing data values}: The developer defines how edits to
-records in the table should map to edits in the state object.
+\begin{itemize}
+\tightlist
+\item
+  \emph{Exposing a table}: In Redux, app developers already represent
+  the state of a user interface as a single object. To expose a table of
+  data, the developer simply writes a function that maps the state
+  object to a table.
+\item
+  \emph{Editing data values}: The developer defines how edits to records
+  in the table should map to edits in the state object.
+\item
   \emph{Sorting/filtering the UI}: The developer inserts an additional
   data processing step before the state object is rendered in the UI, so
-that Wildcard can apply sorting and filtering. \emph{Annotations}: The
-developer adds additional React components managed by Wildcard into
-their UI tree, which render additional data as annotations.
+  that Wildcard can apply sorting and filtering.
+\item
+  \emph{Annotations}: The developer adds additional React components
+  managed by Wildcard into their UI tree, which render additional data
+  as annotations.
+\end{itemize}
 
@@
 
+\emph{todo: mention ajax scraping} \emph{todo: mention privacy benefits}
+
 \hypertarget{sec:themes}{%

### 2020-05-22 paper.tex
finish section 3

@@
 
-\emph{Making edits}: The query engine can issue a request to a table
+\emph{Handling edits}: The query engine can issue a request to a table
 adapter to make an edit to a record. The meaning of making an edit can
@@
 scraping techniques; in practice simple CSS selectors and DOM APIs
-usually suffice. he generic adapter then wraps this scraping function to
-implement the table adapter interface. For example, when the table view
-is sorted, the generic adapter takes the DOM elements corresponding to
-the rows in the table, removes all of them from the page, and then
+usually suffice. The generic adapter then wraps this scraping function
+to implement the table adapter interface. For example, when the table
+view is sorted, the generic adapter takes the DOM elements corresponding
+to the rows in the table, removes all of them from the page, and then
 reinserts them in the new order.
@@
 
-In certain types of software, customizability is often touted as a key
-selling point. An integrated website adapter would provide a way for
-developers to integrate with an existing ecosystem of formulas and
-customization tools, without needing to build all that functionality
-from scratch.
+Customizability is sometimes a key selling point for software. An
+integrated website adapter would provide a way for developers to
+integrate with an ecosystem of formulas and customization tools without
+needing to build all that functionality from scratch.
 
@@
 collaboratively taking notes with friends on options for places to stay
-on Airbnb. The existing Local Storage Adapter could be extended to share
-live synchronized data with other users. This could be achieved through
-a centralized web server, or through P2P connections that might provide
-stronger privacy guarantees.
+together on a trip. The existing local storage adapter could be extended
+to share live synchronized data with other users. This could be achieved
+through a centralized web server, or through P2P connections that might
+provide stronger privacy guarantees.
 
@@
 Queries are processed in three steps. First, the query invokes a primary
-DOM scraping table adapter that associates records in the result with
-elements in the application's user interface. At minimum, the primary
-table adapter needs to return record IDs and have the ability to
-manipulate the application's UI. It can also optionally return data
-about each record. Next, additional tables (AJAX data, local storage
-data) are left-joined by ID. Finally, the result table can be sorted and
-filtered by any column.
-
-This query model can be viewed as a tiny (and rather constrained) subset
-of the SQL query model. Despite its simplicity, this simple model has
-proven sufficient for meeting the needs of customization in practice,
-and minimizes the complexity of supporting more general and arbitrary
-queries. But because it fits into the general paradigm of relational
-queries, it could theoretically be extended to support a wider range of
-queries.
+DOM scraping adapter that associates table rows with elements in the
+application's user interface. Next, additional tables (AJAX data, local
+storage data) are left-joined by ID. Finally, the result table is sorted
+and filtered according to user-specified predicates.
+
+One way to view this query model is as a tiny subset of the SQL query
+model. Despite its simplicity, this model has proven sufficient for
+meeting the needs of customization in practice, and minimizes the
+complexity of supporting arbitrary queries. But because it fits into the
+general paradigm of relational queries, it could theoretically be
+extended to support a wider range of queries.
 
 The query engine is also responsible for executing formulas. We have
-built a small formula language resembling the ones used in visual query
-tools like SIEUFERD. As in those tools, and unlike in spreadsheets,
-formulas automatically apply across an entire column of data, and
-reference other column names instead of values on specific rows. For
-example, the user would write \texttt{=price\ *\ 2} to create a column
-with a value derived from a price attribute. This is more convenient
-than needing to copy-paste a formula across an entire column as in
-spreadsheets, and none of our customization use cases have uncovered a
-need for writing different formulas on different rows of the table.
+built a small formula language resembling a spreadsheet formula
+language, except that formulas automatically apply across an entire
+column of data, and reference other column names instead of values in
+specific rows. This is more convenient than needing to copy-paste a
+formula across an entire column as in spreadsheets, and has worked for
+all of the customizations we have built.
 
@@
 
-In Wildcard we provide a basic table editor as the user interface on top
-of the query engine. It is built with the Handsontable Javascript
-library, which provides UI elements for viewing and editing a table, as
-well as basic query operations like sorting and filtering.
-
-\emph{todo: other table instruments}
+We provide a basic table editor as the user interface on top of the
+query engine. Our table editor is built with the Handsontable Javascript
+library, which provides UI elements for viewing, editing, sorting, and
+filtering a table.
+
+In addition to the basic table editing operations, we also provide
+\emph{cell editors}: UI widgets that expose a custom editing UI for a
+single cell of the table view. A programmer building a cell editor need
+only integrate it with the table viewer; propagating values into the
+website UI is handled by the site-specific DOM adapter. In
+Section~\ref{sec:evaluation} we provide some examples of using cell
+editors.
+
+The table editor only serves as a shallow interface layer over the query
+engine, relaying user commands to the query engine and rendering the
+resulting data table. Because of this architectural split, it would be
+straightforward to develop additional table editor interfaces on top of
+the Wildcard system. One intruiging future possibility is to apply the
+idea of cell editors to the entire table, by developing table editors
+which don't look like a table. For example, we could provide a calendar
+view for displaying a table containing a date column.
+
+\hypertarget{sec:evaluation}{%
+\section{Evaluation}\label{sec:evaluation}}
 
@@
 
-\hypertarget{sec:evaluation}{%
-\section{Evaluation}\label{sec:evaluation}}
-
 To evaluate table-driven customization in practice, we built the
@@
 \subsubsection{Cell editors}\label{cell-editors}}
-
-(\emph{note: written with the assumption that Expedia gets cut from the
-Examples section})
-
-Cell editors are UI widgets that expose a custom editing UI for a single
-cell of the table view. A programmer building a cell editor need only
-integrate it with the table viewer; propagating values into the website
-UI is handled by the site-specific DOM adapter. As a result, creating a
-new cell editor can be as easy as just writing some glue code between
-the table editor and an existing UI widget library.
 

### 2020-05-22 paper.tex
finish section 4

@@
 We have successfully used Wildcard to build customizations for 11
-different websites. In Section~\ref{sec:evaluation}, we present
+different websites. In Section~\ref{sec:reflections}, we present
 reflections from this process, outlining the kinds of customizations we
@@
 customizations can help a user improve their experience of a website.
-Section~\ref{sec:evaluation} explains many other use cases, but first we
-explain how the system works internally.
+Section~\ref{sec:reflections} explains many other use cases, but first
+we explain how the system works internally.
 
@@
 
-In Wildcard, we have built three specific types of table adapters so
-far.
+We have built three specific types of table adapters so far, to enable
+web customization using Wildcard.
 
@@
 We have designed the table adapter API to be general enough to support
-other types of useful adapters in the future. Here are two such
-possibilities:
+other types of useful adapters in the future. Here are two examples:
 
@@
 website UI is handled by the site-specific DOM adapter. In
-Section~\ref{sec:evaluation} we provide some examples of using cell
+Section~\ref{sec:reflections} we provide some examples of using cell
 editors.
@@
 
-\hypertarget{sec:evaluation}{%
-\section{Evaluation}\label{sec:evaluation}}
+\hypertarget{sec:reflections}{%
+\section{Reflections on Usage}\label{sec:reflections}}
 
@@
 
-We have used Wildcard to customize 11 websites, including transactional
+We built Wildcard integrations for 11 websites, including transactional
 sites like Amazon and Uber Eats, and media consumption sites like Hacker
-News and Youtube. Table 1 summarizes these customizations, showing the
-number of lines of code in the adapter configuration for each site.
+News and Youtube. Table 1 summarizes these integrations, showing the
+number of lines of code in the adapter configuration for each site, and
+some example customizations we were able to perform for each site.
 
@@
 useful purposes. Here we expand on some archetypal examples that
-illuminate various aspects of using the system in practice.
+illuminate aspects of using the system in practice.
 
@@
 
-Sometimes, websites have opaque algorithms which they use to rank
-results, which presumably maximize profit but don't offer much control
-to the user. For example, Airbnb doesn't allow users to sort listings by
-price, and Youtube doesn't allow users to sort the homepage by watch
-time. Wildcard enables users to take back some control over this
-process.
-
-In other cases we have found that a lack of sorting options seems less
-intentional, and simply represents a missing feature. For example, the
-Instacart grocery delivery service doesn't allow users to sort their
-grocery cart by price or category. Our guess is that they simply haven't
-gotten around to implementing the feature yet.
+Sometimes, websites have opaque ranking algorithms which presumably
+maximize profit but don't offer much control to the user. For example,
+Airbnb previously allowed users to sort listings by price, but removed
+that feature in 2012. We have found other sites where a lack of sorting
+options seems more like merely an omission; for example, the Instacart
+grocery delivery service has a spartan UI for viewing a pending order,
+which doesn't allow for sorting by price or category. In both of these
+cases, Wildcard enables users to take back some control.
 
@@
 users are not entirely liberated from the suggestions of the opaque
-algorithm. We could work around this restriction in the future by
-fetching content across multiple pages, but we've also found that
-sorting/filtering a single page of a paginated list is often an
-acceptable outcome (and sometimes even a preferable one). For example,
-it's more useful to sort (by length, say) 30 recommended Youtube videos
-than to try to sort all videos on Youtube, or to sort (by read time,
-say) the articles on the Hacker News front page than to sort all
-non-archived articles.
+algorithm. This restriction could be overcome in the future by scraping
+content across multiple pages, or by using an integrated adapter built
+in to the application. However, we've also found that sorting/filtering
+a single page of a paginated list is often an acceptable outcome (and
+sometimes even a preferable one). It's more useful, for example, to sort
+30 recommended Youtube videos than to try to sort all videos on Youtube.
 
@@
 
-Many other web annotation systems focus on annotating text or arbitrary
+Many web annotation systems focus on annotating text or arbitrary
 webpage content, but Wildcard limits annotations to structured objects
-extracted by an adapter, resulting in quite a different set of use
-cases.
-
+extracted by an adapter, resulting in a different set of use cases.
 Annotating with Wildcard has proven most useful when taking notes on a
@@
 language supports only a small number of predefined functions. Adding
-more will allow a very broad range of useful computations, as shown by
-the success of spreadsheets.
+more should allow a broad range of useful computations, as shown by the
+success of spreadsheets.
 
@@
 the current language design is that supporting a new web API requires
-writing conventional code to add a new function to the language, because
+writing Javascript code to add a new function to the language, because
 web APIs typically return complex JSON data structures that can't be
-easily displayed in a single table cell. Adding functions for handling
-JSON data to the formula language might make it possible to access new
-APIs without requiring an entire new function in the formula system.
-
-We have also found instances where simple data manipulation is useful:
-adding together different subtotals into a total price, or manipulating
-the results of a formula expression with arithmetic or concatenating
-string labels, as shown in the example in Section~\ref{sec:example}.
-(\emph{todo: this makes sense iff we switch the examples sectino to HN})
+easily displayed in a single table cell. In the future we would like to
+make it possible to call new APIs without adding a dedicated function,
+which might require adding functions to the formula language that can
+manipulate JSON data.
+
+We have also found instances where simple data manipulation is useful,
+e.g.~transforming the results of an API call with basic arithmetic and
+string operations, as shown in Section~\ref{sec:example}.
 
@@
 
+We developed two cell editors to explore different benefits of adding
+custom UI for editing the table.
+
 An important benefit of cell editors is that they allow users to
-incorporate their personal information within the web UI but without
-uploading their personal data to the web server. To explore this idea,
-we created a cell editor based on the FullCalendar Javascript plugin,
-which can load data from a Google Calendar. This makes it convenient to
-enter dates into a website based on the user's personal calendar
-information.
+incorporate their private information into a web UI without uploading it
+to a website. To explore this idea, we created a datepicker based on the
+\href{https://fullcalendar.io/}{FullCalendar} plugin, which can load
+data from a Google Calendar. This makes it convenient to enter dates
+into a website based on the user's personal calendar information.
 
 Another benefit of cell editors is that a user can choose their
-preferred widget for editing some type of information. We built a cell
-editor based on the CKEditor rich text editor, and demonstrated its use
-by integrating it with Google's Blogger website and using it to edit
-blog posts, in place of Blogger's built-in editor.
+preferred widget for editing some type of information. We built a text
+editor based on the \href{https://ckeditor.com/}{CKEditor} rich text
+editor, and demonstrated its use by integrating it with Google's Blogger
+website and using it to edit blog post content.
 
@@
 
-There are many useful customizations that are not currently possible in
-the table-driven customization paradigm. Wildcard can only make
-customizations that use the available data exposed in the table. UI
-modifications are limited in scope; deleting arbitrary buttons isn't
-possible, for example. At one point, we wanted to build an automation to
-repeatedly load a grocery delivery website to check for open delivery
-slots, but it wasn't clear how to achieve this. Some of these
-limitations are specific to the current implementation of the Wildcard
-extension, but some are more fundamental to the entire idea of
-table-driven customization.
-
-We consider this an acceptable outcome; our goal is to make many useful
-customizations available to end users with a low threshold, and not to
-span all possible customizations. Section 5.1 discusses this point
-further.
-
-One benefit of Wildcard is predictability: once we built a site adapter
-for a website, it was generally obvious to us what types of
-customizations were and weren't possible, and the UI guided us towards
-building customizations that matched the system's model. If data is not
-available in the table, for example, then it's abundantly clear that
-customizations relying on that data will not be expressible.
-
-\hypertarget{viability-of-dom-scraping}{%
-\subsection{Viability of DOM scraping}\label{viability-of-dom-scraping}}
+There are many customizations that are not possible to implement with
+table-driven customization. Some of the limitations are specific to the
+current implementation of the Wildcard extension, but many are more
+fundamental to the general paradigm.
+
+One limitation is that Wildcard can only make customizations that use
+the available data exposed in the table. If the adapter doesn't expose
+some piece of data, the user can't use it in their customization. The
+table data format also rules out customizing certain sites that don't
+have a way to map to a table. The UI modifications available in Wildcard
+are also limited in scope; deleting arbitrary buttons isn't possible,
+for example. There is no facility for running automations when the user
+isn't actively viewing a page---at one point, we wanted to build an
+automation to repeatedly load a grocery delivery website to check for
+open delivery slots, but it didn't seem possible to achieve this in
+Wildcard.
+
+Despite these limitations, we found that Wildcard does offer
+predictability: once we built a site adapter for a website, it was
+generally obvious what customizations were possible, and the UI guided
+us towards building customizations that matched the system's model. If
+data is not available in the table, for example, then it's abundantly
+clear that customizations relying on that data will not be expressible.
+
+We consider all of these limitations to be an acceptable outcome. Our
+goal is to support as many useful customizations as possible with a low
+threshold of difficulty, and not to span all possible customizations. If
+users want to implement more sophisticated customizations, they have the
+option of graduating to more advanced customization tools.
+
+\hypertarget{viability-of-scraping}{%
+\subsection{Viability of scraping}\label{viability-of-scraping}}
 
@@
 
-Most of our DOM scraping adapters were created by members of our team.
-However, an external developer unaffiliated with the project contributed
-one adapter, designed to sort the Github page listing a user's
-repositories. They described the experience as ``very straightforward.''
+Nearly all of our DOM scraping adapters were created by members of our
+team. However, an external developer unaffiliated with the project
+contributed one adapter, designed to sort the Github page listing a
+user's repositories, and they described the experience as ``very
+straightforward.''
 
 The adapters for our test sites ranged from 36 to 117 lines of code,
-averaging 68 lines; Table 1 shows the full lines of code for each
-adapter.
+averaging 68 lines; Table 1 shows the number of lines of code for each
+adapter. Most of the code in the adapters is simply using DOM APIs and
+CSS selectors to implement conventional web scraping logic.
 
@@
 ones as with writing normal web scraping code, but the more interactive
-nature of Wildcard introduces additional challenges. rve changes, but it
-may prove challenging to observe changes on some sites only through the
-DOM. One challenge is triggering updates to the spreadsheet data in
-response to UI changes that happen after initial page load. Site
-adapters are responsible for recognizing these changes by observing the
-DOM. So far, we have been able to use event listeners and the
-MutationObserver API to successfully obse
-
-Another challenge is persisting updates to the DOM---some websites use
-virtual DOM frameworks that can occasionally overwrite changes made by
-Wildcard.
-
-So far, we've managed to work around these issues for all the websites
-we've tried, but we don't claim that any website can be customized
-through DOM scraping. As web frontend code gets increasingly complex
-(and starts to move beyond the DOM to other technologies like Shadow DOM
-or even WebGL for rendering), it may become increasingly difficult to
+nature of Wildcard introduces additional challenges. One challenge is
+registering appropriate event handlers to update the table data in
+response to UI changes that happen after initial page load. Another
+challenge is persisting updates to the DOM---some websites use virtual
+DOM frameworks that can occasionally overwrite changes made by Wildcard.
+So far, in practice we've managed to work around these issues for all of
+the websites we've tried, but we don't claim that any website can be
+customized through DOM scraping. As web frontend code gets increasingly
+complex (and starts to move beyond the DOM to other technologies like
+Shadow DOM or even WebGL), it may become increasingly difficult to
 customize websites from the outside without first-party support.
 
-\emph{todo: mention ajax scraping} \emph{todo: mention privacy benefits}
+AJAX scraping proved highly useful in several cases. The Uber Eats
+website was challenging to scrape because it has a complex DOM structure
+with machine-generated CSS classes, but the site also generates AJAX
+requests which contain all the relevant data in a structured form that
+is much easier to extract. We also found examples where relevant
+information wasn't present in the DOM at all. On the grocery delivery
+site Instacart, we used AJAX scraping to augment grocery items with
+their categorization in the store, enabling use cases like viewing all
+the vegetables in an order together.
 
@@
 tradeoff in exchange for a gentler slope in customization, and we show
-in Section~\ref{sec:example} and Section~\ref{sec:evaluation} that our
+in Section~\ref{sec:example} and Section~\ref{sec:reflections} that our
 model can still implement many useful customizations in practice. Also,
@@
 
+\emph{other platforms?}
+
 As computing plays an ever greater role in our lives, it is increasingly

### 2020-05-22 paper.tex
first final

@@
 customize software? We take inspiration from spreadsheets and visual
-database query interfaces \citep{bakke2016, 2020a}, which have
-successfully enabled millions of end users to compute with data through
-direct manipulation.
+database query interfaces, which have successfully enabled millions of
+end users to compute with data through direct manipulation.
 
@@
 
-Despite these limitations, we found that Wildcard does offer
-predictability: once we built a site adapter for a website, it was
-generally obvious what customizations were possible, and the UI guided
-us towards building customizations that matched the system's model. If
-data is not available in the table, for example, then it's abundantly
-clear that customizations relying on that data will not be expressible.
-
-We consider all of these limitations to be an acceptable outcome. Our
-goal is to support as many useful customizations as possible with a low
+We consider these limitations to be an acceptable outcome. Our goal is
+to support as many useful customizations as possible with a low
 threshold of difficulty, and not to span all possible customizations. If
 users want to implement more sophisticated customizations, they have the
-option of graduating to more advanced customization tools.
+option of graduating to more advanced customization tools like scripting
+languages.
+
+We have found that one benefit of showing structured data is
+predictability: once we build an adapter for a website, it is clear what
+data is available or unavailable for use in customizations. Also, there
+is sometimes a way to reframe an imperative script in terms of our
+direct manipulation model. For example, a script that iterates through
+rows in a page adding some additional information to each row can be
+reproduced using a single formula in Wildcard.
 
@@
 
+Here we discuss some themes that we've explored with this work, which go
+beyond our specific approach and address issues in software
+customization more broadly.
+
 \hypertarget{sec:dm}{%
@@
 objects being manipulated are in the domain of programming, not in the
-domain of the task at hand. Imperative programming is a reasonable
-choice as the model for building customizations. Turing-complete
-programming provides a high ceiling for possible customizations, and a
-sequence of commands is a natural fit for automations that simulate a
-series of steps taken by the user.
+domain of the task at hand.
 
@@
 possible customizations, and a sequence of commands is a natural fit for
-automations which simulate a series of steps taken by the user.
-
-There is, however, a serious drawback to this approach. MacLean et al
+automations that simulate a series of steps taken by the user. There is,
+however, a serious drawback to this approach. MacLean et al
 \citep{maclean1990} describe an ideal for user-tailorable systems: a
@@
 environment as it is to use it''---at least for some subset of changes.
-But in scripting languages, even user-friendly ones like Applescript or
-Chickenfoot, the experience of customization does not remotely resemble
-the experience of use, so these systems can't meet this goal.
+But in scripting languages, the experience of customization does not
+remotely resemble the experience of use.
 
@@
 table, e.g.~sorting/filtering by clicking the relevant column header, or
-adding annotation by clicking on the relevant row. These interactions
+adding annotations by clicking on the relevant row. These interactions
 are common in GUI applications, and Wildcard therefore meets MacLean et
 al's goal: some one-click customizations are as easy as using the
-original application.
+original application. Formulas introduce some additional complexity, but
+the pure functional semantics and the intermediate visibility of results
+make the task still easier than imperative programming.
 
 One aspect of directness that we have chosen not to maintain in Wildcard
-is enabling customization in the context of the original user interface,
-as explored by tools like Scotty \citep{eagan2011}. We have found that
-augmenting the original UI with a distinct, additional representation
-provides a consistent experience across all applications, and clearly
-shows the user what structured data is available to work with.
+is enabling customization in closer proximity to the original user
+interface elements, as explored by tools like Scotty \citep{eagan2011}.
+We have found that augmenting the original UI with a distinct,
+additional representation provides a more consistent experience across
+all applications, and clearly shows the user what structured data is
+available to work with.
 
@@
 representations \citep{ainsworth1999}. In their terms, Wildcard plays a
-complementary role by supporting a different set of tasks from the
-original application, while displaying shared information. Wildcard may
-also help construct deeper understanding by subtraction. By stripping
-away details and only showing the essential data in an interface,
-Wildcard helps users think of new ways of using that data, outside the
-specific constraints of the original application.
-
-It is important to note that there are many specific customizations that
-can be achieved in systems like Chickenfoot \citep{bolin2005} that
-cannot be reproduced in Wildcard. We consider this an acceptable
-tradeoff in exchange for a gentler slope in customization, and we show
-in Section~\ref{sec:example} and Section~\ref{sec:reflections} that our
-model can still implement many useful customizations in practice. Also,
-there is sometimes a way to reframe an imperative script in terms of our
-direct manipulation model; for example, a script that iterates through
-rows in a page adding some additional information can be reproduced
-using a formula in Wildcard.
+\emph{complementary role} by supporting a \emph{different set of tasks}
+from the original application, while displaying \emph{shared
+information}. Wildcard may also help construct \emph{deeper
+understanding by subtraction}. By stripping away details and only
+showing the essential data in an interface, Wildcard helps users think
+of new ways of using that data, outside the specific constraints of the
+original application.
+
+As noted in Section~\ref{sec:reflections}, there are many customizations
+that can be achieved in scripting languages that cannot be implemented
+in Wildcard. We consider this an acceptable tradeoff in exchange for a
+gentler slope in customization, and we show in Section~\ref{sec:example}
+and Section~\ref{sec:reflections} that our model can still implement
+many useful customizations.
 
@@
 \subsection{Semantic wrappers}\label{semantic-wrappers}}
-
-Software customization tools typically fall into one of two categories:
-ad hoc, or semantic.
 
@@
 demonstrated the utility of customizing websites through manipulating
-the DOM, without needing explicit extension APIs to be built in.
-However, ad hoc customization comes with a corresponding cost: these
-tools can typically only operate at a low level of abstraction,
-e.g.~manipulating user interface elements. This makes it harder for end
-users to write scripts, and makes the resulting scripts more brittle.
+the DOM, without websites needing to provide explicit extension APIs.
+However, ad hoc customization comes with a cost: these tools typically
+operate at a low level of abstraction, e.g.~manipulating user interface
+elements, rather than in a meaningful domain model. This makes it harder
+for end users to write scripts, and makes the resulting scripts more
+brittle as the specifics of a user interface change.
 
@@
 include accessing a backend web API, or writing a customization in
-Applescript that uses an application-specific API. The main benefit is
-that this allows the extension author to work with meaningful concepts
-in the application domain---``create a new calendar event'' rather than
-``click the button that says new event.''---which makes customizations
-easier to build and more robust. However, this style limits the range of
-extensions that can be built to those that the official plugin API
-supports.
-
-\emph{todo: copy over more content from gdoc}
+Applescript for an application that exposes its domain model to the
+scripting language. The main benefit of this style is that it allows the
+extension author to work with meaningful concepts in the application
+domain---``create a new calendar event'' rather than ``click the button
+that contains the text `new event'.''---which makes customizations
+easier to build and more robust. However, the plugin API limits the
+types of customizations that can be built, and many applications don't
+have an extension system at all.
 
 With Wildcard, we use a hybrid approach that aims to provide the best of
-both worlds. Programmers implement an API wrapper that is internally
-implemented as an ad hoc customization, but externally provides a
+both worlds. Programmers implement site-specific adapters that are
+internally implemented ad hoc customizations, but externally provide a
 high-level interface to the application, abstracting away the details of
@@
 available to all users of the system. When an end user is using a site
-that already has an adapter, they get a semantic customization
-experience that avoids low-level details.
+that already has an adapter implemented, they benefit from a semantic
+customization experience that avoids low-level details.
 
@@
 architecture has been successfully demonstrated by projects like
-Gmail.js, an open source project that creates a convenient API for
-browser extensions to interface with the Gmail web email client.
-
-(\emph{todo: mention automated wrapper induction here or somewhere
-else?})
+\href{https://github.com/KartikTalwar/gmail.js/}{Gmail.js}, an open
+source project that creates a convenient API for browser extensions to
+interface with the Gmail web email client.
+
+The success of semantic wrappers depends on a key hypothesis: that a
+single wrapper created by a programmer can be used for many different
+purposes by end users. Although we've validated that a single generic
+adapter can support many customizations, so far the people making the
+adapters have largely been the same people building customizations on
+top of them. One helpful future addition would be to allow end users to
+create wrappers themselves, or at least to modify existing ones. There
+are many existing projects exploring end-user web scraping (such as
+Helena \citep{chasins2018}) which might prove helpful for this.
+
+The distribution mechanism for semantic wrappers is also important.
+Currently, the distribution mechanism is simply merging the code for all
+adapters into the main Wildcard codebase. This is the simplest workable
+solution, but has its downsides: contributing involves a fairly high
+barrier of creating a pull request on Github, and using newly
+contributed wrappers requires installing a new version of the extension.
+In the future we might explore other mechanisms, like an online
+repository that the extension pulls from dynamically. Security is also a
+consideration---DOM scraping adapters can execute arbitrary Javascript
+code, which opens up the possibility of malicious adapters being
+contributed. Currently we plan to solve this challenge with centralized
+code review, but another approach we are considering is using or
+inventing a more restricted domain-specific language for specifying
+scraping logic.
 
@@
 This paper extends work reported in a workshop paper by Litt and Jackson
-\citep{litt2020} which presented an early prototype version of Wildcard.
-We have substantially extended their work in this paper by creating the
-table adapter abstraction and reimplementing the system around that
-abstraction, evaluating the system more fully on many more websites, and
-by characterizing the design of the system in more detail.
+\citep{litt2020} which presented an earlier version of the Wildcard
+extension. We have substantially extended their work in this paper by
+creating the table adapter abstarction, reimplementing the internals of
+Wildcard around that abstraction, evaluating the system on many more
+websites, and by characterizing the design of the system in much more
+detail than in their workshop paper.
 

### 2020-05-22 paper.tex
first final

@@
 coverage of more websites, but at the expense of complicating the end
-user experience. We might explore such techniques in Wildcard in the
-future, but we think that it's valuable for end users to have a
-customization experience decoupled from the challenge of web scraping
-the underlying data. Sifter implements scraping across multiple pages, a
-valuable feature for sorting and filtering that isn't present in
-Wildcard.
+user experience. We could integrate end user scraping techniques in
+Wildcard in the future, but we believe that when possible it's valuable
+for end users to have a customization experience decoupled from the
+challenge of web scraping the underlying data. Sifter also implements
+scraping across multiple pages, a valuable feature for sorting and
+filtering that isn't present in Wildcard.
 
 Thresher \citep{hogue2005} helps end users create wrappers that map
-website content to Semantic Web schemas, making customizations available
-using the schema information. Wildcard shares the general idea of
-wrappers with Thresher, but maps to a single generic data type, rather
-than more specific schemas, increasing the range of websites and data
-that Wildcard is able to support.
+website content to Semantic Web schemas like ``Movie'' or ``Director,''
+and augments websites with new functionality by exploiting that schema
+information. Wildcard shares the general idea of wrappers, but maps to a
+generic table data type rather than more specific schemas, increasing
+the range of supported data and allowing for a simpler mapping process.
 
@@
 web applications together visually. As mentioned previously, these tools
-all require writing imperative programs in some form, in contrast to the
-more declarative and direct approach of table-driven customization.
+all require writing imperative programs, in contrast to the more
+declarative and direct approach of table-driven customization.
 
@@
 
-Another relevant area involves direct manipulation interfaces for
-interacting with structured data. We take inspiration from these tools
-in our work, but apply them in a different domain: customizing existing
-software applications, rather than interacting with databases or
-constructing software from scratch.
+Another relevant area involves spreadsheets and visual query systems. We
+take inspiration from these tools in our work, but apply them in a
+different domain: customizing existing software applications, rather
+than interacting with databases or constructing software from scratch.
 
 The most closely related work is in systems that offer spreadsheet-like
-querying of relational data, as proposed by Liu and Jagadish
-\citep{liu2009}. SIEUFERD by Bakke and Karger \citep{bakke2016} is an
-example; their paper presents a survey of other similar tools. Our work
-is particularly influenced by the authors' observation that direct
-manipulation requires that the user manipulate the results of a database
-query rather than the query itself, and that the user must see
-intermediate results at every step of constructing the query. SIEUFERD's
-interface supports a far wider range of queries than Wildcard, but the
-basic ideas of the user interface are similar.
-
-\href{https://airtable.com/}{Airtable} is another example of a modern
-commercial product that offers spreadsheet-like interaction with a
-relational database.
+querying of relational data. SIEUFERD by Bakke and Karger
+\citep{bakke2016} is one such recent system, and their paper presents a
+survey of many other similar tools. Our work is particularly influenced
+by the authors' observation that direct manipulation requires that the
+user manipulate the results of a database query rather than the query
+itself, and that the user must see intermediate results at every step of
+constructing the query. SIEUFERD's interface supports a far wider range
+of queries than Wildcard, but the fundamental principles of the user
+interface are similar. \href{https://airtable.com/}{Airtable} is another
+example of a modern commercial product that offers spreadsheet-like
+interaction with a relational database.
 
@@
 \citep{benson2014}, Gneiss \citep{chang2014}, Marmite \citep{wong2007},
-\href{https://www.glideapps.com/}{Glide}. These projects share the idea
-that a spreadsheet is a convenient interface for editing the data
-underlying a GUI application. We share that idea, but apply it to
-software customization, rather than building software from scratch.
-
-Another system in this space is ScrAPIr, by Alrashed et
+\href{https://www.glideapps.com/}{Glide}. We share the core idea of
+connecting a spreadsheet view to a GUI, but apply it to software
+customization, rather than building software from scratch.
+
+Another related system is ScrAPIr, by Alrashed et
 al.~\citep{alrashed2020}, which enables end users to access backend web

### 2020-05-23 paper.tex
tweak
more pictures

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Customizing Software by Direct Manipulation of Tabular Data}
+\title{End-User Customization by Direct Manipulation of Tabular Data}
 
@@
 \begin{abstract}
-  In this paper we show how the behavior of a software application can
-  be extended and adapted by direct manipulation. We introduce
-  \emph{table-driven customization}, a new paradigm where end users
-  directly manipulate a table view of the structured data inside an
-  application, rather than writing imperative scripts (as in most
-  customization tools). This simple model also accommodates a
-  spreadsheet formula language and custom data editing widgets, which
-  provide sufficient expressivity to implement many useful
-  customizations.
+  In this paper we introduce \emph{table-driven customization}, a new
+  way for end users to customize software. We augment a user interface
+  with a table view showing the structured data inside the application.
+  When users edit the table, their changes are reflected in the original
+  UI. This simple model also accommodates a spreadsheet formula language
+  and custom data editing widgets, providing enough power to implement a
+  variety of useful extensions.
 
   We illustrate the approach with Wildcard, a browser extension that
-  implements table-driven customization in the context of existing web
-  applications. Through concrete examples, we show that this paradigm
-  can support useful customizations for real websites. We share
-  reflections from our experiences using the Wildcard system, on both
-  its strengths and limitations relative to other customization
-  approaches. Finally, we explore how this paradigm might lead to new
-  software architectures that encourage this form of customization.
+  implements table-driven customization for existing applications using
+  web scraping. Through concrete examples, we show that this paradigm
+  can support extensions to many real websites, ranging from sorting and
+  filtering existing data to adding entire new features. We share
+  reflections from our experiences using Wildcard, on both its strengths
+  and limitations relative to other customization approaches. Finally,
+  we explore how this paradigm might lead to new software architectures
+  that encourage this form of end-user customization.
 \end{abstract}
@@
 
-Many applications don't meet the precise needs of their users. End user
-customization systems can help improve the situation, by empowering
-non-programmers to modify their software in the way they would like.
-
-Most end user customization systems offer a simplified version of
-programming. Some scripting languages \citep{bolin2005, cook2007} have a
-friendly syntax that resembles natural language. Other visual
-customization tools eliminate text syntax entirely. Macro recorders
-\citep{cook2007, chasins2018, anupam2000} remove some of the initial
-programming burden by letting a user start with concrete demonstrations.
-Despite their many differences, these approaches all share something in
-common: an imperative programming model, with statement sequencing,
-mutable variables and loops.
-
-We have known for decades about an alternate approach: \emph{direct
+Many applications don't meet the precise needs of their users, and it is
+impossible for developers to anticipate everyone's unique requirements.
+End user customization systems can help close this gap, by empowering
+non-programmers to modify their software to satisfy their personal
+goals.
+
+Many end user customization systems
+\citep{cook2007, bolin2005, leshed2008, chasins2018} offer a scripting
+model. They use various strategies to make programming more
+approachable: friendly syntax, a visual programming environment, or
+macro recording to bootstrap from concrete demonstrations. But all these
+techniques build on the same fundamental foundation: an imperative
+programming model, with statement sequencing, mutable variables, and
+loops.
+
+We have known for decades about an alternative: \emph{direct
 manipulation} \citep{shneiderman1983}, where ``visibility of the object
@@
 manipulation is the \emph{de facto} standard in GUIs today, but when it
-comes to customizing those GUIs, it is rarely to be found. As a result,
-switching from using software to customizing it poses a learning barrier
-for users not familiar with programming, and requires an abrupt shift in
-mental model even for those familiar with scripting.
-
-In this work, we ask: what would it look like to build a direct
-maniulation interface that lowers the threshold for starting to
-customize software? We take inspiration from spreadsheets and visual
-database query interfaces, which have successfully enabled millions of
-end users to compute with data through direct manipulation.
-
-In this paper we present a technique called \emph{table-driven
-customization} inspired by those tools. An application's UI is augmented
-with a table view, where the user can see and manipulate the
-application's internal data. Changes in the table view result in
-immediate corresponding changes to the original user interface, enabling
-the user to customize an application with live feedback.
+comes to customizing those GUIs, it is rarely to be found. Switching
+from using an interface to customizing it via scripting requires an
+abrupt shift in interaction model, and can pose a steep learning barrier
+for users not familiar with programming. We subscribe to MacLean et al's
+vision of a ``gentle slope'' \citep{maclean1990}, where users should
+only need to make minimal and incremental investments in skill to
+achieve their desired customizations.
+
+We seek to add a point to this gentle slope by proposing a new method
+for customizing software via direct manipulation, which requires less
+skill than scripting but still supports powerful customizations. We take
+inspiration from visual database query interfaces and spreadsheets,
+which have successfully enabled millions of end users to compute with
+data through direct manipulation.
+
+In this paper, we present a paradigm called \emph{table-driven
+customization}. An application's UI is augmented with a table view where
+the user can see and manipulate the application's internal data, just as
+in a familiar spreadsheet. These changes don't just apply to the table;
+they also result in immediate changes to the application's original user
+interface. The user can sort/filter data in the UI, inject annotations
+into the UI, pull in related information from other web services, and
+more, all using the table as a mediating interface.
 
@@
 user experience of table-driven customization through an example
-scenario using Wildcard.
+scenario, using Wildcard to add features to Hacker News.
 
@@
 We have successfully used Wildcard to build customizations for 11
-different websites. In Section~\ref{sec:reflections}, we present
-reflections from this process, outlining the kinds of customizations we
-were able to build, limitations we encountered, and the challenges of
-integrating scraping logic with websites.
+different websites which serve our own personal needs. In
+Section~\ref{sec:reflections}, we present reflections from this process.
+We outline the kinds of customizations we were able to build,
+limitations we encountered, and some of the challenges of writing
+scraping logic.
 
@@
 To illustrate the end user experience of table-driven customization, we
-consider an example scenario of customizing Hacker News, a tech news
+consider an example scenario of customizing
+\href{https://news.ycombinator.com/}{Hacker News}, a tech news
 aggregator. Figure~\ref{fig:hacker-news} shows accompanying screenshots.
@@
 
-The table contains additional columns to the right of the original
-attributes for each row, where the user can enter spreadsheet-style
-formulas to compute derived values. We provide generic names for these
-columns by default as spreadsheets do, so that users don't need to
-choose column names themselves. The user enters a formula into the
-\texttt{user1} column: (Figure~\ref{fig:hacker-news}, Note C):
+The table contains additional empty columns (\texttt{user1},
+\texttt{user2}, \ldots), where the user can enter spreadsheet-like
+formulas to compute derived values. The user enters a formula into the
+first column: (Figure~\ref{fig:hacker-news}, Note C):
 
@@
 a public web API to compute an estimated read time for the link's
-contents. The argument \texttt{link} refers to a column name in the
-table; for each row, the formula is evaluated with the link for that
-particular row. The result of evaluating this formula is a numerical
-value in seconds for each article.
+contents. The \texttt{link} argument refers to a column name in the
+table; the formula is automatically evaluated across all rows in the
+table, using the value of \texttt{link} for each row.
 
@@
 
+Here we describe the abstract interface fulfilled by a table adapter,
+and then describe specific types of table adapters we've built so far in
+the Wildcard system.
+
+\hypertarget{abstract-interface}{%
+\subsubsection{Abstract interface}\label{abstract-interface}}
+
 The first two parts of the table adapter interface resemble a typical
@@
 associates named attributes with values. Tables have a typed schema, so
-the same attributes are shared across all records. The columns also
-carry some additional metadata in addition to their type, such as
-whether or not they are read-only or editable.
+the same attributes are shared across all records. We currently support
+strings, numeric values, booleans, and datetimes as types. The columns
+also carry some additional metadata, such as whether or not they are
+read-only or editable.
 
@@
 
-\hypertarget{types-of-table-adapters}{%
-\subsubsection{Types of table adapters}\label{types-of-table-adapters}}
-
-We have built three specific types of table adapters so far, to enable
-web customization using Wildcard.
-
-\textbf{DOM scraping adapters} are the essential component that enables
-Wildcard to interface with an existing website UI. In addition to the
-standard web scraping problem of extracting a table of data from the
-DOM, a scraping adapter must also manipulate the DOM to reorder rows,
-edit form entries, and inject annotations as the table is edited.
+\hypertarget{dom-scraping-adapters}{%
+\subsubsection{DOM scraping adapters}\label{dom-scraping-adapters}}
+
+DOM scraping adapters are the essential component that enables Wildcard
+to interface with an existing website UI. In addition to the standard
+web scraping problem of extracting a table of data from the DOM, a
+scraping adapter must also manipulate the DOM to reorder rows, edit form
+entries, and inject annotations as the table is edited.
 
@@
 
-After specifying basic metadata like which URLs the adapter should apply
-to, the programmer only needs to implement a single Javascript function
-that scrapes relevant elements from the page. They are free to use any
-scraping techniques; in practice simple CSS selectors and DOM APIs
-usually suffice. The generic adapter then wraps this scraping function
-to implement the table adapter interface. For example, when the table
-view is sorted, the generic adapter takes the DOM elements corresponding
-to the rows in the table, removes all of them from the page, and then
-reinserts them in the new order.
-
-The programmer can optionally override other default behavior of the
-generic DOM scraping adapter, including the logic for when to re-run the
-scraping function in response to page changes, how to style injected
-annotations, and how to style the currently selected row in the table.
-
-An \textbf{AJAX scraping adapter} intercepts AJAX requests made by a web
-page, and extracts information from those requests to add to the table.
-When available, this tends to be a helpful technique because the data is
+\emph{Specifying scraping logic}: First, the adapter author specifies
+basic metadata like which URLs the adapter should apply to, and a list
+of columns and their types. Then, implement a single Javascript function
+that extracts relevant elements from the page using web scraping
+technqiues. The generic adapter then wraps this scraping function to
+implement the table adapter interface. For example, when the user sorts
+the table view, the generic adapter removes the DOM elements
+corresponding to the rows in the table, and reinserts them back into the
+page in the correct order.
+
+For each scraped row of data, the programmer must return a unique
+identifier, and a value for each column in the table.
+
+\emph{Choosing an ID}: When possible, we recommend choosing a
+server-side identifier that will remain stable across pageloads. This
+enables user annotations persisted in local storage to be associated
+with the same records on subsequent pageloads. We have found that it's
+usually possible to find such an identifier; for example, each item in a
+page often contains a link to a page with more details, where the URL
+contains a unique identifier.
+
+\emph{Types of scraped values}: For each individual value within a row,
+there are two options for what type of data can be returned by the
+programmer-specified scraping function.
+
+The default option is to return a DOM element, in which case the generic
+adapter extracts the text contents of the DOM element and casts them to
+the type of the column. For example, if the relevant column is a numeric
+column and the returned value is a \texttt{\textless{}div\textgreater{}}
+containing the string ``1.23'', the generic adapter extracts the number
+1.23. The advantage of this approach is that the field becomes
+\emph{writable} in the table, if the adapter author chooses to mark it
+as writable. When the user edits the value in the table, the generic
+adapter can overwrite the inner contents of the DOM element to reflect
+the change. In practice, this is most useful for editing the value of
+form input fields. It's possible to enable the editing of other fields
+(e.g., editing the title of a post on Hacker News) but we haven't found
+many use cases for editing those types of fields.
+
+Another option is to directly return a value, rather than returning a
+DOM element. The advantage of this approach is that the adapter author
+can perform arbitrary computations to derive the returned value---for
+example, they can use a regular expression to extract a substring. The
+disadvantage is that the field is no longer writable, because the
+computation used to derive the value isn't reversible.
+
+\emph{Other site-specific customization}: The programmer can optionally
+override other behavior of the generic DOM scraping adapter, including
+logic for when to re-run the scraping function in response to page
+changes, how to style injected annotations, and how to style the
+currently selected row in the table. We've chosen sensible defaults that
+generally result in a workable adapter without needing to change these
+options, but they can improve the experience on a specific website. For
+example, by default we highlight the currently selected row with a blue
+border, but on some sites this is hard to see or breaks the layout, so
+setting a background color works better.
+
+\hypertarget{ajax-scraping-adapters}{%
+\subsubsection{AJAX scraping adapters}\label{ajax-scraping-adapters}}
+
+An AJAX scraping adapter intercepts AJAX requests made by a web page,
+and extracts information from those requests to add to the table. When
+available, this tends to be a helpful technique because the data is
 already in a structured form so it is easier to scrape, and it often
@@
 
-The \textbf{local storage adapter} simply stores a table of data in the
-browser. This is currently only used to persist annotations: the user
-makes edits in the table, they are stored privately in the browser, and
-can be loaded into the table on future pageloads.
+\hypertarget{local-storage-adapters}{%
+\subsubsection{Local storage adapters}\label{local-storage-adapters}}
+
+The local storage adapter simply stores a table of data in the browser.
+This is currently only used to persist annotations.
+
+The table view is initialized with empty columns such as \texttt{user1}
+which serve as the user's ``scratch space,'' as shown in
+Section~\ref{sec:example}. When the user makes edits to these columns,
+new rows are created in the local storage table. The rows contain the
+record ID from the DOM scraping adapter, which enables them to be
+re-associated with the same records on subsequent pageloads.
 
@@
 Airbnb           & Travel               & 73                                       & Add Walkability Scores to listings. Sort listings by price.                           \\
-Amazon           & Online shopping      & 99                                       & Sort used book sellers by total price, including delivery fees                                \\
-Blogger          & Blogging             & 36                                       & Use alternate text editor to edit blog posts                                                 \\
-Expedia          & Travel               & 41                                       & Use alternate datepicker to enter travel dates                                               \\
-Flux             & Data portal          & 67                                       & Use Wildcard as a faster table editor for editing lab results                                \\
-Github           & Code repository      & 62                                       & Sort a user's code repositories by stars to find popular work                                \\
+Amazon           & Online shopping      & 99                                       & Sort third party sellers by total price, including fees.                                \\
+Blogger          & Blogging             & 36                                       & Use alternate text editor to edit blog posts.                                                 \\
+Expedia          & Travel               & 41                                       & Use alternate datepicker to enter travel dates.                                               \\
+Flux             & Data portal          & 67                                       & Use Wildcard as a faster table editor for editing lab results.                                \\
+Github           & Code repository      & 62                                       & Sort a user's code repositories by stars to find popular work.                                \\
 Hacker News      & News                 & 69                                       & Add read times to links. Filter out links that have been read. \\
@@
 We built Wildcard integrations for 11 websites, including transactional
-sites like Amazon and Uber Eats, and media consumption sites like Hacker
-News and Youtube. Table 1 summarizes these integrations, showing the
-number of lines of code in the adapter configuration for each site, and
-some example customizations we were able to perform for each site.
+sites like Amazon and Uber Eats, and media consumption sites like
+Youtube and Hacker News. Table 1 summarizes these integrations, showing
+the number of lines of code in the adapter configuration for each site,
+and some example customizations we were able to perform for each site.
 
@@
 
+\begin{figure}
+\hypertarget{fig:amazon}{%
+\centering
+\includegraphics[width=\columnwidth]{media/amazon.png}
+\caption{Sorting the used sellers page on Amazon by total price}\label{fig:amazon}
+}
+\end{figure}
+
+\begin{figure}
+\hypertarget{fig:ubereats}{%
+\centering
+\includegraphics[width=\columnwidth]{media/ubereats.png}
+\caption{Organizing takeout restaurants on Uber Eats by ETA and price}\label{fig:ubereats}
+}
+\end{figure}
+
 \hypertarget{annotating}{%
@@
 rent). We have also used it with Instacart's online grocery cart, for
-jotting down notes like ``should we get more milk?''
+jotting down notes like ``do we need more eggs?''
 
@@
 used them to augment Airbnb listings with walkability scores, and to
-augment Hacker News articles with estimated read times. One challenge of
-the current language design is that supporting a new web API requires
-writing Javascript code to add a new function to the language, because
-web APIs typically return complex JSON data structures that can't be
-easily displayed in a single table cell. In the future we would like to
-make it possible to call new APIs without adding a dedicated function,
-which might require adding functions to the formula language that can
-manipulate JSON data.
+augment Hacker News articles with estimated read times as shown in
+Section~\ref{sec:example}. One challenge of the current language design
+is that supporting a new web API requires writing Javascript code to add
+a new function to the language, because web APIs typically return
+complex JSON data structures that can't be easily displayed in a single
+table cell. In the future we would like to make it possible to call new
+APIs without adding a dedicated function, which might require adding
+functions to the formula language that can manipulate JSON data.
 
@@
 string operations, as shown in Section~\ref{sec:example}.
+
+\begin{figure}
+\hypertarget{fig:instacart}{%
+\centering
+\includegraphics[width=\columnwidth]{media/instacart.png}
+\caption{Taking notes on Instacart grocery items, after sorting them by price}\label{fig:instacart}
+}
+\end{figure}
 
@@
 website was challenging to scrape because it has a complex DOM structure
-with machine-generated CSS classes, but the site also generates AJAX
-requests which contain all the relevant data in a structured form that
-is much easier to extract. We also found examples where relevant
-information wasn't present in the DOM at all. On the grocery delivery
-site Instacart, we used AJAX scraping to augment grocery items with
-their categorization in the store, enabling use cases like viewing all
-the vegetables in an order together.
+with machine-generated CSS classes, but the site also use AJAX requests
+which contain all the relevant data in a structured form that is much
+easier to extract. We also found examples where relevant information
+wasn't present in the DOM at all. On the grocery delivery site
+Instacart, we found that AJAX requests contained additional data not
+shown in the UI, enabling us to do things like sort grocery items in an
+order by category.
 
@@
 
-\emph{other platforms?}
-
 As computing plays an ever greater role in our lives, it is increasingly

### 2020-05-23 paper.tex
first cut of final draft
possible final

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{End-User Customization by Direct Manipulation of Tabular Data}
+\title{End-User Software Customization by Direct Manipulation of Tabular
+Data}
 
@@
   In this paper we introduce \emph{table-driven customization}, a new
-  way for end users to customize software. We augment a user interface
-  with a table view showing the structured data inside the application.
-  When users edit the table, their changes are reflected in the original
-  UI. This simple model also accommodates a spreadsheet formula language
-  and custom data editing widgets, providing enough power to implement a
+  way for end users to extend software by direct manipulation without
+  doing traditional programming. We augment a user interface with a
+  table view showing the structured data inside the application. When
+  users edit the table, their changes are reflected in the original UI.
+  This simple model accommodates a spreadsheet formula language and
+  custom data editing widgets, providing enough power to implement a
   variety of useful extensions.
@@
 techniques build on the same fundamental foundation: an imperative
-programming model, with statement sequencing, mutable variables, and
-loops.
+programming model, with statements, mutable variables, and loops.
 
@@
 manipulation} \citep{shneiderman1983}, where ``visibility of the object
-of interest'' replaces ``complex command language syntax''. Direct
+of interest'' replaces ``complex command language syntax.'' Direct
 manipulation is the \emph{de facto} standard in GUIs today, but when it
 comes to customizing those GUIs, it is rarely to be found. Switching
-from using an interface to customizing it via scripting requires an
+from using an application to customizing it via scripting requires an
 abrupt shift in interaction model, and can pose a steep learning barrier
-for users not familiar with programming. We subscribe to MacLean et al's
-vision of a ``gentle slope'' \citep{maclean1990}, where users should
+for users not familiar with programming.
+
+We subscribe to MacLean et al's vision of a ``gentle slope''
+\citep{maclean1990} free of such abrupt ``cliffs,'' where users should
 only need to make minimal and incremental investments in skill to
-achieve their desired customizations.
-
-We seek to add a point to this gentle slope by proposing a new method
-for customizing software via direct manipulation, which requires less
-skill than scripting but still supports powerful customizations. We take
-inspiration from visual database query interfaces and spreadsheets,
-which have successfully enabled millions of end users to compute with
-data through direct manipulation.
-
-In this paper, we present a paradigm called \emph{table-driven
-customization}. An application's UI is augmented with a table view where
-the user can see and manipulate the application's internal data, just as
-in a familiar spreadsheet. These changes don't just apply to the table;
-they also result in immediate changes to the application's original user
-interface. The user can sort/filter data in the UI, inject annotations
-into the UI, pull in related information from other web services, and
-more, all using the table as a mediating interface.
+achieve their desired customizations. We seek to contribute to this
+gentle slope with a new method for customizing software via direct
+manipulation, which requires less skill than scripting but still
+supports powerful customizations. We take inspiration from visual
+database query interfaces and spreadsheets, which have successfully
+enabled millions of end users to compute with data through direct
+manipulation.
+
+In our proposed paradigm, \emph{table-driven customization}, an
+application's UI is augmented with a table view where the user can see
+and manipulate the application's internal data. These changes don't just
+apply to the table; they also result in immediate changes to the
+application's original user interface. The user can sort/filter data in
+the UI, inject annotations, pull in related information from other web
+services, and more, all using the table as a mediating interface.
+Interacting with the table view feels like interacting with a familiar
+spreadsheet, but results in customizing an existing application.
 
@@
 We have successfully used Wildcard to build customizations for 11
-different websites which serve our own personal needs. In
+different websites that serve our own personal needs. In
 Section~\ref{sec:reflections}, we present reflections from this process.
@@
 \item
-  \emph{Customization by direct manipulation}: We agree with MacLean et
-  al \citep{maclean1990} that customization should follow a ``gentle
-  slope'', where users can gradually invest in more complex techniques
-  to achieve deeper customizations. We believe an important point on
-  this slope is the ability to customize an application by directly
-  seeing and changing its internal structured data, rather than by
-  writing imperative scripts.
+  \emph{Customization by direct manipulation}: We discuss further how
+  table-driven customization relates to a gentle slope of customization.
+  We suggest that an important point on this slope is the ability to
+  customize an application by directly seeing and changing its data,
+  rather than by writing imperative scripts.
 \item
-  \emph{Wrapping applications for customization}: Typically, tools that
-  don't rely on official extension APIs resort to offering low-level
-  APIs for customization. Instead, we propose a community-maintained
-  library of semantic wrappers around existing applications, enabling
-  end users to work with domain objects rather than low-level
-  representations.
+  \emph{Semantic wrappers}: Typically, tools that don't rely on official
+  extension APIs resort to offering low-level APIs for customization.
+  Instead, we propose a community-maintained library of semantic
+  wrappers around existing applications, enabling end users to work with
+  domain objects rather than low-level representations.
 \end{itemize}
 
-Table-driven customization relates to existing work in many areas. In
-particular, our goals overlap with many software customization tools,
-and our methods overlap with direct manipulation interfaces for working
-with structured data, including visual database query systems and
-spreadsheets. We explore these connections and more in
-Section~\ref{sec:related-work}.
+Table-driven customization relates to existing work in many areas. Our
+goals overlap with many software customization tools, and our methods
+overlap with direct manipulation interfaces for working with structured
+data, including visual database query systems and spreadsheets. We
+explore these connections in Section~\ref{sec:related-work}.
+
+Finanlly, in Section~\ref{sec:conclusion}, we conclude and describe some
+opportunities for future work.
 
@@
 consider an example scenario of customizing
-\href{https://news.ycombinator.com/}{Hacker News}, a tech news
+\href{https://news.ycombinator.com/}{Hacker News}, a popular tech news
 aggregator. Figure~\ref{fig:hacker-news} shows accompanying screenshots.
@@
 website UI also becomes sorted in the same order
-(Figure~\ref{fig:hacker-news}, Note B)---Wildcard has manipulated the
-website's DOM to synchronize it with the sort order of the table. This
-sort order is also persisted in the browser and reapplied automatically
-the next time the user loads the page, so they can always browse the
-latest set of links using this sort order.
+(Figure~\ref{fig:hacker-news}, Note B). Internally, Wildcard has changed
+the website's DOM to synchronize it with the sort order of the table.
+This sort predicate is also persisted in the browser and reapplied
+automatically the next time the user loads the page, so they can always
+browse the page sorted by points.
 
 \textbf{Adding estimated read times}: Next, the user decides to attempt
-a more substantial customization: to add estimated read times to
-articles on Hacker News.
-
-The table contains additional empty columns (\texttt{user1},
-\texttt{user2}, \ldots), where the user can enter spreadsheet-like
-formulas to compute derived values. The user enters a formula into the
-first column: (Figure~\ref{fig:hacker-news}, Note C):
+a more substantial customization: to add estimated read times to each
+article, in order to prioritize reading deeper content.
+
+The table contains additional empty columns where the user can enter
+spreadsheet-like formulas to compute derived values. The user enters a
+formula into the first column, which is named \texttt{user1} by default
+(Figure~\ref{fig:hacker-news}, Note C):
 
@@
 
-This calls a built-in function that takes a URL as an argument and uses
-a public web API to compute an estimated read time for the link's
-contents. The \texttt{link} argument refers to a column name in the
-table; the formula is automatically evaluated across all rows in the
-table, using the value of \texttt{link} for each row.
+This formula calls a built-in function that takes a URL as an argument
+and uses a public web API to compute an estimated read time for the
+URL's contents. The \texttt{link} argument in the formula refers to a
+column name in the table; the formula is automatically evaluated across
+all rows in the table, using the value of \texttt{link} for each row.
 
 The user clicks the \texttt{user1} column header to sort the articles on
-the page in descending order of estimated read time, helping to
-prioritize reading deeper content. It would be useful to show these read
-times in the page too, but before doing that, the user wants to make the
-time estimates more readable. They enter another formula in the next
-column, \texttt{user2}:
+the page in descending order of estimated read time. They would also
+like to display the read times in the page, but a number in seconds
+isn't the most legible format, so they enter another formula in the
+\texttt{user2} column:
 
@@
 This formula converts seconds to minutes by dividing by 60 and rounding
-to the nearest integer, and then concatenates a label to the number,
-producing strings like ``21 min read''.
+to the nearest integer, and concatenates the result with a string label,
+producing results like ``21 min read''.
 
@@
 (Figure~\ref{fig:hacker-news}, Note D). Each article on the page now
-shows an annotation with the estimated read time in minutes. (The format
-of how annotations appear for each item was determined by the programmer
-who created the Wildcard adapter for Hacker News.)
-
-\textbf{Adding notes to links}: The user can also manually add notes to
-the table, by simply entering values into the table without using
+shows an annotation with the estimated read time in minutes. (The
+formatting of annotations was determined by the programmer who created
+the adapter for Hacker News.)
+
+\textbf{Adding manual annotations}: The user can also manually add notes
+to the table, by simply entering values into the table without using
 formulas. In this case, the user jots down a few notes in another column
@@
 They can then filter the table to only contain rows where this formula
-column contains \texttt{false}; visited rows are hidden both from the
-table view and the original page. This is an example of a customization
-that the original website could not implement, since websites don't have
-access to the browser history for privacy reasons. But by using
-Wildcard, the user can implement the customization locally, without
-needing to expose their entire browser history to Hacker News.
+column contains \texttt{false}; links that the user has already visited
+are hidden both from the table view and the original page. This is an
+example of a customization that the original website could not have
+implemented, since websites don't have access to the browser history for
+privacy reasons. But by using Wildcard, the user could implement the
+customization locally, without needing to expose their browser history
+to Hacker News.
 
@@
 table-driven customization, using a simplified illustration of the
-Hacker News example scenario. In this example, the name and points value
-for each article is scraped from the web page DOM, and user annotations
-are loaded from the brower's local storage.
+Hacker News example scenario. The name and points value for each article
+is scraped from the web page DOM, and user annotations are loaded from
+the brower's local storage.
 
 First, the web page and the browser storage are each wrapped by a
-\textbf{table adapter}, which defines a bidirectional mapping between
-some underlying data source and a table.
-
-In addition to a \emph{read} mapping for how the underlying data should
-be represented as a table, it also defines a \emph{write} mapping for
-what effects edits should have on the original data source. In this
-case, the local storage adapter has a trivial mapping: it loads a table
-of data stored in the browser, and persists edits to that persisted
-state. The mapping logic of the DOM scraping adapter is much more
-involved. It implements web scraping logic to produce a table of data
-from the web page, and turns edits into DOM manipulations, such as
-reordering rows of data on the page.
+\textbf{table adapter}, which defines a bidirectional mapping between an
+underlying data source and a table.
+
+In addition to a \emph{read mapping} for how the underlying data should
+be represented as a table, it also has a \emph{write mapping} defining
+the effects that edits have on the original data source. The local
+storage adapter has a trivial mapping: it loads a table of data stored
+in the browser, and persists edits to that state. The mapping logic of
+the DOM scraping adapter is much more involved. It implements web
+scraping logic to produce a table of data from the web page, and turns
+edits into DOM manipulations, such as reordering rows of data on the
+page.
 
@@
 
-The first two parts of the table adapter interface resemble a typical
-database table:
-
 \emph{Returning a table}: A table adapter exposes a table of data: an
@@
 adapter to make an edit to a record. The meaning of making an edit can
-vary depending on the adapter: in the local storage adapter, an
-annotation may be persisted into local storage; in the DOM scraping
-adapter, an edit may represent filling in a form field.
+vary depending on the adapter: in the local storage adapter, a new value
+may be persisted into local storage; in the DOM scraping adapter, an
+edit may result in changing the value of a form field.
 
@@
 adapter with the entire combined table shown to the user. The DOM
-scraping adapter uses this for injecting annotations---values in
-additional columns from other tables are added to the original web page.
-
-\emph{Currently selected record}: As the user clicks in different parts
-of the table view, the query engine communicates the record currently
-selected by the user. The DOM scraping adapter uses this information to
-highlight the row in the page that corresponds to the selected row in
-the table (and scroll to the row if it's not currently showing on the
-page), which helps clarify the connection between the table and the
-original UI.
+scraping adapter uses this for injecting annotations---values from other
+tables are added to the original web page.
+
+\emph{Currently selected record}: As the user clicks around the table
+view, the query engine broadcasts the record currently selected by the
+user to each table adapter. The DOM scraping adapter uses this
+information to highlight the row in the page that corresponds to the
+selected row in the table, which helps the user understand the mapping
+between the table and the original UI.
 
@@
 
-Because each website has unique content, we rely on programmers to
+Because every website has unique content, we rely on programmers to
 create a DOM scraping adapter for each individual website to make it
@@
 
-\emph{Specifying scraping logic}: First, the adapter author specifies
-basic metadata like which URLs the adapter should apply to, and a list
-of columns and their types. Then, implement a single Javascript function
-that extracts relevant elements from the page using web scraping
-technqiues. The generic adapter then wraps this scraping function to
-implement the table adapter interface. For example, when the user sorts
-the table view, the generic adapter removes the DOM elements
-corresponding to the rows in the table, and reinserts them back into the
-page in the correct order.
-
-For each scraped row of data, the programmer must return a unique
-identifier, and a value for each column in the table.
-
-\emph{Choosing an ID}: When possible, we recommend choosing a
-server-side identifier that will remain stable across pageloads. This
-enables user annotations persisted in local storage to be associated
-with the same records on subsequent pageloads. We have found that it's
-usually possible to find such an identifier; for example, each item in a
-page often contains a link to a page with more details, where the URL
-contains a unique identifier.
+\emph{Basic metadata}: The programmer specifies URL patterns that the
+adapter should activate on, and a set of column names and types to
+return in the table.
+
+\emph{Specifying scraping logic}: The main component of a site-specific
+adapter is a single Javascript function that extracts relevant elements
+from the page using web scraping technqiues. For each scraped row of
+data, the programmer must return a unique identifier, and values for
+each column in the table.
+
+The generic adapter then wraps this scraping function to implement the
+table adapter interface. For example, when the user sorts the table
+view, the generic adapter synchronizes the sort order to the DOM by
+removing the DOM elements corresponding to the rows in the table and
+reinserting them back into the page in the new order.
+
+\emph{Choosing an ID}: When possible, it is best to choose a server-side
+identifier that remains stable across pageloads. This enables user
+annotations persisted in local storage to be associated with the same
+records on subsequent pageloads. We have found that it's usually
+possible to find such an identifier; for example, each item in a page
+often contains a link to a page with more details, with a URL that
+contains a stable ID.
 
@@
 adapter extracts the text contents of the DOM element and casts them to
-the type of the column. For example, if the relevant column is a numeric
-column and the returned value is a \texttt{\textless{}div\textgreater{}}
-containing the string ``1.23'', the generic adapter extracts the number
-1.23. The advantage of this approach is that the field becomes
-\emph{writable} in the table, if the adapter author chooses to mark it
-as writable. When the user edits the value in the table, the generic
-adapter can overwrite the inner contents of the DOM element to reflect
-the change. In practice, this is most useful for editing the value of
-form input fields. It's possible to enable the editing of other fields
-(e.g., editing the title of a post on Hacker News) but we haven't found
-many use cases for editing those types of fields.
+the type of the column. For example, if the scraping function returns a
+\texttt{\textless{}div\textgreater{}} containing the string ``1.23'',
+and the column has a numeric type, the generic adapter extracts the
+number 1.23. The advantage of this approach is that the field can be
+\emph{writable}---when the user edits the value in the table, the
+generic adapter can simply overwrite the inner contents of the DOM
+element to reflect the change. In practice, this is most useful for
+editing the value of form input fields, although it's technically
+possible to enable the editing of other fields (e.g., editing the title
+of a post on Hacker News).
 
@@
 example, they can use a regular expression to extract a substring. The
-disadvantage is that the field is no longer writable, because the
-computation used to derive the value isn't reversible.
+disadvantage is that the field is no longer writable. The computation
+used to derive the value isn't reversible, so there's no way to reflect
+a table edit in the DOM.
 
@@
   Firefox has convenient APIs for intercepting requests. It appears
-  possible to implement in Chrome as well, but we have not finished our
-  implementation.}
+  possible to implement in Chrome and Edge as well, but we have not
+  finished our implementation.}
 
@@
 state to Wildcard with only minimal effort from the application
-developers. To test this hypothesis, we have created an early prototype
-of a Wildcard adapter for the Redux state management library, and used
-it with the TodoMVC todo list application. To install the adapter, the
-programmer only needs to make a few small changes to their app:
+developers. To test this hypothesis, we have created an early
+proof-of-concept of an integrated website adapter for the popular
+\href{https://redux.js.org/}{Redux} Javascript state management library.
+To set up this integrated adapter, the programmer would only need to
+make a few minor changes to their application code:
 
@@
   the state of a user interface as a single object. To expose a table of
-  data, the developer simply writes a function that maps the state
-  object to a table.
+  data to Wildcard, the developer simply writes a function that maps the
+  state object to a table.
 \item
-  \emph{Editing data values}: The developer defines how edits to records
-  in the table should map to edits in the state object.
+  \emph{Editing data values}: The developer writes handlers which define
+  how edits to table records should map to UI actions. For example,
+  editing the value of a todo might trigger an \texttt{EDIT\_TODO}
+  event.
 \item
@@
   \emph{Annotations}: The developer adds additional React components
-  managed by Wildcard into their UI tree, which render additional data
-  as annotations.
+  managed by the Wildcard adapter into their UI tree, which render
+  additional data as annotations.
 \end{itemize}
@@
 model. Despite its simplicity, this model has proven sufficient for
-meeting the needs of customization in practice, and minimizes the
-complexity of supporting arbitrary queries. But because it fits into the
-general paradigm of relational queries, it could theoretically be
-extended to support a wider range of queries.
+meeting our customization needs, and minimizes the complexity of
+supporting arbitrary queries. But because it fits into the general
+paradigm of relational queries, it could theoretically be extended to
+support a wider range of queries.
 
@@
 formula across an entire column as in spreadsheets, and has worked for
-all of the customizations we have built.
+all of the customizations we have built. Visual database query systems
+like SIEUFERD \citep{bakke2016} use a similar approach to using formulas
+in a structured table.
 
@@
 straightforward to develop additional table editor interfaces on top of
-the Wildcard system. One intruiging future possibility is to apply the
-idea of cell editors to the entire table, by developing table editors
-which don't look like a table. For example, we could provide a calendar
-view for displaying a table containing a date column.
+the Wildcard system. For example, we could provide a calendar view for
+displaying a table containing a date column.
 
@@
 
-We built Wildcard integrations for 11 websites, including transactional
-sites like Amazon and Uber Eats, and media consumption sites like
-Youtube and Hacker News. Table 1 summarizes these integrations, showing
-the number of lines of code in the adapter configuration for each site,
-and some example customizations we were able to perform for each site.
-
-So far, most usage of Wildcard has come from members of the project
-team. Here we offer our reflections on using the system, focused on two
-key questions:
+We developed site-specific adapters for 11 websites that we personally
+use frequently, and then built customizations in Wildcard for those
+websites. Table 1 summarizes these results, showing the number of lines
+of code in the adapter for each site, and some example customizations we
+created. Here we offer our reflections from these experiences using the
+system, focused on two key questions:
 
@@
 Sometimes, websites have opaque ranking algorithms which presumably
-maximize profit but don't offer much control to the user. For example,
-Airbnb previously allowed users to sort listings by price, but removed
-that feature in 2012. We have found other sites where a lack of sorting
-options seems more like merely an omission; for example, the Instacart
-grocery delivery service has a spartan UI for viewing a pending order,
-which doesn't allow for sorting by price or category. In both of these
-cases, Wildcard enables users to take back some control.
+maximize profit but restrict user agency. For example, Airbnb previously
+allowed users to sort listings by price, but removed that feature in
+2012. In other cases, a lack of sorting options seems more like an
+innocent omission; for example, the Instacart grocery delivery service
+has a spartan UI for viewing an order, which doesn't allow for sorting
+items by price or category. In both of these cases, Wildcard enables
+users to take back some control.
 
@@
 filter entries that are shown on the current page, which means that
-users are not entirely liberated from the suggestions of the opaque
-algorithm. This restriction could be overcome in the future by scraping
-content across multiple pages, or by using an integrated adapter built
-in to the application. However, we've also found that sorting/filtering
-a single page of a paginated list is often an acceptable outcome (and
-sometimes even a preferable one). It's more useful, for example, to sort
-30 recommended Youtube videos than to try to sort all videos on Youtube.
+users are not entirely liberated from the site's original ranking. This
+restriction could be overcome in the future by scraping content across
+multiple pages, or by using an integrated adapter and avoiding scraping
+altogether. However, we've also realized that sorting/filtering a single
+page of a paginated list is sometimes an acceptable outcome (and even a
+preferable one). It's more useful, for example, to sort 30 recommended
+Youtube videos than to try to sort all videos on Youtube.
 
@@
 \includegraphics[width=\columnwidth]{media/amazon.png}
-\caption{Sorting the used sellers page on Amazon by total price}\label{fig:amazon}
+\caption{Sorting the used sellers page on Amazon by total price, including fees. The original page doesn't have sorting, and doesn't show the combined price.}\label{fig:amazon}
 }
@@
 \includegraphics[width=\columnwidth]{media/ubereats.png}
-\caption{Organizing takeout restaurants on Uber Eats by ETA and price}\label{fig:ubereats}
+\caption{Organizing takeout restaurants on Uber Eats by delivery ETA and price}\label{fig:ubereats}
 }
@@
 rent). We have also used it with Instacart's online grocery cart, for
-jotting down notes like ``do we need more eggs?''
+jotting down notes as we review an order and consider modifications
+(shown in Figure~\ref{fig:instacart}).
 
@@
 
-We developed two cell editors to explore different benefits of adding
-custom UI for editing the table.
-
-An important benefit of cell editors is that they allow users to
-incorporate their private information into a web UI without uploading it
-to a website. To explore this idea, we created a datepicker based on the
+We developed two cell editors, to explore the benefits of adding custom
+UI for editing values in the table.
+
+One valuable feature is that users can incorporate their private
+information into a web UI. We created a datepicker based on the
 \href{https://fullcalendar.io/}{FullCalendar} plugin, which can load
 data from a Google Calendar. This makes it convenient to enter dates
-into a website based on the user's personal calendar information.
-
-Another benefit of cell editors is that a user can choose their
-preferred widget for editing some type of information. We built a text
-editor based on the \href{https://ckeditor.com/}{CKEditor} rich text
-editor, and demonstrated its use by integrating it with Google's Blogger
-website and using it to edit blog post content.
+into a website based on the user's personal calendar information,
+without uploading that information to the website itself.
+
+Another benefit is that a user can choose a preferred widget for editing
+information across different sites that typically provide a fragmented
+editing experience. We built a text editor based on the
+\href{https://ckeditor.com/}{CKEditor} rich text editor, and integrated
+it with Google's Blogger website by representing the contents of a blog
+post as a single table cell containing an HTML string (shown in
+Figure~\ref{fig:blogger}). If there were site adapters available for
+other websites with rich text entry, the user could then use a single
+rich text editor to edit text across the whole web.
+
+\begin{figure}
+\hypertarget{fig:blogger}{%
+\centering
+\includegraphics[width=\columnwidth]{media/blogger.png}
+\caption{Using a custom text editor widget to edit a blog post on Blogger. The text is synchronized with the Blogger editor through a table cell.}\label{fig:blogger}
+}
+\end{figure}
 
@@
 website was challenging to scrape because it has a complex DOM structure
-with machine-generated CSS classes, but the site also use AJAX requests
+with machine-generated CSS classes, but the site also uses AJAX requests
 which contain all the relevant data in a structured form that is much
@@
 Instacart, we found that AJAX requests contained additional data not
-shown in the UI, enabling us to do things like sort grocery items in an
-order by category.
+shown in the UI, enabling us to do things like sort grocery items by
+category.
 
@@
 Here we discuss some themes that we've explored with this work, which go
-beyond our specific approach and address issues in software
-customization more broadly.
+beyond our specific approach and address broader issues in software
+customization.
 
@@
 table, e.g.~sorting/filtering by clicking the relevant column header, or
-adding annotations by clicking on the relevant row. These interactions
-are common in GUI applications, and Wildcard therefore meets MacLean et
-al's goal: some one-click customizations are as easy as using the
-original application. Formulas introduce some additional complexity, but
-the pure functional semantics and the intermediate visibility of results
-make the task still easier than imperative programming.
-
-One aspect of directness that we have chosen not to maintain in Wildcard
+adding annotations by clicking and typing on the relevant row. At every
+step, the user receives intermediate feedback, not only in the table
+view, but also in the original application, so it's clear whether they
+are making progress towards their desired result. These types of
+interactions are common in GUI applications, and Wildcard therefore
+seems to meet MacLean et al's goal: some one-click customizations are as
+easy as using the original application. Formulas introduce some
+additional complexity, but spreadsheets have demonstrated that formula
+programming is still accessible to many users, due to the pure
+functional semantics and the intermediate visibility of results.
+
+One aspect of directness that we have chosen not to pursue in Wildcard
 is enabling customization in closer proximity to the original user
@@
 information}. Wildcard may also help construct \emph{deeper
-understanding by subtraction}. By stripping away details and only
-showing the essential data in an interface, Wildcard helps users think
-of new ways of using that data, outside the specific constraints of the
-original application.
+understanding by subtraction}---by stripping away details and only
+showing the essential data in an interface, Wildcard encourages thinking
+of an application in terms of its core information, rather than the
+specific capabilities provided by the current user interface. In our
+anecdotal experience, we've often found that looking at a site's data in
+table format tends to spur new ideas for customizations which weren't
+evident from looking at the original UI.
 
@@
 gentler slope in customization, and we show in Section~\ref{sec:example}
-and Section~\ref{sec:reflections} that our model can still implement
-many useful customizations.
+and Section~\ref{sec:reflections} that our model can implement many
+useful customizations.
 
@@
 types of customizations that can be built, and many applications don't
-have an extension system at all.
+have any plugin API.
 
 With Wildcard, we use a hybrid approach that aims to provide the best of
-both worlds. Programmers implement site-specific adapters that are
-internally implemented ad hoc customizations, but externally provide a
-high-level interface to the application, abstracting away the details of
-the user interface. These wrappers are added to a shared repository,
-available to all users of the system. When an end user is using a site
-that already has an adapter implemented, they benefit from a semantic
+both worlds. Third-party programmers implement site-specific adapters
+that are internally implemented as ad hoc customizations, but externally
+provide a high-level interface to the application, abstracting away the
+details of the user interface. These wrappers are added to a shared
+repository, available to all users of the system. When an end user is
+using a site that already has an adapter, they benefit from a semantic
 customization experience that avoids low-level details.
@@
 \href{https://github.com/KartikTalwar/gmail.js/}{Gmail.js}, an open
-source project that creates a convenient API for browser extensions to
-interface with the Gmail web email client.
+source project that wraps the Gmail web client in a convenient API for
+browser extensions to build on.
 
@@
 adapters have largely been the same people building customizations on
-top of them. One helpful future addition would be to allow end users to
-create wrappers themselves, or at least to modify existing ones. There
-are many existing projects exploring end-user web scraping (such as
-Helena \citep{chasins2018}) which might prove helpful for this.
-
-The distribution mechanism for semantic wrappers is also important.
-Currently, the distribution mechanism is simply merging the code for all
-adapters into the main Wildcard codebase. This is the simplest workable
-solution, but has its downsides: contributing involves a fairly high
-barrier of creating a pull request on Github, and using newly
-contributed wrappers requires installing a new version of the extension.
-In the future we might explore other mechanisms, like an online
-repository that the extension pulls from dynamically. Security is also a
-consideration---DOM scraping adapters can execute arbitrary Javascript
-code, which opens up the possibility of malicious adapters being
-contributed. Currently we plan to solve this challenge with centralized
-code review, but another approach we are considering is using or
-inventing a more restricted domain-specific language for specifying
-scraping logic.
+top of them, so more work is needed to fully test this hypothesis.
+
+A helpful future addition would be to allow end users to create wrappers
+themselves, so that more people could contribute to the shared library.
+There are many existing projects exploring end-user web scraping (such
+as Helena \citep{chasins2018}) which might prove helpful; however,
+because the needs of a Wildcard adapter are slightly more complex than a
+traditional web scraper (as discussed in
+Section~\ref{sec:architecture}), further work might be needed to enable
+end-user wrapper creation.
+
+The distribution mechanism for semantic wrappers is also important for
+encouraging an ecosystem of shared wrappers. Currently, the distribution
+mechanism is simply merging the code for all adapters into the main
+Wildcard codebase. This is the simplest workable solution, but has its
+downsides: contributing involves a fairly high barrier of creating a
+pull request on Github, and using newly contributed wrappers requires
+installing a new version of the extension. In the future we might
+explore other mechanisms, like an online repository that the extension
+pulls from dynamically. Security is also a consideration---DOM scraping
+adapters can execute arbitrary Javascript code, which opens up the
+possibility of malicious adapters being contributed. Currently we plan
+to solve this challenge with centralized code review, but another
+approach we are considering is using or inventing a more restricted
+domain-specific language for specifying scraping logic.
 
@@
 
-Some recent web customization tools aim to enable end users to modify
-web interfaces without programming.
+More recently, web customization tools have aimed to enable end users to
+modify web interfaces without programming.
 
@@
 constructing the query. SIEUFERD's interface supports a far wider range
-of queries than Wildcard, but the fundamental principles of the user
-interface are similar. \href{https://airtable.com/}{Airtable} is another
-example of a modern commercial product that offers spreadsheet-like
-interaction with a relational database.
+of queries than Wildcard, but the core principles of the user interface
+are similar. \href{https://airtable.com/}{Airtable} is another example
+of a modern commercial product that offers spreadsheet-like interaction
+with a relational database.
 
@@
 \citep{benson2014}, Gneiss \citep{chang2014}, Marmite \citep{wong2007},
-\href{https://www.glideapps.com/}{Glide}. We share the core idea of
-connecting a spreadsheet view to a GUI, but apply it to software
+\href{https://www.glideapps.com/}{Glide}. We share the main idea of
+connecting a spreadsheet view to a GUI, but we apply it to software
 customization, rather than building software from scratch.
@@
 library of wrappers around existing web APIs. Unlike Wildcard, however,
-ScrAPIr targets explicit APIs exposed by developers. It also focused on
+ScrAPIr targets explicit APIs exposed by developers. It also focuses on
 backend services and doesn't aim to extend the frontend interfaces of
@@
 paradigm for customizing software by direct manipulation of the
-underlying structured data. We have demonstrated the paradigm using a
-browser extension, and have used it to create useful customizations for
-a variety of websites.
-
-Our primary goal for future work is to evaluate the system with a
-broader group of users. What barriers do end users face when using the
-system? What types of customizations do they choose to create? What
-formulas prove most helpful?
-
-Another area for future work is exploring whether it is possible to
-express a broader range of customizations by extending our table-editing
+underlying structured data. We have demonstrated the paradigm using the
+Wildcard browser extension, and have used it to create useful
+customizations for a variety of websites.
+
+So far, most usage of Wildcard has come from members of the project
+team. Our primary goal for future work is to evaluate the system with a
+broader group of users. What usability barriers do end users face when
+using the system? What types of customizations do they choose to create?
+What formulas prove most helpful? Through real deployment of the tool,
+we hope to validate the viability of creating a community-maintained
+library of site-specific adapters.
+
+Another area for future work is exploring whether it's possible to
+express a broader range of customizations by extending the table-editing
 paradigm. Are there ways to offer an increase in power and functional
 complexity, while retaining a programming model that is simpler for end
-users than conventional coding?
+users than conventional coding? For example, one possibility would be to
+enable users to set up triggers to perform actions like sending
+notifications when certain conditions are met in the table view.
 

### 2020-05-23 paper.tex
maybe i'm done

@@
   way for end users to extend software by direct manipulation without
-  doing traditional programming. We augment a user interface with a
-  table view showing the structured data inside the application. When
-  users edit the table, their changes are reflected in the original UI.
-  This simple model accommodates a spreadsheet formula language and
+  doing traditional programming. We augment existing user interfaces
+  with a table view showing the structured data inside the application.
+  When users edit the table, their changes are reflected in the original
+  UI. This simple model accommodates a spreadsheet formula language and
   custom data editing widgets, providing enough power to implement a
@@
 We subscribe to MacLean et al's vision of a ``gentle slope''
-\citep{maclean1990} free of such abrupt ``cliffs,'' where users should
-only need to make minimal and incremental investments in skill to
-achieve their desired customizations. We seek to contribute to this
-gentle slope with a new method for customizing software via direct
-manipulation, which requires less skill than scripting but still
-supports powerful customizations. We take inspiration from visual
-database query interfaces and spreadsheets, which have successfully
-enabled millions of end users to compute with data through direct
-manipulation.
+\citep{maclean1990} free of such ``cliffs,'' where users should only
+need to make minimal and incremental investments in skill to achieve
+their desired customizations. We seek to contribute to this gentle slope
+with a new method for customizing software via direct manipulation,
+which requires less skill than scripting but still supports powerful
+customizations. We take inspiration from visual database query
+interfaces and spreadsheets, which have successfully enabled millions of
+end users to compute with data through direct manipulation.
 
@@
 services, and more, all using the table as a mediating interface.
-Interacting with the table view feels like interacting with a familiar
+Interacting with the table view resembles interacting with a familiar
 spreadsheet, but results in customizing an existing application.
@@
 
-Finanlly, in Section~\ref{sec:conclusion}, we conclude and describe some
+Finally, in Section~\ref{sec:conclusion}, we conclude and describe
 opportunities for future work.
@@
 \textbf{Adding estimated read times}: Next, the user decides to attempt
-a more substantial customization: to add estimated read times to each
+a more substantial customization: adding estimated read times to each
 article, in order to prioritize reading deeper content.
@@
 implemented, since websites don't have access to the browser history for
-privacy reasons. But by using Wildcard, the user could implement the
-customization locally, without needing to expose their browser history
-to Hacker News.
+privacy reasons. But by using Wildcard, the user was able to implement
+the customization locally, without needing to expose their browser
+history to Hacker News.
 
@@
 
-The two tables are then combined into a single table for the end user to
+The two tables are then combined into a single table for the user to
 view and edit. The \textbf{query engine} is responsible for creating
@@
 and the column has a numeric type, the generic adapter extracts the
-number 1.23. The advantage of this approach is that the field can be
-\emph{writable}---when the user edits the value in the table, the
-generic adapter can simply overwrite the inner contents of the DOM
-element to reflect the change. In practice, this is most useful for
-editing the value of form input fields, although it's technically
-possible to enable the editing of other fields (e.g., editing the title
-of a post on Hacker News).
+number 1.23. The advantage of returning a DOM element is that the value
+is editable---when the user changes the value in the table, the generic
+adapter can simply overwrite the inner contents of the DOM element. In
+practice, this is most useful for editing the value of form input
+fields. It's technically possible to edit the contents of any element in
+the page (e.g.~editing the title of a post on Hacker News), but this
+often isn't particularly useful, so adapter authors have the ability to
+mark fields as readonly.
 
@@
 state to Wildcard with only minimal effort from the application
-developers. To test this hypothesis, we have created an early
+developers. To test this hypothesis, we created an early
 proof-of-concept of an integrated website adapter for the popular
 \href{https://redux.js.org/}{Redux} Javascript state management library.
-To set up this integrated adapter, the programmer would only need to
-make a few minor changes to their application code:
-
-\begin{itemize}
-\tightlist
-\item
-  \emph{Exposing a table}: In Redux, app developers already represent
-  the state of a user interface as a single object. To expose a table of
-  data to Wildcard, the developer simply writes a function that maps the
-  state object to a table.
-\item
-  \emph{Editing data values}: The developer writes handlers which define
-  how edits to table records should map to UI actions. For example,
-  editing the value of a todo might trigger an \texttt{EDIT\_TODO}
-  event.
-\item
-  \emph{Sorting/filtering the UI}: The developer inserts an additional
-  data processing step before the state object is rendered in the UI, so
-  that Wildcard can apply sorting and filtering.
-\item
-  \emph{Annotations}: The developer adds additional React components
-  managed by the Wildcard adapter into their UI tree, which render
-  additional data as annotations.
-\end{itemize}
+In Redux, app developers conveniently already represent the state of a
+user interface as a single object, so to expose a table of data to
+Wildcard, the developer simply writes a function that maps the
+centralized state object to a table.
 
@@
 built a small formula language resembling a spreadsheet formula
-language, except that formulas automatically apply across an entire
-column of data, and reference other column names instead of values in
-specific rows. This is more convenient than needing to copy-paste a
-formula across an entire column as in spreadsheets, and has worked for
-all of the customizations we have built. Visual database query systems
-like SIEUFERD \citep{bakke2016} use a similar approach to using formulas
-in a structured table.
+language. As in visual database query tools like SIEUFERD
+\citep{bakke2016} and \href{https://airtable.com/}{Airtable}, formulas
+automatically apply across an entire column of data, and reference other
+column names instead of values in specific rows. This is more convenient
+than needing to copy-paste a formula across an entire column as in
+spreadsheets, and has worked for all of the customizations we have
+built.
 
@@
 
-We provide a basic table editor as the user interface on top of the
-query engine. Our table editor is built with the Handsontable Javascript
-library, which provides UI elements for viewing, editing, sorting, and
-filtering a table.
+We provide a table editor view as the user interface on top of the query
+engine. Our table editor is built with the Handsontable Javascript
+library, which provides built-in UI elements for viewing, editing,
+sorting, and filtering a table.
 
@@
 We developed site-specific adapters for 11 websites that we personally
-use frequently, and then built customizations in Wildcard for those
-websites. Table 1 summarizes these results, showing the number of lines
-of code in the adapter for each site, and some example customizations we
-created. Here we offer our reflections from these experiences using the
-system, focused on two key questions:
+use frequently, and then built customizations for those websites using
+the Wildcard table view. Table 1 summarizes these results, showing the
+number of lines of code in the adapter for each site, and some example
+customizations we created. Here we offer our reflections from these
+experiences using the system, focused on two key questions:
 
@@
 table-driven customization. Some of the limitations are specific to the
-current implementation of the Wildcard extension, but many are more
+current implementation of the Wildcard extension, but others are more
 fundamental to the general paradigm.
@@
 Some of the challenges of writing a DOM scraping adapter are the same
-ones as with writing normal web scraping code, but the more interactive
-nature of Wildcard introduces additional challenges. One challenge is
-registering appropriate event handlers to update the table data in
-response to UI changes that happen after initial page load. Another
-challenge is persisting updates to the DOM---some websites use virtual
-DOM frameworks that can occasionally overwrite changes made by Wildcard.
-So far, in practice we've managed to work around these issues for all of
-the websites we've tried, but we don't claim that any website can be
-customized through DOM scraping. As web frontend code gets increasingly
-complex (and starts to move beyond the DOM to other technologies like
-Shadow DOM or even WebGL), it may become increasingly difficult to
-customize websites from the outside without first-party support.
-
-AJAX scraping proved highly useful in several cases. The Uber Eats
-website was challenging to scrape because it has a complex DOM structure
-with machine-generated CSS classes, but the site also uses AJAX requests
+ones as with writing normal web scraping code. Sometimes, addressing the
+desired set of elements can be difficult, and when sites change,
+scrapers can break---we observed several instances where sites changed
+their CSS classes and caused Wildcard adapters to no longer work. One
+benefit of a library of shared wrappers is that if many customizations
+depend on some piece of scraping logic, rather than having the scraping
+logic embedded in a single browser extension, it should be more likely
+to be fixed quickly.
+
+The interactive nature of Wildcard also introduces additional challenges
+beyond normal web scraping. One challenge is registering appropriate
+event handlers to update the table data in response to UI changes that
+happen after initial page load. Another challenge is persisting updates
+to the DOM---some websites use virtual DOM frameworks that can
+occasionally overwrite changes made by Wildcard. So far, in practice
+we've managed to work around these issues for all of the websites we've
+tried, but we don't claim that any website can be customized through DOM
+scraping. As web frontend code gets increasingly complex (and starts to
+move beyond the DOM to other technologies like Shadow DOM or even
+WebGL), it may become increasingly difficult to customize websites from
+the outside without first-party support.
+
+AJAX scraping proved very useful in several cases. The Uber Eats website
+was challenging to scrape because it has a complex DOM structure with
+machine-generated CSS classes, but the site also uses AJAX requests
 which contain all the relevant data in a structured form that is much
@@
 additional complexity, but spreadsheets have demonstrated that formula
-programming is still accessible to many users, due to the pure
+programming is still accessible to many users, helped by the pure
 functional semantics and the intermediate visibility of results.
@@
 is enabling customization in closer proximity to the original user
-interface elements, as explored by tools like Scotty \citep{eagan2011}.
-We have found that augmenting the original UI with a distinct,
-additional representation provides a more consistent experience across
-all applications, and clearly shows the user what structured data is
-available to work with.
+interface elements, as explored by tools like Scotty
+\citep{eagan2011}---all customization occurs in a separate panel next to
+the original UI. While closer proximity might be helpful, we have found
+that augmenting the original UI with a distinct, additional
+representation provides a more consistent experience across all
+applications, and clearly shows what structured data is available to
+work with.
 
@@
 \citep{benson2014}, Gneiss \citep{chang2014}, Marmite \citep{wong2007},
-\href{https://www.glideapps.com/}{Glide}. We share the main idea of
+and \href{https://www.glideapps.com/}{Glide}. We share the main idea of
 connecting a spreadsheet view to a GUI, but we apply it to software
@@
 What formulas prove most helpful? Through real deployment of the tool,
-we hope to validate the viability of creating a community-maintained
-library of site-specific adapters.
+we hope to validate the viability of a community-maintained library of
+site-specific adapters.
 
@@
 incentives are not always aligned with the user's interests. We hope
-that table-driven customization can serve as one point, and maybe a
-guidepost, on our path from normal use to deep modification, in support
-of a more adaptable experience for all computer users.
+that table-driven customization can serve as a point on the path from
+normal use to deep modification, in support of a more adaptable
+experience for all computer users.
 

### 2020-08-09 initial-submission.tex
record initial submission files

--- 2020-05-23 paper.tex
@@
   wrappers around existing applications, enabling end users to work with
-  domain objects rather than low-level representations.
+  domain data rather than low-level representations.
 \end{itemize}
@@
 applications, and clearly shows what structured data is available to
-work with.
+work with. We also emphasize the mapping between the representations by
+highlighting content in the original page, similiar to the way that
+browser developer tools highlight the currently selected element in the
+DOM inspector in the original page.
 
@@
 This paper extends work reported in a workshop paper by Litt and Jackson
-\citep{litt2020} which presented an earlier version of the Wildcard
+\citep{litt2020} which presented an early version of the Wildcard
 extension. We have substantially extended their work in this paper by
-creating the table adapter abstarction, reimplementing the internals of
+creating the table adapter abstraction, reimplementing the internals of
 Wildcard around that abstraction, evaluating the system on many more
-websites, and by characterizing the design of the system in much more
-detail than in their workshop paper.
+websites and use cases, and by characterizing the design of the system
+in much more detail than in their workshop paper.
 
@@
 More recently, web customization tools have aimed to enable end users to
-modify web interfaces without programming.
-
-Sifter \citep{huynh2006} enables end users to sort and filter lists of
-data obtained by web scraping, much like Wildcard's sorting features.
-The main difference between the systems is that table-driven
-customization has many other use cases besides sorting and filtering.
-Also, Sifter involves end users in a semi-automated data extraction
-process, rather than having programmers create wrappers. This provides
-coverage of more websites, but at the expense of complicating the end
-user experience. We could integrate end user scraping techniques in
-Wildcard in the future, but we believe that when possible it's valuable
-for end users to have a customization experience decoupled from the
-challenge of web scraping the underlying data. Sifter also implements
-scraping across multiple pages, a valuable feature for sorting and
-filtering that isn't present in Wildcard.
+modify web interfaces without programming. Sifter \citep{huynh2006}
+enables end users to sort and filter lists of data obtained by web
+scraping, much like Wildcard's sorting features. The main difference
+between the systems is that table-driven customization has many other
+use cases besides sorting and filtering. Also, Sifter involves end users
+in a semi-automated data extraction process, rather than having
+programmers create wrappers. This provides coverage of more websites,
+but at the expense of complicating the end user experience. We could
+integrate end user scraping techniques in Wildcard in the future, but we
+believe that when possible it's valuable for end users to have a
+customization experience decoupled from the challenge of web scraping
+the underlying data. Sifter also implements scraping across multiple
+pages, a valuable feature for sorting and filtering that isn't present
+in Wildcard.
 
@@
 
-\hypertarget{direct-manipulation-programming-interfaces.}{%
-\subsection{Direct manipulation programming
-interfaces.}\label{direct-manipulation-programming-interfaces.}}
-
-Another relevant area involves spreadsheets and visual query systems. We
-take inspiration from these tools in our work, but apply them in a
+\hypertarget{spreadsheets-and-visual-query-interfaces}{%
+\subsection{Spreadsheets and visual query
+interfaces}\label{spreadsheets-and-visual-query-interfaces}}
+
+Another relevant area involves spreadsheets and visual query interfaces.
+We take inspiration from these tools in our work, but apply them in a
 different domain: customizing existing software applications, rather
@@
 survey of many other similar tools. Our work is particularly influenced
-by the authors' observation that direct manipulation requires that the
-user manipulate the results of a database query rather than the query
-itself, and that the user must see intermediate results at every step of
-constructing the query. SIEUFERD's interface supports a far wider range
-of queries than Wildcard, but the core principles of the user interface
-are similar. \href{https://airtable.com/}{Airtable} is another example
-of a modern commercial product that offers spreadsheet-like interaction
-with a relational database.
+by the authors' observation that a user should be able to modify queries
+by interacting with the results of the query rather than some
+representation of the query itself. SIEUFERD's interface supports a far
+more general range of queries than Wildcard, but the core principles of
+the user interface are the same. \href{https://airtable.com/}{Airtable}
+is another example of a modern commercial product that offers
+spreadsheet-like interaction with a relational database.
 

### 2020-08-15 compile-diff.sh
add a scraping sample

@@ -1,0 +1,8 @@
+set -e
+
+latexdiff initial-submission.tex paper.tex > diff.tex
+pdflatex diff.tex
+bibtex diff
+pdflatex -interaction=batchmode diff.tex
+pdflatex -interaction=batchmode diff.tex
+open diff.pdf

### 2020-08-15 hacker-news-code.js

@@ -1,0 +1,51 @@
+
+const HNAdapter = createDomScrapingAdapter({
+  name: "Hacker News",
+
+  // Specify when the adapter should be enabled, based on current URL
+  enabled() {
+    return (
+      urlExact("news.ycombinator.com/") ||
+      urlContains("news.ycombinator.com/news") ||
+      urlContains("news.ycombinator.com/newest")
+    );
+  },
+
+  // Define the name and type of each column in the table
+  attributes: [
+    { name: "id", type: "text", hidden: true },
+    { name: "rank", type: "numeric" },
+    { name: "title", type: "text" },
+    { name: "link", type: "text" },
+  ],
+
+  // Define scraping logic for extracting data
+  scrapePage() {
+    return Array.from(document.querySelectorAll("tr.athing")).map((el) => {
+      let detailsRow = el.nextElementSibling;
+      let spacerRow = detailsRow.nextElementSibling;
+
+      return {
+        // Return a unique ID for each row
+        id: String(el.getAttribute("id")),
+
+        // Return DOM elements corresponding to this row
+        // (this enables DOM manipulation for sorting/filtering)
+        rowElements: [el, detailsRow, spacerRow],
+
+        // Return data values for the row by scraping
+        dataValues: {
+          rank: el.querySelector("span.rank"),
+          title: el.querySelector("a.storylink"),
+          link: el.querySelector("a.storylink").getAttribute("href"),
+        },
+
+        // Specify where annotations should be injected
+        annotationContainer: detailsRow.querySelector("td.subtext") as HTMLElement,
+        annotationTemplate: `| <span style="color: #f60;">$annotation</span>`,
+      };
+    });
+  },
+});
+
+export default HNAdapter;

### 2020-08-15 paper.tex
finish addressing scrapers

@@
 %%\citestyle{acmauthoryear}
-
 %%
 %% end of the preamble, start of the body of the document source.
+
+\usepackage{listings}
+
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\newenvironment{Shaded}{}{}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
+
 \begin{document}
@@
 \begin{abstract}
-  In this paper we introduce \emph{table-driven customization}, a new
-  way for end users to extend software by direct manipulation without
-  doing traditional programming. We augment existing user interfaces
-  with a table view showing the structured data inside the application.
-  When users edit the table, their changes are reflected in the original
-  UI. This simple model accommodates a spreadsheet formula language and
+  In this paper we introduce \emph{data-driven customization}, a new way
+  for end users to extend software by direct manipulation without doing
+  traditional programming. We augment existing user interfaces with a
+  table view showing the structured data inside the application. When
+  users edit the table, their changes are reflected in the original UI.
+  This simple model accommodates a spreadsheet formula language and
   custom data editing widgets, providing enough power to implement a
@@
   We illustrate the approach with Wildcard, a browser extension that
-  implements table-driven customization for existing applications using
+  implements data-driven customization for existing applications using
   web scraping. Through concrete examples, we show that this paradigm
@@
 
-We subscribe to MacLean et al's vision of a ``gentle slope''
+We subscribe to MacLean et al.'s vision of a ``gentle slope''
 \citep{maclean1990} free of such ``cliffs,'' where users should only
@@
 with a new method for customizing software via direct manipulation,
-which requires less skill than scripting but still supports powerful
-customizations. We take inspiration from visual database query
-interfaces and spreadsheets, which have successfully enabled millions of
-end users to compute with data through direct manipulation.
-
-In our proposed paradigm, \emph{table-driven customization}, an
+taking inspiration from visual database query interfaces and
+spreadsheets, which have successfully enabled millions of end users to
+compute with data through direct manipulation.
+
+In our proposed paradigm, \emph{data-driven customization}, an
 application's UI is augmented with a table view where the user can see
@@
 \includegraphics[width=\textwidth]{media/overview.eps}
-\caption{An overview of table-driven customization}\label{fig:overview}
+\caption{An overview of data-driven customization}\label{fig:overview}
 }
@@
 We have developed a browser extension called Wildcard which uses web
-scraping techniques to implement table-driven customization for existing
-Web applications. In Section~\ref{sec:example}, we illustrate the end
-user experience of table-driven customization through an example
-scenario, using Wildcard to add features to Hacker News.
+scraping techniques to implement data-driven customization for existing
+Web applications. In Section~\ref{sec:example}, we concretely
+demonstrate the end user experience of using Wildcard to add features to
+a news website.
 
 In Section~\ref{sec:architecture}, we explain the system architecture of
-table-driven customization. We focus on the \emph{table adapter}
+data-driven customization. We focus on the \emph{table adapter}
 abstraction, which allows many different types of underlying data to be
@@
 
-In Section~\ref{sec:themes}, we discuss some key themes from our work:
+In Section~\ref{sec:visions}, we discuss some key themes from our work:
 
@@
   \emph{Customization by direct manipulation}: We discuss further how
-  table-driven customization relates to a gentle slope of customization.
+  data-driven customization relates to a gentle slope of customization.
   We suggest that an important point on this slope is the ability to
@@
   wrappers around existing applications, enabling end users to work with
-  domain objects rather than low-level representations.
+  domain data rather than low-level representations.
 \end{itemize}
 
-Table-driven customization relates to existing work in many areas. Our
+data-driven customization relates to existing work in many areas. Our
 goals overlap with many software customization tools, and our methods
@@
 
-To illustrate the end user experience of table-driven customization, we
-consider an example scenario of customizing
+To concretely illustrate the user experience of data-driven
+customization, we present a scenario of customizing
 \href{https://news.ycombinator.com/}{Hacker News}, a popular tech news
@@
 
-This formula calls a built-in function that takes a URL as an argument
-and uses a public web API to compute an estimated read time for the
-URL's contents. The \texttt{link} argument in the formula refers to a
-column name in the table; the formula is automatically evaluated across
-all rows in the table, using the value of \texttt{link} for each row.
+This formula calls a built-in function \texttt{ReadTimeInSeconds} that
+uses a third-party public web API to compute an estimated read time for
+the URL's contents. The \texttt{link} argument in the formula refers to
+a column name in the table; the formula is automatically evaluated
+across all rows in the table, using the value of \texttt{link} for each
+row.
 
@@
 
-This scenario has shown a few examples of how table-driven
-customizations can help a user improve their experience of a website.
-Section~\ref{sec:reflections} explains many other use cases, but first
-we explain how the system works internally.
+This scenario has shown a few examples of how data-driven customizations
+can help a user improve their experience of a website.
+Section~\ref{sec:reflections} explains many other use cases and contexts
+where the technique applies, but first we explain how the system works
+internally.
 
@@
 Figure~\ref{fig:table-adapter} summarizes the overall architecture of
-table-driven customization, using a simplified illustration of the
-Hacker News example scenario. The name and points value for each article
-is scraped from the web page DOM, and user annotations are loaded from
-the brower's local storage.
+data-driven customization, using a simplified illustration of the Hacker
+News example scenario. The name and points value for each article is
+scraped from the web page DOM, and user annotations are loaded from the
+brower's local storage.
 
@@
 
-A key idea in table-driven customization is that a wide variety of data
+A key idea in data-driven customization is that a wide variety of data
 sources can be mapped to a generic table abstraction. In a relational
 database, the table matches the underlying storage format, but in
-table-driven customization, the table is merely an \emph{interface
+data-driven customization, the table is merely an \emph{interface
 layer}. The data shown in the table is a projection of some underlying
@@
 
-Here we describe the abstract interface fulfilled by a table adapter,
-and then describe specific types of table adapters we've built so far in
-the Wildcard system.
-
-\hypertarget{abstract-interface}{%
-\subsubsection{Abstract interface}\label{abstract-interface}}
-
-\emph{Returning a table}: A table adapter exposes a table of data: an
-ordered list of records. Each record carries a unique identifier and
-associates named attributes with values. Tables have a typed schema, so
-the same attributes are shared across all records. We currently support
-strings, numeric values, booleans, and datetimes as types. The columns
-also carry some additional metadata, such as whether or not they are
-read-only or editable.
-
-A table adapter can update the contents of a table at any time in
-response to changes in the underlying state (e.g., a DOM scraping
-adapter can update the table when the page body changes). When data
-changes, the query view is reactively updated in response.
-
-\emph{Handling edits}: The query engine can issue a request to a table
-adapter to make an edit to a record. The meaning of making an edit can
-vary depending on the adapter: in the local storage adapter, a new value
-may be persisted into local storage; in the DOM scraping adapter, an
-edit may result in changing the value of a form field.
-
-In addition, the query engine also sends additional information about
-the combined query view to each table adapter:
-
-\emph{Sorting/filtering}: When the user sorts or filters the query view,
-an ordered list of visible IDs is sent to each table adapter. The DOM
-scraping adapter uses this information to change the list of rows shown
-in the web page.
-
-\emph{Data from other tables}: The query engine provides each table
-adapter with the entire combined table shown to the user. The DOM
-scraping adapter uses this for injecting annotations---values from other
-tables are added to the original web page.
-
-\emph{Currently selected record}: As the user clicks around the table
-view, the query engine broadcasts the record currently selected by the
-user to each table adapter. The DOM scraping adapter uses this
-information to highlight the row in the page that corresponds to the
-selected row in the table, which helps the user understand the mapping
-between the table and the original UI.
+Here we describe the three types of table adapters we have built so far
+in Wildcard. These do not represent an exhaustive set of all possible
+table adapters---in Section~\ref{sec:visions} we discuss other types of
+adapters that would fit well into the general paradigm.
 
@@
 
-DOM scraping adapters are the essential component that enables Wildcard
-to interface with an existing website UI. In addition to the standard
-web scraping problem of extracting a table of data from the DOM, a
-scraping adapter must also manipulate the DOM to reorder rows, edit form
-entries, and inject annotations as the table is edited.
-
-Because every website has unique content, we rely on programmers to
-create a DOM scraping adapter for each individual website to make it
-available for customization in Wildcard. To make this approach viable,
-we have built a generic DOM scraping adapter which a programmer can
-configure with the minimal site-specific parts.
-
-\emph{Basic metadata}: The programmer specifies URL patterns that the
-adapter should activate on, and a set of column names and types to
-return in the table.
-
-\emph{Specifying scraping logic}: The main component of a site-specific
-adapter is a single Javascript function that extracts relevant elements
-from the page using web scraping technqiues. For each scraped row of
-data, the programmer must return a unique identifier, and values for
-each column in the table.
-
-The generic adapter then wraps this scraping function to implement the
-table adapter interface. For example, when the user sorts the table
-view, the generic adapter synchronizes the sort order to the DOM by
-removing the DOM elements corresponding to the rows in the table and
-reinserting them back into the page in the new order.
-
-\emph{Choosing an ID}: When possible, it is best to choose a server-side
-identifier that remains stable across pageloads. This enables user
-annotations persisted in local storage to be associated with the same
-records on subsequent pageloads. We have found that it's usually
-possible to find such an identifier; for example, each item in a page
-often contains a link to a page with more details, with a URL that
+DOM scraping adapters enable Wildcard to interface with an existing
+website UI. A DOM scraping adapter fulfills the standard web scraping
+task of extracting a table of data from the DOM, but it must also handle
+the reverse direction: manipulating the DOM to reflect edits to the
+table.
+
+Currently, we require DOM scraping adapters to be programmed manually
+for each website using Javascript code. It might seem that this
+prohibits non-programmer users from using the system, but we solve this
+problem via a shared repository of adapters. Once an adapter is
+programmed for a website, it is added to a shared code repository,
+enabling any end user to perform customizations on that website.
+
+There are other ways a DOM scraping adapter could be produced which
+would reduce this dependence on programmers: an end user could specify
+the scraping logic via demonstration, or the desired data table could be
+automatically inferred from the page. While we are interested in these
+techniques and discuss them later in Section~\ref{sec:visions}, we
+believe that a shared repository of manually programmed adapters is a
+pragmatic starting point; given that many users visit the same popular
+websites,a critical mass of adapters could serve the needs of many
+users.
+
+To ease the burden on programmers of creating these adapters, we have
+created a framework makes the process feel more like writing
+unidirectional scraping code than performing a complex bidirectional
+synchronization. The key idea is this: programmers return pointers to
+DOM elements representing table rows and table cells; Wildcard extracts
+data from these DOM elements, but it also uses the pointers to
+synchronize table edits back into the page. For example, when the user
+sorts the table, the DOM elements representing the table rows are moved
+around in the DOM to reflect the new sorted order.
+
+\begin{figure*}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{const}\NormalTok{ HNAdapter }\OperatorTok{=} \AttributeTok{createDomScrapingAdapter}\NormalTok{(}\OperatorTok{\{}
+  \DataTypeTok{name}\OperatorTok{:} \StringTok{"Hacker News"}\OperatorTok{,}
+
+  \CommentTok{// Specify when the adapter should be enabled, based on current URL}
+  \AttributeTok{enabled}\NormalTok{() }\OperatorTok{\{}
+    \ControlFlowTok{return}\NormalTok{ (}
+      \AttributeTok{urlExact}\NormalTok{(}\StringTok{"news.ycombinator.com/"}\NormalTok{) }\OperatorTok{||}
+      \AttributeTok{urlContains}\NormalTok{(}\StringTok{"news.ycombinator.com/news"}\NormalTok{) }\OperatorTok{||}
+      \AttributeTok{urlContains}\NormalTok{(}\StringTok{"news.ycombinator.com/newest"}\NormalTok{)}
+\NormalTok{    )}\OperatorTok{;}
+  \OperatorTok{\},}
+
+  \CommentTok{// Define the name and type of each column in the table}
+  \DataTypeTok{attributes}\OperatorTok{:}\NormalTok{ [}
+    \OperatorTok{\{} \DataTypeTok{name}\OperatorTok{:} \StringTok{"id"}\OperatorTok{,} \DataTypeTok{type}\OperatorTok{:} \StringTok{"text"}\OperatorTok{,} \DataTypeTok{hidden}\OperatorTok{:} \KeywordTok{true} \OperatorTok{\},}
+    \OperatorTok{\{} \DataTypeTok{name}\OperatorTok{:} \StringTok{"rank"}\OperatorTok{,} \DataTypeTok{type}\OperatorTok{:} \StringTok{"numeric"} \OperatorTok{\},}
+    \OperatorTok{\{} \DataTypeTok{name}\OperatorTok{:} \StringTok{"title"}\OperatorTok{,} \DataTypeTok{type}\OperatorTok{:} \StringTok{"text"} \OperatorTok{\},}
+    \OperatorTok{\{} \DataTypeTok{name}\OperatorTok{:} \StringTok{"link"}\OperatorTok{,} \DataTypeTok{type}\OperatorTok{:} \StringTok{"text"} \OperatorTok{\},}
+    \CommentTok{// ... other columns omitted for brevity}
+\NormalTok{  ]}\OperatorTok{,}
+
+  \CommentTok{// Iterate over DOM elements, returning information about each row}
+  \AttributeTok{scrapePage}\NormalTok{() }\OperatorTok{\{}
+    \ControlFlowTok{return} \VariableTok{Array}\NormalTok{.}\AttributeTok{from}\NormalTok{(}\VariableTok{document}\NormalTok{.}\AttributeTok{querySelectorAll}\NormalTok{(}\StringTok{"tr.athing"}\NormalTok{)).}\AttributeTok{map}\NormalTok{((el) }\KeywordTok{=>} \OperatorTok{\{}
+      \KeywordTok{let}\NormalTok{ detailsRow }\OperatorTok{=} \VariableTok{el}\NormalTok{.}\AttributeTok{nextElementSibling}\OperatorTok{;}
+      \KeywordTok{let}\NormalTok{ spacerRow }\OperatorTok{=} \VariableTok{detailsRow}\NormalTok{.}\AttributeTok{nextElementSibling}\OperatorTok{;}
+
+      \ControlFlowTok{return} \OperatorTok{\{}
+        \CommentTok{// Return a unique ID for each row}
+        \DataTypeTok{id}\OperatorTok{:} \AttributeTok{String}\NormalTok{(}\VariableTok{el}\NormalTok{.}\AttributeTok{getAttribute}\NormalTok{(}\StringTok{"id"}\NormalTok{))}\OperatorTok{,}
+
+        \CommentTok{// Return DOM elements corresponding to this row}
+        \CommentTok{// (this enables moving/hiding the elements for sorting/filtering)}
+        \DataTypeTok{rowElements}\OperatorTok{:}\NormalTok{ [el}\OperatorTok{,}\NormalTok{ detailsRow}\OperatorTok{,}\NormalTok{ spacerRow]}\OperatorTok{,}
+
+        \CommentTok{// Return data for each column}
+        \DataTypeTok{dataValues}\OperatorTok{:} \OperatorTok{\{}
+          \DataTypeTok{rank}\OperatorTok{:} \VariableTok{el}\NormalTok{.}\AttributeTok{querySelector}\NormalTok{(}\StringTok{"span.rank"}\NormalTok{)}\OperatorTok{,}
+          \DataTypeTok{title}\OperatorTok{:} \VariableTok{el}\NormalTok{.}\AttributeTok{querySelector}\NormalTok{(}\StringTok{"a.storylink"}\NormalTok{)}\OperatorTok{,}
+          \DataTypeTok{link}\OperatorTok{:} \VariableTok{el}\NormalTok{.}\AttributeTok{querySelector}\NormalTok{(}\StringTok{"a.storylink"}\NormalTok{).}\AttributeTok{getAttribute}\NormalTok{(}\StringTok{"href"}\NormalTok{)}\OperatorTok{,}
+          \CommentTok{// ... other columns omitted for brevity}
+        \OperatorTok{\},}
+
+        \CommentTok{// Specify where annotations should be injected, and what they should look like}
+        \DataTypeTok{annotationContainer}\OperatorTok{:} \VariableTok{detailsRow}\NormalTok{.}\AttributeTok{querySelector}\NormalTok{(}\StringTok{"td.subtext"}\NormalTok{) }\ImportTok{as}\NormalTok{ HTMLElement}\OperatorTok{,}
+        \DataTypeTok{annotationTemplate}\OperatorTok{:} \VerbatimStringTok{\textasciigrave{}| <span style="color: \#f60;">$annotation</span>\textasciigrave{}}\OperatorTok{,}
+      \OperatorTok{\};}
+    \OperatorTok{\}}\NormalTok{)}\OperatorTok{;}
+  \OperatorTok{\},}
+\OperatorTok{\}}\NormalTok{)}\OperatorTok{;}
+\end{Highlighting}
+\end{Shaded}
+
+\caption{Source code for the Hacker News scraper. Some details removed for brevity.}\label{fig:scraper-code}
+\end{figure*}
+
+Figure~\ref{fig:scraper-code} shows an example of the scraper code used
+for the Hacker News example (with some code eliminated for brevity.) It
+defines the following main components:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{enabled}: defines when this adapter should run, usually based
+  on the active URL in the browser.\footnote{Currently Wildcard can only
+    show a single table at a time, so if multiple adapters are enabled
+    for a single page, we arbitrarily pick one. It would be a
+    straightforward extension to allow the user to switch between
+    multiple possible tables available on the page.}
+\item
+  \texttt{attributes}: defines a schema for the table, with a name and
+  type for each column
+\item
+  \texttt{scrapePage}: defines a scraping function which returns an
+  array of objects, each containing the data for a single row of the
+  table.
+\end{itemize}
+
+Here are some of the concerns that emerge when building adapters in
+practice:
+
+\emph{Choosing a row ID}: When possible, it is best to choose a
+server-side identifier that remains stable across pageloads. This
+enables user annotations persisted in local storage to be associated
+with the same records on subsequent pageloads. We have found that it's
+usually possible to find such an identifier; for example, each item in a
+page often contains a link to a page with more details, with a URL that
 contains a stable ID.
@@
 adapter extracts the text contents of the DOM element and casts them to
-the type of the column. For example, if the scraping function returns a
-\texttt{\textless{}div\textgreater{}} containing the string ``1.23'',
-and the column has a numeric type, the generic adapter extracts the
-number 1.23. The advantage of returning a DOM element is that the value
-is editable---when the user changes the value in the table, the generic
-adapter can simply overwrite the inner contents of the DOM element. In
-practice, this is most useful for editing the value of form input
-fields. It's technically possible to edit the contents of any element in
-the page (e.g.~editing the title of a post on Hacker News), but this
-often isn't particularly useful, so adapter authors have the ability to
-mark fields as readonly.
+the type of the column. The advantage of returning a DOM element is that
+the value is editable---when the user changes the value in the table,
+the generic adapter can simply overwrite the inner contents of the DOM
+element.
 
@@
 
-\emph{Other site-specific customization}: The programmer can optionally
-override other behavior of the generic DOM scraping adapter, including
-logic for when to re-run the scraping function in response to page
-changes, how to style injected annotations, and how to style the
-currently selected row in the table. We've chosen sensible defaults that
-generally result in a workable adapter without needing to change these
-options, but they can improve the experience on a specific website. For
-example, by default we highlight the currently selected row with a blue
-border, but on some sites this is hard to see or breaks the layout, so
-setting a background color works better.
+\emph{Optional overrides}: In order to turn a unidirectional scraping
+function into a bidirectional scraping adapter, there are a number of
+behaviors that must be specified:
+
+\begin{itemize}
+\tightlist
+\item
+  when should the scraping function be re-run in response to changes on
+  the page?
+\item
+  how should injected annotations appear in table rows?
+\item
+  when the user selects a row in the table, how should the corresponding
+  row in the DOM be highlighted?
+\end{itemize}
+
+The scraping framework defines sensible defaults that work well on many
+sites, but the programmer can optionally override them to provide better
+site-specific behavior. For example, the Hacker News adapter specifies
+annotation options which make user annotations appear more naturally in
+the context of the original design.
 
@@
 re-associated with the same records on subsequent pageloads.
-
-\hypertarget{future-adapters}{%
-\subsubsection{Future Adapters}\label{future-adapters}}
-
-We have designed the table adapter API to be general enough to support
-other types of useful adapters in the future. Here are two examples:
-
-\textbf{Integrated website adapters}: A key benefit of the table adapter
-abstraction is that Wildcard is not coupled to web scraping as the only
-means for integrating with existing sites, but can also accommodate
-first party developers adding support directly into their own websites.
-An ``integrated website adapter'' installed by the developer could
-directly accesses the internal state of the application, providing the
-same functionality as a DOM scraping adapter but in a more robust way.
-
-With the advent of rich frontend web frameworks, structured application
-state is now often available in the web client. We suspect it is
-possible to create plugins for frontend frameworks that expose this
-state to Wildcard with only minimal effort from the application
-developers. To test this hypothesis, we created an early
-proof-of-concept of an integrated website adapter for the popular
-\href{https://redux.js.org/}{Redux} Javascript state management library.
-In Redux, app developers conveniently already represent the state of a
-user interface as a single object, so to expose a table of data to
-Wildcard, the developer simply writes a function that maps the
-centralized state object to a table.
-
-Customizability is sometimes a key selling point for software. An
-integrated website adapter would provide a way for developers to
-integrate with an ecosystem of formulas and customization tools without
-needing to build all that functionality from scratch.
-
-\textbf{Shared storage adapter}: It would be useful to share user
-annotations among users and across devices---for example,
-collaboratively taking notes with friends on options for places to stay
-together on a trip. The existing local storage adapter could be extended
-to share live synchronized data with other users. This could be achieved
-through a centralized web server, or through P2P connections that might
-provide stronger privacy guarantees.
 
@@
 \end{tabular}
-\caption{A list of table-driven customizations that we have implemented using Wildcard.}
+\caption{A list of data-driven customizations that we have implemented using Wildcard.}
 \label{tab:websites}
@@
 
-To evaluate table-driven customization in practice, we built the
-Wildcard browser extension, which implements table-driven customization
-in the context of existing websites. It is implemented in Typescript,
-and works across three major browsers: Chrome, Firefox, and Edge.
+To evaluate data-driven customization in practice, we built the Wildcard
+browser extension, which implements data-driven customization in the
+context of existing websites. It is implemented in Typescript, and works
+across three major browsers: Chrome, Firefox, and Edge.
 
@@
 
-We have found that table-driven customization can serve a broad range of
+We have found that data-driven customization can serve a broad range of
 useful purposes. Here we expand on some archetypal examples that
@@
 There are many customizations that are not possible to implement with
-table-driven customization. Some of the limitations are specific to the
+data-driven customization. Some of the limitations are specific to the
 current implementation of the Wildcard extension, but others are more
@@
 Our second evaluation area relates less to the conceptual approach of
-table-driven customization, and more to the specific implementation of
+data-driven customization, and more to the specific implementation of
 customizing existing web applications. In order for third-party
@@
 
-\hypertarget{sec:themes}{%
-\section{Key themes}\label{sec:themes}}
-
-Here we discuss some themes that we've explored with this work, which go
-beyond our specific approach and address broader issues in software
-customization.
+\hypertarget{sec:visions}{%
+\section{Vision}\label{sec:visions}}
+
+\begin{itemize}
+\tightlist
+\item
+  Wildcard just one instance of a broader data-driven customization
+
+  \begin{itemize}
+  \tightlist
+  \item
+    particular view: table
+  \item
+    particular context: web
+  \end{itemize}
+\item
+  here we share some of the underlying ideas. go beyond Wildcard, room
+  to grow
+\end{itemize}
 
@@
 automations that simulate a series of steps taken by the user. There is,
-however, a serious drawback to this approach. MacLean et al
-\citep{maclean1990} describe an ideal for user-tailorable systems: a
+however, a serious drawback to this approach. MacLean et
+al.~\citep{maclean1990} describe an ideal for user-tailorable systems: a
 ``gentle slope'' from using to customizing, where small incremental
@@
 investment in learning even to implement the simplest customizations.
-Another goal of MacLean et al is to make it ``as easy to change the
+Another goal of MacLean et al.~is to make it ``as easy to change the
 environment as it is to use it''---at least for some subset of changes.
@@
 
-With table-driven customization we aim to provide a gentler slope, by
+With data-driven customization we aim to provide a gentler slope, by
 using direct manipulation for software customization. The data shown in
@@
 interactions are common in GUI applications, and Wildcard therefore
-seems to meet MacLean et al's goal: some one-click customizations are as
-easy as using the original application. Formulas introduce some
+seems to meet MacLean et al.'s goal: some one-click customizations are
+as easy as using the original application. Formulas introduce some
 additional complexity, but spreadsheets have demonstrated that formula
 programming is still accessible to many users, helped by the pure
-functional semantics and the intermediate visibility of results.
+functional semantics and the visibility of intermediate results.
 
@@
 applications, and clearly shows what structured data is available to
-work with.
-
-Ainsworth et al provide a helpful taxonomy of the value of multiple
+work with. We also emphasize the map ping between the representations by
+highlighting content in the original page, similiar to the way that
+browser developer tools highlight the currently selected element in the
+DOM inspector in the original page.
+
+Ainsworth et al.~provide a helpful taxonomy of the value of multiple
 representations \citep{ainsworth1999}. In their terms, Wildcard plays a
@@
 
+\hypertarget{generic-data-interfaces}{%
+\subsection{Generic data interfaces}\label{generic-data-interfaces}}
+
+\begin{itemize}
+\item
+  what makes a UI good for customization?
+
+  \begin{itemize}
+  \tightlist
+  \item
+    generic: can show data from many apps
+  \item
+    neutral: doesn't prioritize a particular use case, leaves the
+    imagination open
+  \item
+    good affordances for performing common actions
+  \end{itemize}
+\item
+  could be others besides a table
+
+  \begin{itemize}
+  \tightlist
+  \item
+    tree?
+  \item
+    text box?
+  \item
+    compose these?
+  \end{itemize}
+\end{itemize}
+
+\hypertarget{decouple-data-from-application-logic}{%
+\subsection{Decouple data from application
+logic}\label{decouple-data-from-application-logic}}
+
+\begin{itemize}
+\item
+  thin client architecture of web has promoted a coupling of data + app
+
+  \begin{itemize}
+  \tightlist
+  \item
+    only way user can view the data is in the first-party tool
+  \item
+    APIs provide some flexibility in picking a client. but, APIs aren't
+    meant for end user consumption. and, there are limits (eg, Twitter
+    API restrictions)
+  \item
+    limited user agency over how they use their data
+  \item
+    compare to, say, opening a file locally
+  \end{itemize}
+\item
+  Some people aiming towards ambitious rearchitectures in this
+  space\ldots{}
+
+  \begin{itemize}
+  \tightlist
+  \item
+    cite local-first
+  \item
+    TBL Solid
+  \end{itemize}
+\item
+  Data-driven customization: an iterative approach towards this vision;
+  complementary to more radical rearchitecting.
+\item
+  Start with third party extraction
+\item
+  Integrate across apps
+\item
+  open the data in a different UI
+\item
+  Then move to first party data providing in the client (could drive
+  wildcard from existing apis??)
+\end{itemize}
+
 \hypertarget{sec:related-work}{%
@@
 This paper extends work reported in a workshop paper by Litt and Jackson
-\citep{litt2020} which presented an earlier version of the Wildcard
+\citep{litt2020} which presented an early version of the Wildcard
 extension. We have substantially extended their work in this paper by
-creating the table adapter abstarction, reimplementing the internals of
+creating the table adapter abstraction, reimplementing the internals of
 Wildcard around that abstraction, evaluating the system on many more
-websites, and by characterizing the design of the system in much more
-detail than in their workshop paper.
-
-Table-driven customization relates to two broad areas of related work.
+websites and use cases, and by characterizing the design of the system
+in much more detail than in their workshop paper.
+
+data-driven customization relates to two broad areas of related work.
 Our problem statement is related to software customization tools, and
@@
 
-Table-driven customization is most closely related to other tools that
+data-driven customization is most closely related to other tools that
 aim to empower end users to customize software without traditional
@@
 
-This lineage goes back at least to the Buttons system by MacLean et al
-\citep{maclean1990}, where Xerox Lisp users could share buttons that
+This lineage goes back at least to the Buttons system by MacLean et
+al.~\citep{maclean1990}, where Xerox Lisp users could share buttons that
 performed various ``tailoring'' actions on the system. The authors
 proposed the ``gentle slope'' idea which has greatly influenced our
-approach to table-driven customization (as discussed in
+approach to data-driven customization (as discussed in
 Section~\ref{sec:dm}). The authors also point out the importance of a
@@
 More recently, web customization tools have aimed to enable end users to
-modify web interfaces without programming.
-
-Sifter \citep{huynh2006} enables end users to sort and filter lists of
-data obtained by web scraping, much like Wildcard's sorting features.
-The main difference between the systems is that table-driven
-customization has many other use cases besides sorting and filtering.
-Also, Sifter involves end users in a semi-automated data extraction
-process, rather than having programmers create wrappers. This provides
-coverage of more websites, but at the expense of complicating the end
-user experience. We could integrate end user scraping techniques in
-Wildcard in the future, but we believe that when possible it's valuable
-for end users to have a customization experience decoupled from the
-challenge of web scraping the underlying data. Sifter also implements
-scraping across multiple pages, a valuable feature for sorting and
-filtering that isn't present in Wildcard.
+modify web interfaces without programming. Sifter \citep{huynh2006}
+enables end users to sort and filter lists of data obtained by web
+scraping, much like Wildcard's sorting features. The main difference
+between the systems is that data-driven customization has many other use
+cases besides sorting and filtering. Also, Sifter involves end users in
+a semi-automated data extraction process, rather than having programmers
+create wrappers. This provides coverage of more websites, but at the
+expense of complicating the end user experience. We could integrate end
+user scraping techniques in Wildcard in the future, but we believe that
+when possible it's valuable for end users to have a customization
+experience decoupled from the challenge of web scraping the underlying
+data. Sifter also implements scraping across multiple pages, a valuable
+feature for sorting and filtering that isn't present in Wildcard.
 
@@
 all require writing imperative programs, in contrast to the more
-declarative and direct approach of table-driven customization.
-
-\hypertarget{direct-manipulation-programming-interfaces.}{%
-\subsection{Direct manipulation programming
-interfaces.}\label{direct-manipulation-programming-interfaces.}}
-
-Another relevant area involves spreadsheets and visual query systems. We
-take inspiration from these tools in our work, but apply them in a
+declarative and direct approach of data-driven customization.
+
+\hypertarget{spreadsheets-and-visual-query-interfaces}{%
+\subsection{Spreadsheets and visual query
+interfaces}\label{spreadsheets-and-visual-query-interfaces}}
+
+Another relevant area involves spreadsheets and visual query interfaces.
+We take inspiration from these tools in our work, but apply them in a
 different domain: customizing existing software applications, rather
@@
 
-In this paper, we have presented table-driven customization, a new
+In this paper, we have presented data-driven customization, a new
 paradigm for customizing software by direct manipulation of the
@@
 incentives are not always aligned with the user's interests. We hope
-that table-driven customization can serve as a point on the path from
+that data-driven customization can serve as a point on the path from
 normal use to deep modification, in support of a more adaptable

### 2020-08-15 paper.tex
visions

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{End-User Software Customization by Direct Manipulation of Tabular
-Data}
+\title{Data-Driven Customization}
 
@@
 \begin{abstract}
-  In this paper we introduce \emph{data-driven customization}, a new way
-  for end users to extend software by direct manipulation without doing
-  traditional programming. We augment existing user interfaces with a
-  table view showing the structured data inside the application. When
-  users edit the table, their changes are reflected in the original UI.
-  This simple model accommodates a spreadsheet formula language and
-  custom data editing widgets, providing enough power to implement a
-  variety of useful extensions.
+  Customizing software should be as easy as using it. But most
+  customization methods require a dramatic shift, from using a GUI to
+  writing scripts in a programming language.
+
+  We introduce \emph{data-driven customization}, a new way for end users
+  to extend software by direct manipulation without doing traditional
+  programming. We augment existing user interfaces with a table view
+  showing the structured data inside the application. When users edit
+  the table, their changes are reflected in the original UI. This simple
+  model accommodates a spreadsheet formula language and custom data
+  editing widgets, providing enough power to implement a variety of
+  useful extensions.
 
   We illustrate the approach with Wildcard, a browser extension that
-  implements data-driven customization for existing applications using
-  web scraping. Through concrete examples, we show that this paradigm
-  can support extensions to many real websites, ranging from sorting and
-  filtering existing data to adding entire new features. We share
-  reflections from our experiences using Wildcard, on both its strengths
-  and limitations relative to other customization approaches. Finally,
-  we explore how this paradigm might lead to new software architectures
-  that encourage this form of end-user customization.
+  implements data-driven customization on the web using web scraping.
+  Through concrete examples, we show that this paradigm can support
+  useful extensions to many real websites, and we share reflections from
+  our experiences using Wildcard.
+
+  Finally, we share our broader vision for a future where end users have
+  greater access to the data in their applications, and better tools for
+  taking advantage of that data in the context of everyday software
+  usage.
 \end{abstract}
@@
 
-data-driven customization relates to existing work in many areas. Our
+Data-driven customization relates to existing work in many areas. Our
 goals overlap with many software customization tools, and our methods
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Wildcard just one instance of a broader data-driven customization
-
-  \begin{itemize}
-  \tightlist
-  \item
-    particular view: table
-  \item
-    particular context: web
-  \end{itemize}
-\item
-  here we share some of the underlying ideas. go beyond Wildcard, room
-  to grow
-\end{itemize}
+We envision data-driven customization as a broad paradigm that could
+extend well beyond the Wildcard proof-of-concept, and ultimately result
+in new software architectures that empower end users to mold software to
+their specific needs. Here we explore some of the deeper ideas
+underlying our work, and future possibilities beyond the Wildcard tool.
+
+\hypertarget{decouple-data-from-applications}{%
+\subsection{Decouple data from
+applications}\label{decouple-data-from-applications}}
+
+When data is freely available outside the context of a specific
+application, users have more freedom to choose a suitable application
+for their needs. For example, an RSS feed can be consumed by many reader
+applications---a journalist can use a power tool optimized for skimming
+hundreds of news sources a day, while a casual reader can use a simple
+app to keep up with a few blogs. Motivated users can even create their
+own custom workflows for filtering and combining RSS feeds, either via
+traditional programming or in an end-user programming environment like
+\href{https://en.wikipedia.org/wiki/Yahoo!_Pipes}{Yahoo Pipes}.
+
+However, on the Web today, data is often siloed and only accessible
+through a single prescribed application. A Facebook feed can only be
+viewed through the Facebook application. Podcasts, originally served
+openly through RSS, are
+\href{https://stratechery.com/2019/spotifys-podcast-aggregation-play/}{beginning
+to become exclusive} to specific platforms like Spotify. This coupling
+between data and applications leaves users at the mercy of using a
+single client optimized for specific purposes (e.g., maximizing
+engagement) that may not be aligned with users' individual desires.
+
+Some services provide APIs that mitigate these siloing effects, but APIs
+fail to provide a full solution to the problem. First, APIs often
+provide limited access, especially when an open client ecosystem would
+harm the economic incentives of a web service built on advertising in a
+first-party client---in 2012, Twitter infamously
+\href{https://www.theverge.com/2012/8/16/3248079/twitter-limits-app-developers-control}{imposed
+restrictions} on third-party applications that mimicked the ``mainstream
+Twitter consumer client experience.'' A second problem is that web APIs
+have a high barrier to entry---they tend to be designed more for
+programmers creating entire applications or heavy-duty automations than
+for end users casually modifying their own experience.
+
+There have been compelling suggestions for more decentralized
+architectures that would give users more control of their data.
+Local-first software \citep{kleppmann2019} suggests that productivity
+applications should run logic and store data locally, while retaining
+the benefits of realtime collaboration through peer-to-peer
+synchronization. The SOLID project \citep{berners-lee2019} envisions a
+decentralized future where users store data on their own servers, and
+choose to grant limited access to applications. We see data-driven
+customization as aligned with these decentralized visions, but
+complementary to them in two ways.
+
+\textbf{Incremental:} Rather than proposing that web services be totally
+rearchitected, data-driven customization suggests a more incremental
+path for adding user agency to existing software.
+
+Data-driven customization allows for lightly augmenting a centralized
+website with decentralized data storage. Section~\ref{sec:example}
+demonstrated how a user's private annotations on a news site could be
+stored in their browser, without needing to upload the annotation to the
+website's server. A hybrid storage model is reasonable here: a
+centralized storage model makes sense for most of the information on
+Hacker News that is viewed by all users, but a user's private
+annotations can easily just be stored in their browser. This model could
+even be extended with peer-to-peer sharing---a column in a Wildcard
+table could be shared directly among friends, providing shared
+annotation of a common website without needing to use the website itself
+as a centralized intermediary.
+
+By using third-party data extraction, data-driven customization also
+works with existing websites that do not expose structured data to the
+user. Ultimately, in adversarial situations where websites are strongly
+incentivized to restrict access to their data, scraping is unlikely to
+be a sustainable solution. But we hypothesize that there are many more
+situations where websites are \emph{neutral}: not opposed to the idea of
+end user customization, but also not sufficiently motivated to create
+and maintain a public API. We see these neutral situations as a context
+where this kind of incremental approach could succeed. Perhaps some of
+these websites might be more motivated to provide official extension
+hooks if they saw the value that users were getting from unofficial
+community-provided ones.
+
+\textbf{Focus on end user customization:} Having access to the data is a
+necessary but not sufficient condition for empowering end users to craft
+their own software experience. Another key ingredient is providing
+usable tools and interfaces for working with the data.
+
+In data-driven customization, we focus heavily on this part of the
+solution. By showing raw data in the context of a user interface and
+allowing small tweaks to the original application's behavior, we provide
+a smooth path for people to move from using an application to tweaking
+it.
+
+In this sense, data-driven customization is a complementary approach to
+other projects that focus on getting users greater access to their data.
+In a decentralized future where data is stored locally rather than in
+cloud silos, interfaces like Wildcard would be one technique for
+actually making use of this data in service of greater end user
+flexibility.
 
@@
 
-\hypertarget{decouple-data-from-application-logic}{%
-\subsection{Decouple data from application
-logic}\label{decouple-data-from-application-logic}}
-
-\begin{itemize}
-\item
-  thin client architecture of web has promoted a coupling of data + app
-
-  \begin{itemize}
-  \tightlist
-  \item
-    only way user can view the data is in the first-party tool
-  \item
-    APIs provide some flexibility in picking a client. but, APIs aren't
-    meant for end user consumption. and, there are limits (eg, Twitter
-    API restrictions)
-  \item
-    limited user agency over how they use their data
-  \item
-    compare to, say, opening a file locally
-  \end{itemize}
-\item
-  Some people aiming towards ambitious rearchitectures in this
-  space\ldots{}
-
-  \begin{itemize}
-  \tightlist
-  \item
-    cite local-first
-  \item
-    TBL Solid
-  \end{itemize}
-\item
-  Data-driven customization: an iterative approach towards this vision;
-  complementary to more radical rearchitecting.
-\item
-  Start with third party extraction
-\item
-  Integrate across apps
-\item
-  open the data in a different UI
-\item
-  Then move to first party data providing in the client (could drive
-  wildcard from existing apis??)
-\end{itemize}
-
 \hypertarget{sec:related-work}{%
@@
 
-data-driven customization relates to two broad areas of related work.
+Data-driven customization relates to two broad areas of related work.
 Our problem statement is related to software customization tools, and
@@
 
-data-driven customization is most closely related to other tools that
+Data-driven customization is most closely related to other tools that
 aim to empower end users to customize software without traditional

### 2020-08-15 paper.tex
wrapping up visions

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Data-Driven Customization}
+\title{Data-Driven Customization of Web Applications}
 
@@
 
-We have developed a browser extension called Wildcard which uses web
-scraping techniques to implement data-driven customization for existing
-Web applications. In Section~\ref{sec:example}, we concretely
-demonstrate the end user experience of using Wildcard to add features to
-a news website.
-
-In Section~\ref{sec:architecture}, we explain the system architecture of
-data-driven customization. We focus on the \emph{table adapter}
-abstraction, which allows many different types of underlying data to be
-bidirectionally mapped to a table. We describe several types of table
-adapters we've built in Wildcard, and also describe future adapters
-supported by the general paradigm.
-
-We have successfully used Wildcard to build customizations for 11
-different websites that serve our own personal needs. In
-Section~\ref{sec:reflections}, we present reflections from this process.
-We outline the kinds of customizations we were able to build,
-limitations we encountered, and some of the challenges of writing
-scraping logic.
-
-In Section~\ref{sec:visions}, we discuss some key themes from our work:
+To explore this idea in a real context, we have developed a browser
+extension called Wildcard that uses web scraping techniques to implement
+data-driven customization for existing Web applications. We introduce
+the tool with an example customization of Hacker News in
+Section~\ref{sec:example}, and then describe the implementation in
+{[}Section~\ref{sec:architecture}. In Section~\ref{sec:reflections}, we
+present evidence that Wildcard can produce useful customizations, by
+sharing reflections from customizing 11 different websites in ways that
+met our own personal needs.
+
+Wildcard is just an initial proof of concept of data-driven
+customization. In Section~\ref{sec:visions}, we discuss our broader
+vision for how this style of customization could change the relationship
+between users and creators of software, focusing on three ideas:
 
@@
 \item
-  \emph{Customization by direct manipulation}: We discuss further how
-  data-driven customization relates to a gentle slope of customization.
-  We suggest that an important point on this slope is the ability to
+  \emph{Decoupling data from applications}: On the modern Web, data is
+  often stored in proprietary silos, limiting the agency of users to
+  choose their applications and flexibly work with data. We propose
+  data-driven customization as an incremental step towards a more
+  decentralized architecture, where users gain more control over the
+  storage, processing and display of information from web services.
+\item
+  \emph{Customization by direct manipulation}: We discuss how
+  data-driven customization promotes a gentle slope of customization. We
+  propose that an important point on this slope is the ability to
   customize an application by directly seeing and changing its data,
@@
 
-Data-driven customization relates to existing work in many areas. Our
-goals overlap with many software customization tools, and our methods
-overlap with direct manipulation interfaces for working with structured
-data, including visual database query systems and spreadsheets. We
-explore these connections in Section~\ref{sec:related-work}.
+In Section~\ref{sec:related-work} we discuss connections to related
+work. Our goals overlap with software customization tools, and our
+methods overlap with direct manipulation interfaces for working with
+structured data, including visual database query systems and
+spreadsheets.
 
@@
 
-\hypertarget{decouple-data-from-applications}{%
-\subsection{Decouple data from
-applications}\label{decouple-data-from-applications}}
+\hypertarget{decoupling-data-from-applications}{%
+\subsection{Decoupling data from
+applications}\label{decoupling-data-from-applications}}
 

### 2020-08-15 paper.tex
new title and abstract

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Data-Driven Customization of Web Applications}
+\title{Test}
 
@@
 \begin{abstract}
-  Customizing software should be as easy as using it. But most
-  customization methods require a dramatic shift, from using a GUI to
-  writing scripts in a programming language.
-
-  We introduce \emph{data-driven customization}, a new way for end users
-  to extend software by direct manipulation without doing traditional
-  programming. We augment existing user interfaces with a table view
-  showing the structured data inside the application. When users edit
-  the table, their changes are reflected in the original UI. This simple
-  model accommodates a spreadsheet formula language and custom data
-  editing widgets, providing enough power to implement a variety of
-  useful extensions.
-
-  We illustrate the approach with Wildcard, a browser extension that
-  implements data-driven customization on the web using web scraping.
-  Through concrete examples, we show that this paradigm can support
-  useful extensions to many real websites, and we share reflections from
-  our experiences using Wildcard.
-
-  Finally, we share our broader vision for a future where end users have
-  greater access to the data in their applications, and better tools for
-  taking advantage of that data in the context of everyday software
-  usage.
+  Test.
 \end{abstract}

### 2020-08-15 paper.tex
new abstract, new diff

--- 2020-08-15 paper.tex
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Data-Driven Customization of Web Applications}
+\title{End-User Software Customization by Direct Manipulation of Tabular
+Data}
 
@@
   Customizing software should be as easy as using it. But most
-  customization methods require a dramatic shift, from using a GUI to
-  writing scripts in a programming language.
+  customization methods require a dramatic shift from using a GUI to
+  writing customization scripts in a programming language.
 

### 2020-08-15 paper.tex
done

@@
 
-A helpful future addition would be to allow end users to create wrappers
-themselves, so that more people could contribute to the shared library.
-There are many existing projects exploring end-user web scraping (such
-as Helena \citep{chasins2018}) which might prove helpful; however,
-because the needs of a Wildcard adapter are slightly more complex than a
-traditional web scraper (as discussed in
-Section~\ref{sec:architecture}), further work might be needed to enable
-end-user wrapper creation.
-
 The distribution mechanism for semantic wrappers is also important for
@@
 mechanism is simply merging the code for all adapters into the main
-Wildcard codebase. This is the simplest workable solution, but has its
-downsides: contributing involves a fairly high barrier of creating a
-pull request on Github, and using newly contributed wrappers requires
-installing a new version of the extension. In the future we might
-explore other mechanisms, like an online repository that the extension
-pulls from dynamically. Security is also a consideration---DOM scraping
-adapters can execute arbitrary Javascript code, which opens up the
-possibility of malicious adapters being contributed. Currently we plan
-to solve this challenge with centralized code review, but another
-approach we are considering is using or inventing a more restricted
-domain-specific language for specifying scraping logic.
-
-\hypertarget{generic-data-interfaces}{%
-\subsection{Generic data interfaces}\label{generic-data-interfaces}}
-
-\begin{itemize}
-\item
-  what makes a UI good for customization?
-
-  \begin{itemize}
-  \tightlist
-  \item
-    generic: can show data from many apps
-  \item
-    neutral: doesn't prioritize a particular use case, leaves the
-    imagination open
-  \item
-    good affordances for performing common actions
-  \end{itemize}
-\item
-  could be others besides a table
-
-  \begin{itemize}
-  \tightlist
-  \item
-    tree?
-  \item
-    text box?
-  \item
-    compose these?
-  \end{itemize}
-\end{itemize}
+Wildcard codebase. This is a simple solution, but makes it fairly
+difficult to contribute new wrappers and requires installing a new
+version of the extension to gain access to new warppers. In the future
+we might explore other mechanisms, like an online repository that the
+extension pulls from dynamically. Security is also a consideration---DOM
+scraping adapters can execute arbitrary Javascript code, which means a
+malicious adapter could exfiltrate sensitive information from a page.
+Approaches to security could include centralized code review, using a
+restricted scraping DSL, or creating a sandboxed context for scrapers
+without access to networking APIs.
+
+\hypertarget{alternate-mechanisms-for-wrapper-creation}{%
+\subsubsection{Alternate mechanisms for wrapper
+creation}\label{alternate-mechanisms-for-wrapper-creation}}
+
+Requiring programming to create wrappers has an obvious limitation. If
+an end user wants to customize a site and no programmer has contributed
+a wrapper for that site, then they have no means of customizing it.
+Although a sufficiently vibrant community of programmers could produce
+wrappers for many popular sites, it's unrealistic to imagine covering
+all websites that end users might want to customize. We envision two
+strategies for dealing with this problem:
+
+\textbf{End user wrapper creation:} If end users could create wrappers
+without programming, there would be fewer limitations on the sites they
+could customize. There are existing projects exploring end-user web
+scraping (such as Helena \citep{chasins2018}) which might prove helpful;
+however, because the needs of a Wildcard adapter are slightly more
+complex than a traditional web scraper (as discussed in
+Section~\ref{sec:architecture}), further work might be needed to enable
+end-user wrapper creation.
+
+\textbf{First party wrapper creation:} An integrated adapter installed
+by the developer of an application could directly access internal state,
+providing the same functionality as a DOM scraping adapter but in a more
+robust way.
+
+With the advent of rich frontend web frameworks, structured application
+state is now often available in the web client. We suspect it is
+possible to create plugins for frontend frameworks that expose this
+state to Wildcard with only minimal effort from the application
+developers. Customizability is sometimes a key selling point for
+software. An integrated website adapter would provide a way for
+developers to integrate with an ecosystem of formulas and customization
+tools without needing to build all that functionality from scratch.
 

### 2020-08-15 paper.tex
final tweaks

@@
   useful extensions to many real websites, and we share reflections from
-  our experiences using Wildcard.
-
-  Finally, we share our broader vision for a future where end users have
-  greater access to the data in their applications, and better tools for
-  taking advantage of that data in the context of everyday software
-  usage.
+  our experiences using the tool.
+
+  Finally, we share our broader vision for data-driven customization: a
+  future where end users have greater access to the data inside their
+  applications, and better tools for flexibly making use of that data in
+  the context of everyday software usage.
 \end{abstract}
@@
 Section~\ref{sec:example}, and then describe the implementation in
-{[}Section~\ref{sec:architecture}. In Section~\ref{sec:reflections}, we
+Section~\ref{sec:architecture}. In Section~\ref{sec:reflections}, we
 present evidence that Wildcard can produce useful customizations, by
@@
 website UI. A DOM scraping adapter fulfills the standard web scraping
-task of extracting a table of data from the DOM, but it must also handle
-the reverse direction: manipulating the DOM to reflect edits to the
-table.
-
-Currently, we require DOM scraping adapters to be programmed manually
-for each website using Javascript code. It might seem that this
-prohibits non-programmer users from using the system, but we solve this
+task of extracting a table of data from the DOM, but it also acts in the
+reverse direction: manipulating the DOM to reflect edits to the table.
+
+In Wildcard, DOM scraping adapters are programmed manually for each
+website using Javascript code. It might seem that this prohibits
+non-programmer users from using the system at all, but we solve this
 problem via a shared repository of adapters. Once an adapter is
-programmed for a website, it is added to a shared code repository,
-enabling any end user to perform customizations on that website.
-
-There are other ways a DOM scraping adapter could be produced which
-would reduce this dependence on programmers: an end user could specify
+programmed for a website, it is added to the shared repository, enabling
+any end user to perform customizations on that website.
+
+In the future, other strategies for producing DOM scraping adapters
+could reduce this dependence on programmers: an end user could specify
 the scraping logic via demonstration, or the desired data table could be
 automatically inferred from the page. While we are interested in these
-techniques and discuss them later in Section~\ref{sec:visions}, we
-believe that a shared repository of manually programmed adapters is a
-pragmatic starting point; given that many users visit the same popular
-websites,a critical mass of adapters could serve the needs of many
-users.
-
-To ease the burden on programmers of creating these adapters, we have
-created a framework makes the process feel more like writing
-unidirectional scraping code than performing a complex bidirectional
-synchronization. The key idea is this: programmers return pointers to
-DOM elements representing table rows and table cells; Wildcard extracts
-data from these DOM elements, but it also uses the pointers to
-synchronize table edits back into the page. For example, when the user
-sorts the table, the DOM elements representing the table rows are moved
-around in the DOM to reflect the new sorted order.
+techniques and discuss them in Section~\ref{sec:visions}, we believe
+that a shared repository of manually programmed adapters is a pragmatic
+starting point; given that many users visit the same popular websites,a
+critical mass of adapters could serve the needs of many users.
+
+To make it easier to create these adapters, Wildcard provides a
+framework that makes the process feel more like writing unidirectional
+scraping code than performing a complex bidirectional synchronization.
+The key idea is this: programmers return pointers to DOM elements
+representing table rows and table cells; Wildcard extracts data from
+these DOM elements, but it also uses the pointers to synchronize table
+edits back into the page. For example, when the user sorts the table,
+the DOM elements representing the table rows are moved around in the DOM
+to reflect the new sorted order.
 

### 2020-08-15 paper.tex
final tweaks

@@
 
-Here we describe the three types of table adapters we have built so far
-in Wildcard. These do not represent an exhaustive set of all possible
-table adapters---in Section~\ref{sec:visions} we discuss other types of
-adapters that would fit well into the general paradigm.
+\hypertarget{abstract-interface}{%
+\subsubsection{Abstract interface}\label{abstract-interface}}
+
+We begin by describing the abstract interface fulfilled by a table
+adapter.
+
+\emph{Returning a table}: A table adapter exposes a table of data: an
+ordered list of records. Each record carries a unique identifier and
+associates named attributes with values. Tables have a typed schema, so
+the same attributes are shared across all records. We currently support
+strings, numeric values, booleans, and datetimes as types. The columns
+also carry some additional metadata, such as whether or not they are
+read-only or editable.
+
+A table adapter can update the contents of a table at any time in
+response to changes in the underlying state (e.g., a DOM scraping
+adapter can update the table when the page body changes). When data
+changes, the query view is reactively updated in response.
+
+\emph{Handling edits}: The query engine can issue a request to a table
+adapter to make an edit to a record. The meaning of making an edit can
+vary depending on the adapter: in the local storage adapter, a new value
+may be persisted into local storage; in the DOM scraping adapter, an
+edit may result in changing the value of a form field.
+
+In addition, the query engine also sends additional information about
+the combined query view to each table adapter:
+
+\emph{Sorting/filtering}: When the user sorts or filters the query view,
+an ordered list of visible IDs is sent to each table adapter. The DOM
+scraping adapter uses this information to change the list of rows shown
+in the web page.
+
+\emph{Data from other tables}: The query engine provides each table
+adapter with the entire combined table shown to the user. The DOM
+scraping adapter uses this for injecting annotations---values from other
+tables are added to the original web page.
+
+\emph{Currently selected record}: As the user clicks around the table
+view, the query engine broadcasts the record currently selected by the
+user to each table adapter. The DOM scraping adapter uses this
+information to highlight the row in the page that corresponds to the
+selected row in the table, which helps the user understand the mapping
+between the table and the original UI.
+
+Next, we present the three types of table adapters we have built in
+Wildcard so far. These do not represent an exhaustive set of all
+possible table adapters---in Section~\ref{sec:visions} we discuss other
+types of adapters that would fit well into the general paradigm.
 

### 2020-08-17 paper.tex
second round submission version

@@
 
-We developed two cell editors, to explore the benefits of adding custom
-UI for editing values in the table.
-
-One valuable feature is that users can incorporate their private
-information into a web UI. We created a datepicker based on the
-\href{https://fullcalendar.io/}{FullCalendar} plugin, which can load
-data from a Google Calendar. This makes it convenient to enter dates
-into a website based on the user's personal calendar information,
-without uploading that information to the website itself.
-
-Another benefit is that a user can choose a preferred widget for editing
-information across different sites that typically provide a fragmented
-editing experience. We built a text editor based on the
-\href{https://ckeditor.com/}{CKEditor} rich text editor, and integrated
-it with Google's Blogger website by representing the contents of a blog
-post as a single table cell containing an HTML string (shown in
-Figure~\ref{fig:blogger}). If there were site adapters available for
-other websites with rich text entry, the user could then use a single
-rich text editor to edit text across the whole web.
+We developed two \emph{cell editors}: custom UI widgets for editing
+values in the table.
+
+One benefit that cell editors provide is enabling users to incorporate
+their private information into a web UI. We created a datepicker widget
+(based on the \href{https://fullcalendar.io/}{FullCalendar} plugin),
+which can load calendar data from a Google Calendar. This makes it
+convenient to enter dates into a website based on the user's personal
+calendar information, without needing to upload a user's calendar to the
+website itself.
+
+Another benefit is that a user can choose a single preferred widget for
+editing a certain type of information across different sites. For
+example, a user could use their favorite rich text editor to edit text
+in various websites like blogging platforms and task trackers. To
+demonstrate this capability, we built a text editor based on the
+\href{https://ckeditor.com/}{CKEditor} rich text editor. We used the
+editor with Google's Blogger website, by building a site adapter that
+represented the contents of a blog post as a single table cell
+containing an HTML string (shown in Figure~\ref{fig:blogger}).
 
